### 2025-02-09 paper.qmd
abstract submission for DASC'25

---
title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"

filters:
  - acronyms
  - highlight-text
  - abstract-section

format:
  html: default
  ieee-pdf:
    keep-tex: true
    journaltype: conference
    papersize: a4
  ieee-html: default

author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium

bibliography: references.bib

acronyms:
  insert_loa: false
  fromfile:
    - ./media/acronyms.yml

---

# Abstract
The risk of collision of re-entering space debris with aircraft is 
increasing due to both increasing number of airborne flights and
re-entries. National and international authorities need to balance
safety and economic impact of adopted measures.  
We asses airspaces density as flown by commercial airlines
distinguishing seasonal and hourly patterns.
We base our hourly density estimation on (enhanced) flown trajectories as
recorded by EUROCONTROL's \acr{NM} binned over Uber's H3
gridding system. Impact risk probabilities will then calculated based on work
from [@patera2008].
This combined with a buffered area along the the path of re-entry,
can provide risk impact probabilities for deciding on safety measure such as
airspace closures and/or re-routings.

# Introduction

We base our hourly flight density estimation on (enhanced) trajectories flown
by \acr{IFR} commercial aircraft as
recorded by EUROCONTROL's \acr{NM}[^1] and overlapped on Uber's H3
gridding system at different resolutions [@h3resolutions], see @fig-cells 
for resolution 2 which corresponds to hexagons of average area $86801.78\ km^2$
and average edge length of $182.51 km$.

::: {#fig-cells}

```{r cells-map}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("figures/cells_2.png")
```

H3 cells (red) at resolution 2 and their centers (blue) covering EUROCONTROL Area.
:::

We characterise densities on an hourly basis for representative days
in summer and winter.[^2]

We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008; @ailorLargeConstellationDisposal2020] and
provide a reproducible example for States, Agencies or relevant authorities
to build upon.

[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States
and 2 Comprehensive Member States covering
almost all \acr{ECAC} States.

[^2]: We will consider representative days in \acr{IATA} summer 2024 
and winter 2025 season.

@fig-density-winter-res-2 provides a promising result in the hourly flight
density calculation which at least shows
expected outcomes such as cells at zero density for war or no-fly zones such as
Ukraine, Libia or Syria.

::: {#fig-density-winter-res-2}

```{r density-winter-res-2}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("figures/flight_density_2024-08-01T12_2.png")
```

Flight density on 2024-08-01T12 for H3 cells at resolution 2 covering EUROCONTROL Area.
:::

# Conclusion / Outcome

We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled reentry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

# References

::: {#refs}
:::

### 2025-02-09 R/cells-map.R

@@ -13,8 +13,8 @@ source("R/cells.R")
 
 res <- 3
 g <- plot_hexes_map(res)
-ggsave(filename = here("figures", str_glue("cells-{res}.png")), plot = g)
+ggsave(filename = here("figures", str_glue("cells_{res}.png")), plot = g)
 
 res <- 2
 g <- plot_hexes_map(res)
-ggsave(filename = here("figures", str_glue("cells-{res}.png")), plot = g)
+ggsave(filename = here("figures", str_glue("cells_{res}.png")), plot = g)

### 2025-02-17 paper.qmd
shuffle files around

@@
 format:
-  html: default
   ieee-pdf:
@@
 
-bibliography: references.bib
+bibliography: media/references.bib
 
@@
 for resolution 2 which corresponds to hexagons of average area $86801.78\ km^2$
-and average edge length of $182.51 km$.
+and average edge length of $182.51 km$, see also table
+
+```{r}
+library(tidyverse)
+library(gt)
+library(gtExtras)
+
+tribble(
+  ~resolution, ~hex_area, ~hex_edge,
+  2L , 86801.780398997, 182.5129565,
+  3L,  12393.434655088,  68.97922179,
+  4L,   1770.347654491,  26.07175968
+
+) |> 
+  gt() |> 
+   cols_label(
+    resolution ~ "Resolution",
+    tidyselect::ends_with("area") ~ "Density, {{*persons* km^-2}}",
+    # hex_area ~ md("Cell Area ({km^2}})"),
+    hex_edge ~ "Edge length ({km}})"
+  ) |> 
+    gt::tab_header(
+    title = md("H3 Cell Statistics"),
+    subtitle = md("average hexagons")
+  ) |> 
+  opt_align_table_header(align = "left")
+```
 
@@
 #| echo: false
-knitr::include_graphics("figures/cells_2.png")
+knitr::include_graphics("media/figures/cells_2.png")
 ```
@@
 #| echo: false
-knitr::include_graphics("figures/flight_density_2024-08-01T12_2.png")
+knitr::include_graphics("media/figures/flight_density_2024-08-01T12_2.png")
 ```

### 2025-03-07 abstract_dasc25.qmd
added plot for density (of aircraft, population) data storage

---
title: "Call for Papers --- DASC 2025"
filters:
  - highlight-text
format: html
---

## DASC '25
The Conference will be held in Montreal, Canada, on

## Conference Theme

AVIONICS’ RESPONSIBILITY FOR SAFE AND SUSTAINABLE NEXT GENERATION AEROSPACE OPERATIONS

Digital avionics are the foundation for the kind of aviation and space flight we feel familiar with today, but aerospace finds itself in an era of tremendous changes: societal demands and geopolitical changes are transforming aerospace towards higher customization, more autonomy, greener operation and faster adaptability.

Digital avionics systems are the core enablers for many of the new technologies, functions, and operations necessitated by these changes. Furthermore, avionics systems will need to undergo a transformation towards reduced carbon footprint, higher calculation power, lower latency, and higher flexibility - all while maintaining safety and security levels and decreasing certification times: a huge responsibility for the digital avionics systems of the future.

The 44th DASC will investigate the responsibility of digital avionics for the next generation of air and space vehicles to come. We provide the forum to present solutions making aerospace transformation possible, analyze open issues, and discuss disruptive ideas.

You are invited to present your research addressing current and future challenges of avionics systems and exchange diverse perspectives with the world’s leading experts in the field.

### Fields of Interest
Emphasized are methods, tool, systems and functions for:

The DASC encourages leading research & applications of the following fields but is open to other related topics as well.

* Air Traffic Management (ATM)
* Avionics Platforms
* Avionics Technologies
* Communications, Navigation, And Surveillance
* And Information Networks 
* Cyber, Systems, And Software (-Engineering)
* Human Factors
* Unmanned Aircraft Systems & Uncrewed
* Spacecraft (UAS/USS)
* Urban, Advanced Air Mobility & New Space

## Proposal

**Title**: Prepariing for Potential Closure of European Airspaces due to Re-entering Space Objects

One of each in the abstract:

i.  motivation
i. short background why topic important
i. approach chosen/research method
i. experimental work
i. results obtained/hoped~planned
i. conclusions

### Abstract
(background, motivation)
The risk of collision of re-entering space debris with aircraft is 
increasing both due to ever increasing number of airborne flights and
re-entries. National and international authorities need tools to balance
safety and economic/logistic impact of adopted measures.

(approach)
We asses airspaces density as flown by commercial airlines
distinguishing seasonal (summer and winter IATA season) and hourly patterns.
(method)
We base our hourly density estimation on (enhanced) flown trajectories as
recorded by EUROCONTROL's Network Manager (NM)[^1] and overlapped on Uber's H3
gridding system. Impact risk probabilities will then calculated based on work
from [@patera2008].
(conclusion)
This combined with a buffered area along the the path of re-entry,
can provide factual risk impact probabilities for deciding on closures and/or
re-routings.
(conclusion)

[^1]: NM coordinates ATFCM operations on behalf of EUROCONTROL's 42 Member States
and 2 Comprehensive Member States covering
almost all European Civil Aviation Conference (ECAC) States.

::: {#fig-cells}

```{r cells-map}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("figures/flight_density_2024-08-01T12_2.png")
```

Flight density on 2024-08-01T12 for H3 cells at resolution 2 covering EUROCONTROL Area.
Note the empty cells for the war areas / no-fly zone in Ukraine, Syria and Libia.

:::

## Text in Web Based submission

### Synopsis
The risk of collision of re-entering space debris with aircraft is increasing due to both increasing number of airborne flights and re-entries. National and international authorities need to balance safety and economic impact of adopted measures.
We asses airspaces density as flown by commercial airlines distinguishing seasonal and hourly patterns. We base our hourly density estimation on (enhanced) flown trajectories as recorded by EUROCONTROL's Network Manager binned over Uber's H3 gridding system. Impact risk probabilities will then calculated based on work such as from (Patera R. "Risk ro commercial aircraft from reentering space debris", 2008). This combined with a buffered area along the the path of re-entry, can provide risk impact probabilities for deciding on safety measure such as airspace closures and/or re-routings.

### Extended Abstract
We base our hourly flight density estimation on (enhanced) trajectories flown by Instrument Flight Rule (IFR) commercial flights as recorded by EUROCONTROL's Network Manager (NM)1 and overlapped on Uber's H3 gridding system at different resolutions [2], see <Figure 1> for resolution 2 which corresponds to hexagons of average area 86801.78 km2 and average edge length of 182.51 km.

<Figure 1: plot of the grid system at resolution 2>

We characterise densities on an hourly basis for representative days in summer and winter.[^2]
We expand on the works of [1], [3], [4] and provide a reproducible example for States, Agencies or relevant authorities to build upon.

<Figure 2> provides a promising result in the hourly flight density calculation which shows expected outcomes such as cells at zero density for war or no-fly zones such as Ukraine, Libia or Syria. Typical traffic flows such as from central Europe to Canary Islands or Atlantic crossing are clearly visible, providing a visual confirmation of the approach.

<Figure 2: Flight density on 2024-08-01T12 for H3 cells at resolution 2 covering EUROCONTROL Area.>

Conclusion / Outcome

We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled reentry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

Footnotes

[1]: NM coordinates Air Traffic Flow and Capacity Management (ATFCM) operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all European Civil Aviation Conference (ECAC) States.

[2]: We will consider representative days in International Air Transport Association (IATA) summer 2024 and winter 2025 season.

### 2025-03-07 paper.qmd
added plot for density (of aircraft, population) data storage

@@
     resolution ~ "Resolution",
-    tidyselect::ends_with("area") ~ "Density, {{*persons* km^-2}}",
+    tidyselect::ends_with("area") ~ "Cell Area {{*persons* km^-2}}",
     # hex_area ~ md("Cell Area ({km^2}})"),
@@
 
+# Data
+
+::: {#fig-data-densities}
+
+```{r densities}
+#| out-width: "120%"
+#| echo: false
+knitr::include_graphics("media/figures/cubes.pdf")
+```
+
+Flight and population densities data (hex on Germany as an example).
+:::
+
 # Conclusion / Outcome

### 2025-03-07 setup.qmd
added plot for density (of aircraft, population) data storage

---
title: "Setup"
filters:
  - highlight-text
format: 
  html: default
  pdf: default
---

This works takes NM CTFM profiles, resamples them and calculates number of
flight present per hour, `H`, of date, `D` in H3 hexagonal cells
(at resolution `R`) covering the EUROCONTROL Member's airspace.

R and Python scripts are in the `R/` and `python/` directories respectively.

We use the following values:

* H3 hexagons at resolution `R = 2`: 
  - Average Area (km2): `86801.78`
  - Average edge length (Km): `182.51`
* H3 hexagons at resolution `R = 3`: 
  - Average Area (km2): `12393.43`
  - Average edge length (Km): `68.98`
* bounding box approximating EUROCONTROL Area (decimal degrees):
  - xmin = -25.488281
  - ymin = 26.638253
  - xmax = 45.407181
  - ymax = 71.864780
* bounding box containing the HEX at resolution 2 that match
  the bounding box approximating EUROCONTROL Area:
  - xmin = -27.78166
  - xmax =  48.56037
  - ymin =  24.90098
  - ymax =  72.85065

::: {#fig-cells}

```{r cells-map}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("figures/cells_2.png")
```

H3 **[cells]{colour="#ff0000"}** ([red]{colour="#ffffff" bg-colour="#ff0000"})
at resolution 2 and
their **[centers]{colour="#0000ff"}** ([blue]{colour="#ffffff" bg-colour="#0000ff"}) covering EUROCONTROL Area.

:::

## Data Preparation

### Data Extraction
We first extract data for data `D` (for example `2024-08-01`)
1. flight list for date `D`
1. point profile for flown trajectories, i.e. M3/CTFM for date `D`
   in the correct bounding box.
   We use `point_profiles_tidy()` from the `{eurocontrol}` package

This is achieved via script `export_data.R`.
The relevant trajectories are saved in
`data-raw/trjs_{D}.parquet`, i.e. `data-raw/trjs_2024-08-01.parquet`.

### Trajectory Resampling
We load the trajectories and resample them at 30s using the Python `traffic` library.
This is performed by the `resample_aug.py` and `resample_dec.py`
scripts and is time consuming (40 min each).
Execute as:

> $ python resample_aug.py

The result is in (52min 51s)
`data/trajectories_2024-08-01_resampled_30s.parquet`

### Fast H3 Labelling
The script `prepare_pre_density.R` uses DuckDB for fast labelling
of each lon/lat to the relevant H3 cell (at resolution `R`)
and filtering for samples between bbox.
Keep only point in the hex cells of the bbox and for day `D`.
The result is saved in `trajectories_<YYYY-MM-DD>_resampled_30s_bbox_res_3.parquet` for date `D`.

## Cell Occupancy and Density

The script `prepare_density.R` calculates the occupancy per hour `H` of `D` 
for each cell at resolution `R`
.

### 2025-03-07 R/data-cube.R

@@ -1,0 +1,140 @@
+# adapted from Fig 6.9 (and relevant R code) in
+# https://r-spatial.org/book/06-Cubes.html#sec-switching
+
+set.seed(1331)
+library(here)
+library(stars)
+library(colorspace)
+tif <- system.file("tif/L7_ETMs.tif", package = "stars")
+r <- read_stars(tif)
+
+
+nrow <- 2
+ncol <- 8
+m <- matrix(runif(nrow * ncol), nrow = nrow, ncol = ncol)
+m <- r[[1]][1:nrow, 1:ncol, 1]
+dim(m) <- c(x = nrow, y = ncol) # named dim
+s <- st_as_stars(m)
+# s
+attr(s, "dimensions")[[1]]$delta = 3
+attr(s, "dimensions")[[2]]$delta = -.5
+attr(attr(s, "dimensions"), "raster")$affine = c(-1.2, 0.0)
+
+# plot plate
+plt <- function(x, yoffset = 0, add, li = TRUE) {
+  attr(x, "dimensions")[[2]]$offset = attr(x, "dimensions")[[2]]$offset +
+    yoffset
+  l = st_as_sf(x, as_points = FALSE)
+  pal = sf.colors(10)
+  if (li) pal = lighten(pal, 0.3 + rnorm(1, 0, 0.1))
+  if (!add)
+    plot(
+      l,
+      axes = FALSE,
+      breaks = "equal",
+      pal = pal,
+      reset = FALSE,
+      border = grey(.75),
+      key.pos = NULL,
+      main = NULL,
+      xlab = "hour"
+    ) else
+    plot(
+      l,
+      axes = TRUE,
+      breaks = "equal",
+      pal = pal,
+      add = TRUE,
+      border = grey(.75)
+    )
+  u = st_union(l)
+  plot(st_geometry(u), add = TRUE, col = NA, border = 'black', lwd = 2.5)
+}
+
+# plot the cube
+pl <- function(s, x, y, add = TRUE, randomize = FALSE) {
+  attr(s, "dimensions")[[1]]$offset = x
+  attr(s, "dimensions")[[2]]$offset = y
+  m = r[[1]][y + 1:nrow, x + 1:ncol, 1]
+  if (randomize) m = m[sample(y + 1:nrow), x + 1:ncol]
[…]

### 2025-03-07 R/density-map.R

@@ -1,0 +1,96 @@
+library(conflicted)
+library(tidyverse)
+library(arrow)
+library(sf)
+library(h3jsr)
+library(ggplot2)
+library(h3jsr)
+library(magrittr)
+library(giscoR)
+
+conflicts_prefer(dplyr::filter)
+
+source("R/cells.R")
+
+density <- arrow::read_parquet("data/flight_density.parquet")
+
+res  <- 3
+YYYY <- 2024
+MM   <- 8
+DD   <- 1
+HH   <- 10
+
+eur_hex <- bbox_nm() |>
+  hexes_for_bbox_at_res(resolution = res)
+
+
+ddd <- density |>
+  filter(
+    year == YYYY,
+    month == MM,
+    day == DD,
+    hour == HH,
+    h3_resolution == res) |>
+  mutate(occupancy = mean(occupancy), .by = cell) |>
+  filter(cell %in% (eur_hex |> pull(h3_address))) |>
+  select(cell, everything()) |>
+  cell_to_polygon(simple = FALSE)
+
+MAX <- ddd |> pull("occupancy") |> max()
+min <- ddd |> pull("occupancy") |> min()
+min <- 0
+
+
+
+eur_hex_union <- bbox_nm() |>
+  bbox_of_hexes_for_bbox_at_res(resolution = res)
+
+gisco_RG <- gisco_get_countries(
+  resolution = "20",
+  epsg = "4326",
+  spatialtype = "RG",
+  region = c("Europe", "Africa", "Asia")) |>
+  mutate(res = "20M")
+
+gisco <- gisco_RG
+
+EUR_res20 <- gisco |>
+  st_intersection(eur_hex_union) |>
+  select(res)
+
+
+ggplot() +
+  geom_sf(data = EUR_res20,
+          fill = 'lightgrey',
+          linewidth = 0.1,
+          # alpha = 0.5,
+          colour = NA) +
+  geom_sf(data = eur_hex,
+          fill = NA,
+          colour = 'red',
+          linewidth = 0.05) +
+  geom_sf(data = ddd, aes(fill = occupancy), alpha = 0.8) +
+  scale_fill_viridis_c(
+    trans = "log", breaks = c(1, 5, 10, 20, 30, 40, 50),
+    name = "Flights per hour",
+    guide = guide_legend(
+      position = "top",
+      keyheight = unit(3, units = "mm"),
+      keywidth = unit(12, units = "mm"),
+      label.position = "bottom",
+      title.position = "top",
[…]

### 2025-03-07 R/fig_6-9.R

@@ -1,0 +1,84 @@
+# from Fig 6.9 (and relevant R code) in
+# https://r-spatial.org/book/06-Cubes.html#sec-switching
+
+set.seed(1331)
+library(stars)
+library(colorspace)
+tif <- system.file("tif/L7_ETMs.tif", package = "stars")
+r <- read_stars(tif)
+
+nrow <- 5
+ncol <- 8
+#m = matrix(runif(nrow * ncol), nrow = nrow, ncol = ncol)
+m <- r[[1]][1:nrow,1:ncol,1]
+dim(m) <- c(x = nrow, y = ncol) # named dim
+s <- st_as_stars(m)
+# s
+attr(s, "dimensions")[[1]]$delta = 3
+attr(s, "dimensions")[[2]]$delta = -.5
+attr(attr(s, "dimensions"), "raster")$affine = c(-1.2, 0.0)
+
+plt <- function(x, yoffset = 0, add, li = TRUE) {
+  attr(x, "dimensions")[[2]]$offset = attr(x, "dimensions")[[2]]$offset + yoffset
+  l = st_as_sf(x, as_points = FALSE)
+  pal = sf.colors(10)
+  if (li)
+    pal = lighten(pal, 0.3 + rnorm(1, 0, 0.1))
+  if (! add)
+    plot(l, axes = FALSE, breaks = "equal", pal = pal, reset = FALSE, border = grey(.75), key.pos = NULL, main = NULL, xlab = "time")
+  else
+    plot(l, axes = TRUE, breaks = "equal", pal = pal, add = TRUE, border = grey(.75))
+  u = st_union(l)
+  plot(st_geometry(u), add = TRUE, col = NA, border = 'black', lwd = 2.5)
+}
+
+pl <- function(s, x, y, add = TRUE, randomize = FALSE) {
+  attr(s, "dimensions")[[1]]$offset = x
+  attr(s, "dimensions")[[2]]$offset = y
+  m = r[[1]][y + 1:nrow,x + 1:ncol,1]
+  if (randomize)
+    m = m[sample(y + 1:nrow),x + 1:ncol]
+  dim(m) = c(x = nrow, y = ncol) # named dim
+  s[[1]] = m
+  plt(s, 0, add)
+  plt(s, 1, TRUE)
+  plt(s, 2, TRUE)
+  plt(s, 3, TRUE)
+  plt(s, 4, TRUE)
+  plt(s, 5, TRUE)
+  plt(s, 6, TRUE)
[…]

### 2025-03-07 R/layered_map.R

@@ -1,0 +1,195 @@
+# from https://www.urbandemographics.org/post/figures-map-layers-r/
+# BUT see also https://marcosci.github.io/layer/
+
+library("sf")
+library("raster")
+library("stars")
+library("r5r")
+library("geobr")
+library("aopdata")
+library("gtfs2gps")
+library("ggplot2")
+library("osmdata")
+library("h3jsr")
+library("viridisLite")
+library("ggnewscale")
+library("dplyr")
+library("magrittr")
+
+withr::local_options(list(java.parameters = '-Xmx2G'))
+
+rotate_data <- function(data, x_add = 0, y_add = 0) {
+
+  shear_matrix <- function(){ matrix(c(2, 1.2, 0, 1), 2, 2) }
+
+  rotate_matrix <- function(x){
+    matrix(c(cos(x), sin(x), -sin(x), cos(x)), 2, 2)
+  }
+  data %>%
+    dplyr::mutate(
+      geometry = .$geometry * shear_matrix() * rotate_matrix(pi/20) + c(x_add, y_add)
+    )
+}
+
+rotate_data_geom <- function(data, x_add = 0, y_add = 0) {
+  shear_matrix <- function(){ matrix(c(2, 1.2, 0, 1), 2, 2) }
+
+  rotate_matrix <- function(x) {
+    matrix(c(cos(x), sin(x), -sin(x), cos(x)), 2, 2)
+  }
+  data %>%
+    dplyr::mutate(
+      geom = .$geom * shear_matrix() * rotate_matrix(pi/20) + c(x_add, y_add)
+    )
+}
+
+
+
+### get terrain data ----------------
+
+elev <- system.file("extdata/poa/poa_elevation.tif", package = "r5r")
+
+# read terrain raster and calculate hill Shade
+dem <- stars::read_stars(elev)
+dem <- st_as_sf(dem)
+
+# crop
+bbox <- st_bbox(dem)
+
+
+### get public transport network data ----------------
+
+gtfs <- gtfs2gps::read_gtfs( system.file("extdata/poa/poa_eptc.zip", package = "r5r") )
+gtfs <- gtfs2gps::gtfs_shapes_as_sf(gtfs)
+
+# crop
+gtfs <- gtfs[bbox,]
+gtfs <- st_crop(gtfs, bbox)
+plot(gtfs['shape_id'])
+
+
+### get OSM data ----------------
+
+# roads from OSM
+roads <- opq('porto alegre') %>%
+  add_osm_feature(key = 'highway',
[…]

### 2025-03-07 R/prepare_pre_density.R

@@ -3,7 +3,8 @@ library(duckdb)
 library(dplyr, warn.conflicts = FALSE)
 
 withr::local_envvar(c(TZ = "UTC", ORA_SDTZ = "UTC"))
-con <- withr::local_db_connection(DBI::dbConnect(duckdb(), path = ":memory:"))
+# con <- withr::local_db_connection(DBI::dbConnect(duckdb(), path = ":memory:"))
+con <- DBI::dbConnect(duckdb(), path = ":memory:")
 
 dbExecute(con, "SET extension_directory = '/Users/spi/.duckdb/extensions';")
 # dbExecute(con, "INSTALL H3;")

### 2025-03-07 R/quality-checks.R

@@ -1,0 +1,70 @@
+# check frequency of points: 30s?
+# NOT homogeneous!
+
+library(conflicted)
+library(tidyverse)
+library(arrow)
+
+conflicts_prefer(dplyr::filter)
+
+trj <- read_parquet('data/trajectory_2024-08-01.parquet')
+
+occupancy <- trj |>
+  arrange(TIME_OVER) |>
+  mutate(
+    delta = lead(TIME_OVER)- TIME_OVER,
+    .by = c("FLIGHT_ID", "CELL", "HOUR")) |>
+  filter(!is.na(DELTA))
+
+trj |>
+  # duration of each segment inside a CELL in a given HOUR
+  mutate(
+    DELTA = lead(TIME_OVER)- TIME_OVER,
+    .by = c("FLIGHT_ID", "CELL_ID", "HOUR")) |>
+  filter(!is.na(DELTA)) |>
+  group_by(FLIGHT_ID, CELL_ID, HOUR) |>
+  summarise(occupancy = sum(delta)) |>
+  arrange(FLIGHT_ID, desc(CELL_ID), HOUR)
+
+# this proves that TIME_OVER rates are not at 30s for a single flight
+trj |>
+  filter(FLIGHT_ID == 274761935,
+         HOUR == 0) |>
+  arrange(TIME_OVER) |>
+  mutate(DELTA = lead(TIME_OVER) - TIME_OVER) |>
+  filter(!is.na(DELTA))
+
+
+
+
+
+
+library(conflicted)
+library(duckdb)
+library(tidyverse, warn.conflicts = FALSE)
+
+conflict_prefer("filter", "dplyr", quiet = TRUE)
+conflict_prefer("lag", "dplyr")
+conflict_prefer("lead", "dplyr")
+
+
+## Instantiate the in-memory DuckDB connection
+con = dbConnect(duckdb(), shutdown = TRUE)
+
+## Register our parquet dataset as table in our connection (and that assign it
+## to an object that R understands)
+trj = tbl(con, "read_parquet('data/trajectory_2024-08-01.parquet')")
+
+
+trj |>
+  group_by(CELL, HOUR, FLIGHT_ID) |>
+  mutate(NEXT = lead(TIME_OVER)) |>
+  mutate(DELTA = NEXT - TIME_OVER)
+  # duration of each segment inside a CELL in a given HOUR
+  mutate(
[…]

### 2025-03-07 gcat.R

@@ -1,0 +1,98 @@
+library(conflicted)
+library(tidyverse)
+library(here)
+library(fs)
+
+conflicts_prefer(dplyr::filter)
+
+read_catalogue <- function(path) {
+  fn <- fs::path_file(path) |> fs::path_ext_remove()
+  gcat_file <- path
+  gcat_date <- read_lines(gcat_file, skip = 1, n_max = 1) |>
+    str_sub(start = 10L) |>
+    as_datetime(format = "%Y %b  %e  %H%M:%S")
+
+  gcat_col_names <- names(read_tsv(gcat_file, n_max = 0))
+  gcat <- read_tsv(
+    gcat_file,
+    col_names = gcat_col_names,
+    skip = 2,
+    trim_ws = FALSE,
+    quote = ""
+  )
+  gcat |>
+    mutate(
+      catalogue = fn,
+      update_date = gcat_date
+    )
+}
+
+gcat <- dir_ls(here("data-raw"), glob = "*.tsv") |>
+  as_tibble() |>
+  rename(path = value) |>
+  mutate(
+    filename = path_file(path)
+  ) |>
+  filter(filename != "currentcat.tsv") |>
+  pull(path) |>
+  map(.f = read_catalogue) |>
+  bind_rows() |>
+  janitor::clean_names()
+
+
+is_upper_stage <- function(type) {
+  str_detect(type, "R1|R2|R3|R4|R5")
+}
+
+is_rocket_component <- function(type) {
+  str_detect(type, "^C..[A|M].*$")
+}
+
+end_date <- '2024 Mar 01' |> as_datetime()
+start_date <- end_date - ddays(10 * 365.25)
+
+# d_date = phase end time
+gcat_base <- gcat |>
+  # a reentry type
+  filter(status == "R") |>
+  mutate(
+    string_date = str_sub(d_date, 1L, 11L)
+  ) |>
+  mutate(
+    uncertainty = str_sub(d_date, -1L, -1L) == "?",
+    d_date_length = str_length(d_date),
+    # uncertainty = if_else(uncertainty == TRUE,
+    #                       case_when(
+    #                         d_date_length ==
+    #                       )),
+    reentry_date = if_else(is_upper_stage(type), str_sub(d_date, 1L, 11L), NA),
+    object = if_else(is_upper_stage(type), "upper stage", NA)
+  ) |>
+  mutate(
+    reentry_date = if_else(
+      is_rocket_component(type),
+      str_sub(d_date, 1L, 11L),
[…]

### 2025-03-09 R/data-cube.R
cubes plot: increased font size and saved at high res png

@@ -98,9 +98,17 @@ eur_hex <- eur_hex * 6 + c(-8.8, 8.5)
 
 de = de * 6 + c(-8.5, 8.5)
 
-pdf(here("media", "figures", "cubes.pdf"))
+# pdf(here("media", "figures", "cubes.pdf"))
+png(
+  filename = here("media", "figures", "cubes.png"),
+  width = 19,
+  height = 18.5,
+  units = "cm",
+  res = 600
+)
+
 plot.new()
-par(mar = c(5, 0, 5, 0))
+par(mar = c(0, 0, 0, 0), cex = 1.5)
 plot.window(xlim = c(-11, 8), ylim = c(-4, 14.5), asp = 1)
 #
 pl(s, 0, 0, TRUE, randomize = TRUE)
@@ -108,18 +116,18 @@ plot(de, add = TRUE, border = grey(.5))
 plot(eur_hex, add = TRUE, border = grey(.8))
 text(-11, 0, "hour", srt = -90, col = 'black')
 text(3.3, 9.5, "densities", srt = 0, col = 'black')
-text(1.45, 8.6, expression(rho[act]), col = 'black', cex = 1.05)
-text(4.45, 8.6, expression(rho[pop]), col = 'black', cex = 1.05)
+text(1.45, 8.6, expression(rho[act]), col = 'black')
+text(4.45, 8.6, expression(rho[pop]), col = 'black')
 # hours
-text(-10.1, 4, "23", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, 3, "22", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, 2, "21", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, 1, ".", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, 0, ".", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, -1, ".", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, -2, "2", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, -3, "1", srt = -90, col = 'black', cex = 0.7)
-text(-10.1, -4, "0", srt = -90, col = 'black', cex = 0.7)
+text(-10.1, 4, "23", srt = -90, col = 'black', cex = 0.8)
+text(-10.1, 3, "22", srt = -90, col = 'black', cex = 0.8)
[…]

### 2025-03-09 paper.qmd
use png for cubes plot; impact density probability

@@
 
+# Method
+
+From Equation 15 in [@patera2008], the impact (on the ground)
+probability density, $\sigma(L, i)$, at latitude $L$ of a reentring object
+whose orbit has inclination $i$ is given by:
+
+$$ \sigma(L, i) = \frac{1}{2 \pi^2 R^2 \sqrt{(\sin^2(i) - \sin^2(L))}}$$
+
+where $R$ is the Earth's radius.
+
 # Data
@@
 #| echo: false
-knitr::include_graphics("media/figures/cubes.pdf")
+knitr::include_graphics("media/figures/cubes.png")
 ```

### 2025-03-19 R/gcat.R
move under R

@@ -1,0 +1,343 @@
+library(conflicted)
+library(tidyverse)
+library(here)
+library(fs)
+
+conflicts_prefer(dplyr::filter)
+conflicts_prefer(purrr::map)
+
+read_catalogue <- function(path) {
+  fn <- fs::path_file(path) |> fs::path_ext_remove()
+  gcat_file <- path
+  gcat_date <- read_lines(gcat_file, skip = 1, n_max = 1) |>
+    str_sub(start = 10L) |>
+    as_datetime(format = "%Y %b  %e  %H%M:%S")
+
+  gcat_col_names <- names(read_tsv(gcat_file, n_max = 0))
+  gcat <- read_tsv(
+    gcat_file,
+    col_names = gcat_col_names,
+    skip = 2,
+    trim_ws = FALSE,
+    quote = ""
+  )
+  gcat |>
+    mutate(
+      catalogue = fn,
+      update_date = gcat_date
+    )
+}
+
+gcat <- dir_ls(here("data-raw", "gcat"), glob = "*.tsv") |>
+  as_tibble() |>
+  rename(path = value) |>
+  mutate(
+    filename = path_file(path)
+  ) |>
+  filter(filename != "currentcat.tsv") |>
+  pull(path) |>
+  map(.f = read_catalogue) |>
+  bind_rows() |>
+  janitor::clean_names()
+
+
+is_upper_stage <- function(type) {
+  str_detect(type, "R1|R2|R3|R4|R5")
+}
+
+is_rocket_component <- function(type) {
+  str_detect(type, "^C..[A|M].*$")
+}
+
+end_date <- '2024 Mar 01' |> as_datetime()
+end_date <- '2025 Jan 01' |> as_datetime()
+start_date <- end_date - ddays(10 * 365.25)
+start_date <- '2000 Jan 01' |> as_datetime()
+
+# try to determine (vague) d_date uncertainty
+# uncertainty if "\?$"
+
+# the specs say there could be date with hour 24 and minute 60
+# I do not see any occurrences.
+gcat |>
+  select(number_jcat, d_date) |>
+  filter(str_length(d_date) >= 19) |>
+  mutate(
+    hour = str_sub(d_date, 13L, 14L),
+    minute = str_sub(d_date, 15L, 16L),
+    second = str_sub(d_date, 18L, 19L) |> as.numeric()
+  ) |>
+  filter(minute == "60" | hour == "24" | second > 59)
+
+gcat |>
[…]

### 2025-03-19 gcat.R

@@ -1,98 +1,0 @@
-library(conflicted)
-library(tidyverse)
-library(here)
-library(fs)
-
-conflicts_prefer(dplyr::filter)
-
-read_catalogue <- function(path) {
-  fn <- fs::path_file(path) |> fs::path_ext_remove()
-  gcat_file <- path
-  gcat_date <- read_lines(gcat_file, skip = 1, n_max = 1) |>
-    str_sub(start = 10L) |>
-    as_datetime(format = "%Y %b  %e  %H%M:%S")
-
-  gcat_col_names <- names(read_tsv(gcat_file, n_max = 0))
-  gcat <- read_tsv(
-    gcat_file,
-    col_names = gcat_col_names,
-    skip = 2,
-    trim_ws = FALSE,
-    quote = ""
-  )
-  gcat |>
-    mutate(
-      catalogue = fn,
-      update_date = gcat_date
-    )
-}
-
-gcat <- dir_ls(here("data-raw"), glob = "*.tsv") |>
-  as_tibble() |>
-  rename(path = value) |>
-  mutate(
-    filename = path_file(path)
-  ) |>
-  filter(filename != "currentcat.tsv") |>
-  pull(path) |>
-  map(.f = read_catalogue) |>
-  bind_rows() |>
-  janitor::clean_names()
-
-
-is_upper_stage <- function(type) {
-  str_detect(type, "R1|R2|R3|R4|R5")
-}
-
-is_rocket_component <- function(type) {
-  str_detect(type, "^C..[A|M].*$")
-}
-
-end_date <- '2024 Mar 01' |> as_datetime()
-start_date <- end_date - ddays(10 * 365.25)
-
-# d_date = phase end time
-gcat_base <- gcat |>
-  # a reentry type
-  filter(status == "R") |>
-  mutate(
-    string_date = str_sub(d_date, 1L, 11L)
-  ) |>
-  mutate(
-    uncertainty = str_sub(d_date, -1L, -1L) == "?",
-    d_date_length = str_length(d_date),
-    # uncertainty = if_else(uncertainty == TRUE,
-    #                       case_when(
-    #                         d_date_length ==
-    #                       )),
-    reentry_date = if_else(is_upper_stage(type), str_sub(d_date, 1L, 11L), NA),
-    object = if_else(is_upper_stage(type), "upper stage", NA)
-  ) |>
-  mutate(
-    reentry_date = if_else(
-      is_rocket_component(type),
-      str_sub(d_date, 1L, 11L),
[…]

### 2025-03-19 R/scrape_aircraft_info.R
scrape aircraft info, i.e. cruise range and TAS and dimensions

@@ -1,0 +1,331 @@
+# scrape aircraft dimensions
+library(conflicted)
+library(tidyverse)
+library(sf)
+library(arrow)
+library(here)
+library(rvest)
+library(httr)
+library(polite)
+library(eurocontrol)
+
+conflict_prefer("filter", "dplyr", quiet = TRUE)
+
+# circa 15 secs
+trjs <- c(
+  "data/trajectories_2024-08-01_resampled_30s_bbox_res_2.parquet",
+  "data/trajectories_2024-12-05_resampled_30s_bbox_res_2.parquet",
+  "data/trajectories_2024-08-01_resampled_30s_bbox_res_3.parquet",
+  "data/trajectories_2024-12-05_resampled_30s_bbox_res_3.parquet",
+  NULL
+) |>
+  map(.f = read_parquet) |>
+  bind_rows()
+
+ac_types <- trjs |> pull(aircraft_type) |> unique() |> sort() |> setdiff("ZZZZ")
+
+
+scrape_aircraft_type_info_perfdb <- function(ac_type, session) {
+  # For some verbose status updates
+  cli::cli_process_start(
+    "Scrape {.val {ac_type}} at EUROCONTROL Performance DB"
+  )
+  # Create full url and scrape
+  full_url <- polite::nod(
+    session,
+    str_glue(
+      "aircraftperformance/details.aspx?ICAO={ac_type}",
+      ac_type = toupper(ac_type)
+    )
+  )
+  page <- full_url |> polite::scrape()
+
+  icao <- ac_type |> toupper()
+  if (
+    page |> html_text2() |> str_detect("The URL you have used returns no data")
+  ) {
+    # fmt: skip
+    response <- tibble_row(
+      icao           = icao,
+      name           = NA_character_,
+      manufacturer   = NA_character_,
+      type           = NA_character_,
+      wtc            = NA_character_,
+      recat_eu       = NA_character_,
+      mtow_kg        = NA,
+      cruise_tas_kt  = NA,
+      cruise_mach    = NA,
+      cruise_range_nm   = NA,
+      cruise_ceiling_fl = NA,
+      wing_span_m      = NA,
+      length_m         = NA,
+      height_m         = NA,
+      accomodation   = NA_character_
+    )
+  } else {
+    name <- page |>
+      html_elements("#MainContent_wsAcftNameLabel") |>
+      rvest::html_text2()
+
+    manufacturer <- page |>
[…]

### 2025-03-19 R/scrape_airport_passengers.R
scrape airport passengers

@@ -1,0 +1,59 @@
+# To clean data
+library(tidyverse)
+library(lubridate)
+library(janitor)
+library(here)
+
+# To scrape data
+library(rvest)
+library(httr)
+library(polite)
+library(magrittr)
+
+
+url <- "https://en.wikipedia.org/wiki/List_of_the_busiest_airports_in_Europe#2024"
+url_bow <- polite::bow(url)
+
+
+eu_pass <- url_bow |>
+  polite::scrape() |>
+  rvest::html_nodes("table.wikitable") |>
+  rvest::html_table(header = TRUE, trim = TRUE, dec = ".") |>
+  simplify() |>
+  extract2(1) |>
+  clean_names() |>
+  slice(-1) |>
+  select(-2) |>
+  select(1:5) |>
+  mutate(
+    rank2024 = as.integer(rank2024),
+    passengers = str_remove(passengers, "\\[\\d*\\]$"),
+    passengers = str_remove_all(passengers, ","),
+    passengers = as.numeric(passengers),
+    NULL
+  ) |>
+  mutate(
+    year = 2024,
+    passengers = if_else(city_served == "Toulouse", 7844953, passengers),
+    passengers = if_else(
+      airport == "Brussels South Charleroi Airport",
+      10500000,
+      passengers
+    ),
+    # keep same number as 2023
+    passengers = if_else(
+      airport == "Bordeaux–Mérignac Airport",
+      6584194,
+      passengers
+    ),
+    passengers = if_else(
+      airport == "Beauvais–Tillé Airport",
+      5638955,
+      passengers
+    ),
+    NULL
+  ) |>
+  rename(rank = rank2024)
+
+eu_pass |>
+  write_csv(here("data", "airport_passengers.csv"))

### 2025-03-19 R/density-map.R
make empty hexes transparent; add background color for sea and land

@@ -7,6 +7,7 @@ library(ggplot2)
 library(h3jsr)
 library(magrittr)
 library(giscoR)
+library(grid)
 
 conflicts_prefer(dplyr::filter)
 
@@ -14,11 +15,11 @@ source("R/cells.R")
 
 density <- arrow::read_parquet("data/flight_density.parquet")
 
-res  <- 3
+res <- 3
 YYYY <- 2024
-MM   <- 8
-DD   <- 1
-HH   <- 10
+MM <- 8
+DD <- 1
+HH <- 10
 
 eur_hex <- bbox_nm() |>
   hexes_for_bbox_at_res(resolution = res)
@@ -30,7 +31,8 @@ ddd <- density |>
     month == MM,
     day == DD,
     hour == HH,
-    h3_resolution == res) |>
+    h3_resolution == res
+  ) |>
   mutate(occupancy = mean(occupancy), .by = cell) |>
   filter(cell %in% (eur_hex |> pull(h3_address))) |>
   select(cell, everything()) |>
@@ -41,7 +43,6 @@ min <- ddd |> pull("occupancy") |> min()
 min <- 0
 
 
-
 eur_hex_union <- bbox_nm() |>
   bbox_of_hexes_for_bbox_at_res(resolution = res)
 
@@ -49,7 +50,8 @@ gisco_RG <- gisco_get_countries(
   resolution = "20",
   epsg = "4326",
   spatialtype = "RG",
-  region = c("Europe", "Africa", "Asia")) |>
+  region = c("Europe", "Africa", "Asia")
+) |>
   mutate(res = "20M")
 
 gisco <- gisco_RG
@@ -58,39 +60,134 @@ EUR_res20 <- gisco |>
   st_intersection(eur_hex_union) |>
   select(res)
 
+# draw hexagon for key shape in legend
+draw_key_hex <- function(data, params, size) {
+  # make fill inherit color if NA
+  if (is.na(data$fill)) data$fill <- data$col
+
+  # hexagon vertex coordinates
+  v <- list(
+    x = c(0.95, 0.725, 0.275, 0.05, 0.275, 0.725),
+    y = c(
+      0.5,
+      0.110288568297003,
+      0.110288568297003,
+      0.5,
+      0.889711431702997,
+      0.889711431702997
+    )
+  )
+  # hexagon grob
+  grid::polygonGrob(
+    v$x,
+    v$y,
[…]

### 2025-03-19 paper.qmd
use new figure and add formula

@@
 ```{r}
+#| label: tbl-resolutions
+#| tbl-cap: H3 hexagons, average values
+#| tbl-cap-location: bottom
+
 library(tidyverse)
@@
   4L,   1770.347654491,  26.07175968
-
 ) |> 
@@
     resolution ~ "Resolution",
-    tidyselect::ends_with("area") ~ "Cell Area {{*persons* km^-2}}",
-    # hex_area ~ md("Cell Area ({km^2}})"),
-    hex_edge ~ "Edge length ({km}})"
+    hex_area ~ md("Cell Area"),
+    hex_edge ~ md("Edge length")
   ) |> 
@@
     subtitle = md("average hexagons")
+  ) |> 
+  cols_units(
+    hex_area = "km^2",
+    hex_edge = "km"
   ) |>
@@
 
-@fig-density-winter-res-2 provides a promising result in the hourly flight
+@fig-density-winter-res-3 provides a promising result in the hourly flight
 density calculation which at least shows
@@
 
-::: {#fig-density-winter-res-2}
+::: {#fig-density-winter-res-3}
 
-```{r density-winter-res-2}
-#| out-width: "100%"
+```{r density-winter-res-3}
+#| out-width: "110%"
 #| echo: false
-knitr::include_graphics("media/figures/flight_density_2024-08-01T12_2.png")
+knitr::include_graphics("media/figures/flight_density_2024-08-01T12_3.png")
 ```
 
-Flight density on 2024-08-01T12 for H3 cells at resolution 2 covering EUROCONTROL Area.
+Flight density in the EUROCONTROL Area on 2024-08-01T12 (H3 cells at resolution 3).
 :::
@@
 
-From Equation 15 in [@patera2008], the impact (on the ground)
-probability density, $\sigma(L, i)$, at latitude $L$ of a reentring object
-whose orbit has inclination $i$ is given by:
+## Impact probability density
+
+From Equation 15 in [@patera2008], the (on the ground) impact probability
+density, $\sigma(L, i)$, at latitude $L$ of a reentring object
+whose orbit has inclination $i$ (see @fig-orbital-inclination) is given by:
 
@@
 where $R$ is the Earth's radius.
+
+::: {#fig-orbital-inclination}
+
+```{r orbital-inclination}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/figures/orbital_inclination_clip.png")
+```
+
+Orbital inclination is the angle between the plane of an orbit and the equator.
+An orbital inclination of 0° is directly above the equator, 90° crosses right above the pole, and 180° orbits above the equator in the opposite direction of Earth’s spin.
+(NASA illustration by Robert Simmon.).
+:::
+
+## Aircraft effective exposed area
+From Patera
+
+> Even for a ballistic coefficient of 50 lbs/ft2, the flight path angle
+> of a reentering object at 40,000 feet is -86.6 degrees, which corresponds
+> to nearly vertical descent.
+> Since most debris objects have ballistic coefficients equal to or 
+> less than 50 lbs/ft2, the vertical descent assumption is a very
+> accurate approximation.
+
+$$
+\varepsilon_i = \frac{s_{a} a_{h} + s_{d} a_{v}}{s_d}
+$$
+where $s_{a}$ is the average aircraft speed, $s_{d}$ is the average debris 
+object fall speed, $a_v$ area of the aircraft as viewed from above,
+$a_h$ area of the aircraft as viewed from the from the front of the aircraft.
 

### 2025-03-21 R/act_passengers.R
add seats numbers

@@ -1,0 +1,227 @@
+# fmt: skip
+tribble(
+  ~icao, ~crew, ~pax_min, ~pax_typical, ~pax_max, ~accomodation,
+  "A124", 4,  88,  88,  88, "Flightcrew of 6 (later versions 4) and up to 88 passengers in upper deck and up to 150t payload.",
+  "A19N", 2, 120, 150, 160, "Flight crew of two and typical 120-150 passengers in two classes, max 160 in high density configuration.",
+  "A20N", 2, 150, 180, 194, "Flight crew of two and typical 150-180 passengers in two classes, max 194 in high density configuration",
+  "A21N", 2, 180, 220, 244, "Flight crew of two and typical 180-220 passengers in two classes, max 244 in high density configuration",
+  "A306", 2, 266, 266, 298, "Flightcrew of 2 and 266 passengers in typical seating configuration, up to 298 in single class configuration.",
+  "A310", 2, 220, 220, 247, "Flightcrew of 2 and 220 passengers in typical seating configuration, up to maximum of 247.",
+  "A318", 2, 107, 107, 117, "Flightcrew of two and typical 107 passengers in two classes, up to 117 in single class configuration.",
+  "A319", 2, 124, 124, 142, "Flightcrew of two and typical 124 passengers in two classes, up to 142 in high density configuration.",
+  "A320", 2, 150, 150, 180, "Flightcrew of two and typical 150 passengers in two classes, up to 180 in a one class layout.",
+  "A321", 2, 185, 185, 220, "Flightcrew of two and typical 185 passengers in two classes, up to 220 in all economy high density configuration.",
+  "A332", 2, 253, 253, 293, "Flightcrew of two and typical 253 passengers in three classes, up to 293 in two classes configuration.",
+  "A333", 2, 295, 295, 335, "Flightcrew of two and typical 295 passengers in three classes, up to 335 in high density configuration.",
+  "A338", 2, 220, 260, 406, "Flight crew of 2 and typical 220-260 passengers in three classes, max 406 in high density configuration.",
+  "A339", 2, 260, 300, 440, "Flight crew of 2 and typical 260-300 passengers in three classes, max 440 in high density configuration.",
[…]

### 2025-03-21 R/scrape_aircraft_info.R

@@ -309,8 +309,11 @@ all_acts <- ac_types |>
   map(.f = scraper) |>
   bind_rows()
 
-all_acts |>
-  write_csv("all_acts.csv")
+source("R/act_passengers.R")
+
+all_acts <- all_acts |>
+  left_join(act_all_pax) |>
+  write_csv("data/all_acts.csv")
 
 
 # ---- SkyBrary aircraft pages ----

### 2025-04-06 paper.qmd
removed old orbital figures; regenerate data cubes figure; handle SI units properly

@@
     keep-tex: true
+    # fig-pos: 'H'
     journaltype: conference
     papersize: a4
-  ieee-html: default
+    include-in-header:
+      - file: media/preamble.tex
+    include-before-body:
+      - text: |
+          % NOTE: make siunitx's \qty work without clashing with the one in physics
+          \let\qty\SI
+  ieee-html: 
+    html-math-method: mathjax
+    include-in-header:
+      file: media/siunitx.html
 
@@
 
-We base our hourly flight density estimation on (enhanced) trajectories flown
+We base our hourly flight density estimation on trajectories flown
 by \acr{IFR} commercial aircraft as
 recorded by EUROCONTROL's \acr{NM}[^1] and overlapped on Uber's H3
-gridding system at different resolutions [@h3resolutions], see @fig-cells 
-for resolution 2 which corresponds to hexagons of average area $86801.78\ km^2$
-and average edge length of $182.51 km$, see also table
+gridding system at different resolutions [@h3resolutions].
+@fig-cells shows cells for resolution $2$ which corresponds to hexagons
+of average area $\qty{86801.78}{km^2}$ and average edge length of 
+$\qty{182.51}{km}$, see also @tbl-resolutions for other resolutions.
 
@@
 #| echo: false
+#| fig-scap: "H3 cells at resolution 2"
 knitr::include_graphics("media/figures/cells_2.png")
@@
 
-Flight density in the EUROCONTROL Area on 2024-08-01T12 (H3 cells at resolution 3).
+Flight density in the EUROCONTROL Area on 2024-08-01T12 (H3 cells at resolution $3$).
 :::
@@
 
+The probability of impact for debris object $k$, $p_k$, is expressed as.
+
+$$
+p_k = \varepsilon_k \int_a \rho(\vec{x}) \sigma(\vec{x}) da
+$$ {#eq-impact-probability}
+
+where $\rho({\vec{x}})$ is the probability density of an aircraft
+and $\sigma(\vec{x})$ is probability density of a debris object in the
+position $\vec{x}$.
+
 ## Impact probability density
 
-From Equation 15 in [@patera2008], the (on the ground) impact probability
-density, $\sigma(L, i)$, at latitude $L$ of a reentring object
-whose orbit has inclination $i$ (see @fig-orbital-inclination) is given by:
+Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
+circular orbits prior to reentry even if initially on an elliptical ones.
+In fact drag at perigee will lower apogee and similarly drag at apogee will
+lower perigee. But because drag at perigee is higher due to denser atmosphere,
+the apogee  will decrease faster than perigee and the orbit will then tend to be
+a circular one.
+The angular position of the reentry point from the ascending node, 
+see @sec-orbitals, is uniformly distributed, $\sigma(\theta) = 1 / (2\pi)$,
+due to uncertainties in atmospheric drag and the gravitational effects of
+the Equatorial bulge, hence we assume that the impact longitude, $\lambda$,
+is uniformely distributed $\sigma(\lambda) = 1/(2\pi)$.
 
-$$ \sigma(L, i) = \frac{1}{2 \pi^2 R^2 \sqrt{(\sin^2(i) - \sin^2(L))}}$$
+The impact probability density, $\sigma(\varphi, i)$,
+at latitude $\varphi$ of a reentring object whose orbit has inclination $i$
+(see @fig-orbital-elements) is given by [[Equation 15 in @patera2008]]:
 
-where $R$ is the Earth's radius.
+$$ \sigma(\varphi, i) = \frac{1}{2 \pi^2 R^2 \sqrt{(\sin^2(i) - \sin^2(\varphi))}}$$ {#eq-impact-probability-density}
 
-::: {#fig-orbital-inclination}
+So for example [@patera2008], if a building is located at latitude 30 degrees
+north (or south) and has an area of $\qty{1000}{m^2}$ and a piece of debris
+reenters the Earth’s atmosphere from an orbit having an inclination of 45 degrees,
+the impact probability is given by
 
-```{r orbital-inclination}
-#| out-width: "100%"
-#| echo: false
-knitr::include_graphics("media/figures/orbital_inclination_clip.png")
-```
+$$\begin{aligned}
+P &= \Delta A \times \sigma(\varphi, i) = 1000 \times \sigma\Bigl(\frac{\pi}{6}, \frac{\pi}{4}\Bigl) \\
+  &= 1000 \times \num{2.490756e-15} = \num{2.490756e-12}
+\end{aligned}$$
 
-Orbital inclination is the angle between the plane of an orbit and the equator.
-An orbital inclination of 0° is directly above the equator, 90° crosses right above the pole, and 180° orbits above the equator in the opposite direction of Earth’s spin.
-(NASA illustration by Robert Simmon.).
-:::
+where we used the equatorial value of $\qty{6378}{km}$, for the Earth's radius,
+$R$, as the distance of the building from the center of the planet.
+
+$\num{0.046}$
 
@@
 
+The effective expose area, $\varepsilon_i$, is
 $$
@@
 ```{r densities}
-#| out-width: "120%"
+#| out-width: "100%"
 #| echo: false
@@
 
-Flight and population densities data (hex on Germany as an example).
+Data cube of hourly flight densities data over H3 hexes (in Germany as an example) for multiple dates, $d_1 \cdots d_n$. 
 :::
@@
 
-# References
+# Appendix {.appendix}
+
+## Orbital Elements {#sec-orbitals .appendix}
+We refer to @fig-orbital-elements to describe the orbital parameters of
+interest for the current work.
+
+:::{#fig-orbital-elements layout="[40,-3,57]" fig-env="figure*"}
+
+![Keplerian elements: semi-major axis, $a$, the eccentricity, $e = \frac{2c}{2a}$.
+(the semiparameter (a.k.a.
+_semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as
+$p = b^2 / a = a (1 - e^2)$)](media/figures/classical_orbital_elements_2.png)
+
+![Keplerian elements: inclination, $i$, the right ascension of the ascending node, $\Omega$ , and the argument of periapsis, $\omega$.(Earth rotates around $\hat{K}$)](media/figures/classical_orbital_elements_4.png)
+
+Definition of Keplerian elements.
+The orbital elements describing the trajectory of orbiting object are illustrated in association with an inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$.
+The $I$-axis, $\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$, is perpendicular to the $I$-axis in the equatorial plane n the sense of Earth's rotation with the $K$-axis pointing up.
+The semi-major axis, $a$, the eccentricity, $e$, the inclination, $i$, the right ascension of the ascending node, $\Omega$ , and the argument of perigee, $\omega$, constitute, together with the mean anomaly at the initial position, $M_0$, a set of six constants of motion that are preserved in the absence of perturbation.
+True anomaly, $\nu$, specifies the location of an orbiting object within the orbit.
+It is the angle between perigee and the orbiting object’s position vector measured in the direction of the orbiting object’s motion.
+Figures from [@sellers2003understanding], respectively Figure 5-2 and Figure 5-9.
+
+:::
+
+The angular position, $\theta$, of the reentry point from the ascending node
+in the orbit plane is $\theta = \omega + \nu$.
+It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
+$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or
+
+$$\sin(\varphi) = \sin(\theta)\sin(i)$$ {#eq-spheric-coords}
+
+Mass of Earth, $M_{\oplus} = \qty{5.9722e24}{kg}$.
+
+# References {.unnumbered}
 

### 2025-04-06 R/data-cube.R

@@ -9,7 +9,7 @@ tif <- system.file("tif/L7_ETMs.tif", package = "stars")
 r <- read_stars(tif)
 
 
-nrow <- 2
+nrow <- 3
 ncol <- 8
 m <- matrix(runif(nrow * ncol), nrow = nrow, ncol = ncol)
 m <- r[[1]][1:nrow, 1:ncol, 1]
@@ -101,7 +101,7 @@ de = de * 6 + c(-8.5, 8.5)
 # pdf(here("media", "figures", "cubes.pdf"))
 png(
   filename = here("media", "figures", "cubes.png"),
-  width = 19,
+  width = 19.5,
   height = 18.5,
   units = "cm",
   res = 600
@@ -115,9 +115,10 @@ pl(s, 0, 0, TRUE, randomize = TRUE)
 plot(de, add = TRUE, border = grey(.5))
 plot(eur_hex, add = TRUE, border = grey(.8))
 text(-11, 0, "hour", srt = -90, col = 'black')
-text(3.3, 9.5, "densities", srt = 0, col = 'black')
-text(1.45, 8.6, expression(rho[act]), col = 'black')
-text(4.45, 8.6, expression(rho[pop]), col = 'black')
+text(4.5, 9.5, "date", srt = 0, col = 'black')
+text(1.45, 8.6, expression(d[1]), col = 'black')
+text(4.45, 8.6, "...", col = 'black')
+text(7.45, 8.6, expression(d[n]), col = 'black')
 # hours
 text(-10.1, 4, "23", srt = -90, col = 'black', cex = 0.8)
 text(-10.1, 3, "22", srt = -90, col = 'black', cex = 0.8)

### 2025-04-06 media/siunitx.js

@@ -1,0 +1,1 @@
+(function(){"use strict";var __webpack_modules__={660:function(e,n,t){t.d(n,{W:function(){return s}});var r=t(516),i=JSON.parse('{"Bi":"Invalid control sequence at the end of the %1 option.","lJ":"A comparator has already been set.  %1 is already present, but you\'re trying to add %2.","L$":"Couldn\'t parse %1.  Option \\"exponent-threshold\\" must be in the form of \\"min-number:max-number\\".","xg":"There are two many semi-colons.  Should only be two at most.","Gx":"%1 contains invalid characters and cannot be parsed by siunitx for the ang macro.","Kv":"%1 contains invalid characters and cannot be parsed by siunitx for the num macro.","sT":"Key %1 accepts only values of type %2.","kq":"%1 is detected as a literal unit.  These are forbidden globally. Use interpreted units instead (i.e. \\\\kilo\\\\gram).","Eh":"The unit macro, %1, has not been defined.","mu":"Trying to close an uncertainty that doesn\'t exist.","P4":"Uncertainty was already closed.","qq":"Couldn\'t parse %1 using per-mode option of %2."}');function a(e,n){for(var t=0;t<n.length;t++){var r=n[t];r.enumerable=r.enumerable||!1,r.configurable=!0,"value"in r&&(r.writable=!0),Object.defineProperty(e,r.key,r)}}function o(e,n,t){return n&&a(e.prototype,n),t&&a(e,t),Object.defineProperty(e,"prototype",{writable:!1}),e}var s=o((function e(){!function(e,n){if(!(e instanceof n))throw new TypeError("Cannot call a class as a function")}(this,e)}));s.BadOptionChars=function(e){return new r.Z("siunitx:badOptionChars",i.Bi,e)},s.ComparatorAlreadySet=function(e,n){return new r.Z("siunitx:comparatorAlreadySet",i.lJ,e,n)},s.ExponentThresholdsError=function(e){return new r.Z("sinunitx:exponentThresholdsError",i.L$,e)},s.ExtraSemicolon=function(){return new r.Z("siunitx:extraSemicolon",i.xg)},s.InvalidAngArgument=function(e){return new r.Z("siunitx:invalidAngArgument",i.Gx,e)},s.Inva
[…]

### 2025-04-06 R/gcat.R
finalize and save on CSV

@@ -311,33 +311,15 @@ gcat_base |>
   count()
 
 # keep the objects that reentered between start_date and end_date
-gcat_base |>
+gcat <- gcat_base |>
   filter(!(til < start_date), !(end_date < wef)) |>
   rename(inclination = `inc`) |>
   mutate(
     inclination = as.numeric(inclination),
     inclination = if_else(inclination < 0, 180 - inclination, inclination),
     NULL
-  )
+  ) |>
+  select(jcat, piece, type, inclination, reentry_date, dry_mass)
 
-# ---- weighting function ----
-# RE=6378000 # equatorial radius in m
-#
-# weighting_function = np.zeros(360)
-# num_of_satellites = len(RBGCAT)
-# timer = 0
-#
-# for line in RBGCAT:
-#   inclination = int(float(line[36]))
-# drymass = line[20]
-# if inclination > 90:
-#   inclination = 180 - inclination
-#
-# vals, lats = cs.latWeights(0.5, 550e3+RE, inclination) # get latitude weights
-#
-# weighting_function += vals # add the normalised times
-#
-# timer +=1
-# print("Working on satellite:", timer, "of", num_of_satellites)
-#
-# np.savetxt('Output data/10 year weighting function.csv', weighting_function)
+gcat |>
+  write_csv(here("data", "gcat.csv"))

### 2025-04-06 R/orbitals.R
keplerian orbits helper

@@ -1,0 +1,97 @@
+#' convert from orbital elements to cartesian coordinates
+#'
+#' @param nu the true anomaly, i.e. the angular displacement measured from
+#'           periapsis to the position vector along the direction of motion
+#' @param a  the semimajor axis of the orbit
+#' @param e  the eccentricity of the orbit
+#' @param i the inclination of the orbit
+#' @param Omega the Right ascension of the ascending node
+#' @param w the argument of perigee
+#' @param M the mass of the first body (more massive one), i.e. Earth
+#' @param m the mass of the orbiting body (less massive), i.e. satellite.
+#'          In the case of m << M, m can be passed as 0
+#'
+#' @returns a vector with I, J, K, Vi, Vj, Vk of the values of position and
+#'          velocity in a geocentric-equatorial reference system
+#'
+#' @examples
+#' \dontrun{
+#' rE  <- 6378e3 # Earth equatorial radius
+#' alt <- 550e3 # altitude of satellite, i.e. 500 km for LEO
+#' ecc <- 0 # circular orbit, i.e. when uncontrolled reentry of debris
+#' m   <- 0 # mass of debris/satellite << mass Earth, so it can be 0
+#' o   <- 0
+#' w   <- 0
+#' i <- 10
+#' kepler_to_cartesian(
+#'   nu = 0.5,
+#'   a = rE + alt,
+#'   e = ecc,
+#'   i = i, Omega = o,
+#'   w = w,
+#'   m = m)
+#' }
+kepler_to_cartesian <- function(
+  nu,
+  a,
+  e,
+  i,
+  Omega,
+  w,
+  M = 5.9722e24, # mass of the major body, i.e. Earth
+  m = 0 # mass of the other body, i.e. satellite or debris
+) {
+  G <- 6.67430e-11 # gravitational constant
+  mu <- G * (M + m) #standard gravitational parameter
+
+  p <- a * (1 - e^2) # semiparameter (semilatus rectum)
+  r_0 = p / (1 + e * cos(nu))
+
+  #--------------- Coordinates in the perifocal reference system Oxyz -----------------
+  # position vector coordinates
+  x <- r_0 * cos(nu)
+  y <- r_0 * sin(nu)
+  # velocity vector coordinates
+  Vx_ <- -sqrt(mu / p) * sin(nu)
[…]

### 2025-04-06 R/scrape_aircraft_info.R
scrape aircraft performance info and save CSV

@@ -8,6 +8,8 @@ library(rvest)
 library(httr)
 library(polite)
 library(eurocontrol)
+library(purrr)
+library(rlang)
 
 conflict_prefer("filter", "dplyr", quiet = TRUE)
 
@@ -25,6 +27,7 @@ trjs <- c(
 ac_types <- trjs |> pull(aircraft_type) |> unique() |> sort() |> setdiff("ZZZZ")
 
 
+# ---- EUROCONTROL Perfromance DB ----
 scrape_aircraft_type_info_perfdb <- function(ac_type, session) {
   # For some verbose status updates
   cli::cli_process_start(
@@ -149,7 +152,41 @@ scrape_aircraft_type_info_perfdb <- function(ac_type, session) {
   response
 }
 
+# Create a polite session
+host <- "https://contentzone.eurocontrol.int/"
+session <- polite::bow(host, force = TRUE)
+scraper <- purrr::partial(scrape_aircraft_type_info_perfdb, session = session)
 
+# some aircraft types will be missing, i.e. helicopters
+all_acts <- ac_types |>
+  map(.f = scraper) |>
+  bind_rows()
+
+source("R/act_passengers.R")
+
+all_acts_missing <- all_acts |>
+  filter(if_all(c(-icao), is.na)) |>
+  pull(icao)
+
+all_acts <- all_acts |>
+  filter(!icao %in% all_acts_missing) |>
+  mutate(across(
+    c(
+      "mtow_kg",
+      "cruise_tas_kt",
+      "cruise_mach",
+      "cruise_range_nm",
+      "cruise_ceiling_fl"
+    ),
+    as.numeric
+  )) |>
+  left_join(act_all_pax) |>
+  mutate(pax = crew + pax_max) |>
+  select(-c("accomodation", "crew", "pax_min", "pax_typical", "pax_max")) |>
+  write_csv("data/ectrl_acts.csv")
+
+
+# ---- SkyBrary aircraft pages ----
 scrape_aircraft_type_info_skybrary <- function(ac_type, session) {
   make_valid <- function(value) {
     value <- ifelse(
@@ -211,7 +248,8 @@ scrape_aircraft_type_info_skybrary <- function(ac_type, session) {
     path <- ".field-name-field-aircraft-type-code.field-type-entity-reference.field-label-above.has-single > div.field-items > div"
     type <- page |>
       html_elements(path) |>
-      rvest::html_text2()
+      rvest::html_text2() |>
+      make_valid()
 
[…]

### 2025-04-06 R/aircraft_types.R
consolidate scraped aircraft types in one dataset and save as CSV

@@ -1,0 +1,32 @@
+library(tidyverse)
+
+ac_types <- c(
+  "data/ectrl_acts.csv",
+  "data/sky_acts.csv",
+  "data/doc8643_acts.csv"
+) |>
+  purrr::map(.f = function(fn) {
+    # fmt: skip
+    cols <- cols(
+      icao              = col_character(),
+      name              = col_character(),
+      manufacturer      = col_character(),
+      type              = col_character(),
+      wtc               = col_character(),
+      recat_eu          = col_character(),
+      mtow_kg           = col_double(),
+      cruise_tas_kt     = col_double(),
+      cruise_mach       = col_double(),
+      cruise_range_nm   = col_double(),
+      cruise_ceiling_fl = col_double(),
+      wing_span_m       = col_double(),
+      length_m          = col_double(),
+      height_m          = col_double(),
+      pax               = col_double()
+    )
+    read_csv(fn, col_types = cols)
+  }) |>
+  bind_rows()
+
+ac_types |>
+  write_csv("data/aircraft_type_info.csv")

### 2025-04-06 R/act_passengers.R
passengers info per aircraft type

@@ -225,3 +225,31 @@ tribble(
   "TOBA", 1,   3,   3,   3, "One pilot and 3 passengers.",
   "TOR",  2,   0,   0,   0, "Flightcrew of two."
 ) -> act_all_pax
+
+
+# fmt: skip
+tribble(
+  ~icao, ~crew, ~pax_min, ~pax_typical, ~pax_max, ~accomodation,
+  "A139", 2,  0,  0,  15, "Crew: one or two; Capacity: up to 15 passengers, depending on configuration",
+  "A169", 2,  0,  8,  12, "Crew: 1 or 2; Capacity: 8 to 12 passengers, plus 250 kg (551 lbs) baggage and cargo; or 1 stretcher, plus 7 medical attendants; or 2 stretchers, plus 5 medical attendants",
+  "A189", 2,  0,  0,  19, "Crew: 1 or 2; Capacity: Up to 19 passengers",
+  "A337", 3,  0,  0,   0, "Crew: 2 pilots and 1 loadmaster",
+  "AC90", 2,  0,  0,   4, "Crew: Two; Capacity: four passengers",
+  "AS50", 1,  0,  0,   5, "Crew: 1; Capacity: 5 pax",
+  "AS65", 2,  0,  0,  10, "Crew: 1 or 2 pilots; Capacity: 10 troops",
+  "AT46", 2,  0,  0,  48, "Crew: 2 pilots",
+  "AT73", 4,  0,  0,  66, "",
+  "AT76", 4, 68, 68,  74, "",
+  "B429", 1,  0,  7,   7, "Crew: 1 pilot; Capacity: 7 passengers (six in passenger compartment; one beside pilot)",
+  "BE30", 2,  0, 10,  11, "",
+  "BER2", 5,  0, 44,  72, "Crew: 5; Capacity: 44 pax Be-200ES / 72 pax Be-210",
+  "BN2P", 1,  0,  9,   9, "Crew: 1; Capacity: 9 passengers",
+  "C10T", 1,  0,  5,   5, "Crew: 1; Capacity: 5 passengers",
+) -> sky_pax
+
+
+# fmt: skip
+tribble(
+  ~icao, ~crew, ~pax_min, ~pax_typical, ~pax_max, ~accomodation,
+  "AT8T", 2,  0,  1,  2, "Crew: One or Two (AT-802A – single-seat)",
+) -> doc8643_pax

### 2025-04-11 R/gcat.R
extract correct set of debris

@@ -28,14 +28,11 @@ read_catalogue <- function(path) {
     )
 }
 
-gcat <- dir_ls(here("data-raw", "gcat"), glob = "*.tsv") |>
-  as_tibble() |>
-  rename(path = value) |>
-  mutate(
-    filename = path_file(path)
-  ) |>
-  filter(filename != "currentcat.tsv") |>
-  pull(path) |>
+# NOTE: use only
+# * satcat	Standard Satellite Catalog (S) Objects from the US Satellite Catalog
+# * auxcat	Auxiliary Satellite Catalog	(A) Objects omitted from the US catalog
+# * tmpcat	Temporary Catalog	(T)	Objects expected to receive an S catalog number soon
+gcat <- here("data-raw", "gcat", c("satcat.tsv", "auxcat.tsv", "tmpcat.tsv")) |>
   map(.f = read_catalogue) |>
   bind_rows() |>
   janitor::clean_names()
@@ -49,9 +46,8 @@ is_rocket_component <- function(type) {
   str_detect(type, "^C..[A|M].*$")
 }
 
-end_date <- '2024 Mar 01' |> as_datetime()
+# 25 years of reentries
 end_date <- '2025 Jan 01' |> as_datetime()
-start_date <- end_date - ddays(10 * 365.25)
 start_date <- '2000 Jan 01' |> as_datetime()
 
 # try to determine (vague) d_date uncertainty
@@ -69,148 +65,130 @@ gcat |>
   ) |>
   filter(minute == "60" | hour == "24" | second > 59)
 
-gcat |>
-  select(number_jcat, d_date) |>
-  mutate(
-    d_date_uncertainty_flag = str_detect(d_date, "\\?$"),
-    d_date_uncertainty_mult = case_when(
-      str_detect(d_date, "s\\?$") ~ 30,
-      str_detect(d_date, "s$") ~ 10,
-      str_detect(d_date, "\\?$") ~ 3,
-      .default = 1
-    )
-  ) |>
-  filter(str_detect(d_date, "s\\?$")) |>
-  mutate(dd = as_datetime(d_date)) |>
-  filter(is.na(dd) | d_date_uncertainty_flag)
 
-# fmt: skip
-vdate_test <- tribble(
-  ~precision,                          ~vdate, ~range,  ~width,
[…]

### 2025-04-11 R/orbitals.R
add latitude weights

@@ -1,8 +1,8 @@
-#' convert from orbital elements to cartesian coordinates
+#' convert from orbital elements to Cartesian coordinates
 #'
 #' @param nu the true anomaly, i.e. the angular displacement measured from
 #'           periapsis to the position vector along the direction of motion
-#' @param a  the semimajor axis of the orbit
+#' @param a  the semi-major axis of the orbit
 #' @param e  the eccentricity of the orbit
 #' @param i the inclination of the orbit
 #' @param Omega the Right ascension of the ascending node
@@ -44,7 +44,7 @@ kepler_to_cartesian <- function(
   G <- 6.67430e-11 # gravitational constant
   mu <- G * (M + m) #standard gravitational parameter
 
-  p <- a * (1 - e^2) # semiparameter (semilatus rectum)
+  p <- a * (1 - e^2) # semi-parameter (semi-latus rectum)
   r_0 = p / (1 + e * cos(nu))
 
   #--------------- Coordinates in the perifocal reference system Oxyz -----------------
@@ -77,21 +77,82 @@ kepler_to_cartesian <- function(
   c(i = I, j = J, k = K, vi = Vi, vj = Vj, vk = Vk)
 }
 
-# test
-# rE <- 6378e3 # Earth equatorial radius
-# alt <- 550e3 # altitude of satellite, i.e. 500 km for LEO
-# ecc <- 0 # circlar orbit when uncontrolled reentry of debris
-# mSat <- 0 # mass of debris/satellite << mass Earth, so it can be 0
-# Omega <- 0
-# w <- 0
-# i <- 10
-#
-# kepler_to_cartesian(
-#   nu = 0.5,
-#   a = rE + alt,
-#   e = 0,
-#   i = i,
-#   Omega = Omega,
-#   w = w,
-#   m = mSat
-# )
+
+#' latitude weight for a space object circularly orbiting at `inclination`
+#'
+#' @param delta_lat delta altitude [decimal degrees]
+#' @param altitude  altitude of the orbiting space object [m]
+#' @param inclination_deg inclination [decimal degrees]
+#' @param n number of samples of the orbit
+#'
+#' @returns a tibble of latitudes and values for the density
+#'
+#' @examples
+#' \dontrun{
+#' rE  <- 6378e3 # Earth equatorial radius
[…]

### 2025-04-11 R/weighting_function.R
calculate weighting function

@@ -1,0 +1,69 @@
+# ---- weighting function ----
+# build reentry probability density from past debris reentry episodes
+library(conflicted)
+library(tidyverse)
+library(magrittr)
+
+conflicts_prefer(dplyr::filter)
+rE <- 6378e3 # Earth equatorial radius
+alt <- 550e3 # altitude of satellite, i.e. 500 km for LEO
+
+source(here::here("R", "orbitals.R"))
+
+gcat <- read_csv("data/gcat.csv") |>
+  # not strictly needed, because there are unique rows
+  filter(row_number() == 1, .by = c("jcat", "piece")) |>
+  select(jcat, piece, inclination, reentry_date) |>
+  mutate(
+    id = str_c(jcat, "_", piece)
+  )
+
+fff <- purrr::partial(
+  latitude_weights,
+  delta_lat = 0.5,
+  altitude = rE + alt
+)
+
+gcat_orig <- read_csv(
+  "../AirspaceDebris/Output data/10_year_reentries.csv",
+  col_names = FALSE
+) |>
+  select(jcat = X1, piece = X3, inclination = X37, reentry_date = X11) |>
+  mutate(
+    reentry_date = str_sub(reentry_date, start = 1L, end = 12L) |> as_date(),
+    # normalize latitude to [-90, 90] range
+    inclination = ((inclination + 90) %% 180) - 90,
+    id = str_c(jcat, "_", piece)
+  )
+
+
+uuu <- gcat |>
+  # filter(id %in% (gcat_orig |> pull(id))) |>
+  pull(inclination) |>
+  purrr::map(
+    \(.x) fff(inclination_deg = .x)
+    # |>
+    #   pivot_wider(names_from = lat, values_from = val)
+  ) |>
+  bind_cols() |>
+  rename(lll = `lat...1`) |>
+  select(-starts_with("lat")) |>
+  mutate(
+    vvv = rowSums(pick(starts_with("val"))),
+    nnn = length(pick(starts_with("val"))),
+    vvv = vvv / nnn
+  ) |>
+  select(-starts_with("val"), -nnn) |>
+  rename(lat = "lll", val = "vvv")
+
+uuu |>
+  ggplot() +
+  geom_line(aes(x = lat, y = val), linewidth = 1.3) +
+  scale_x_continuous("latitude", breaks = c(-90, -60, -30, 0, 30, 60, 90)) +
+  theme_minimal(base_size = 20) +
+  theme(
[…]

### 2025-04-11 R/weighting_function.R
bigger font for paper's figure

@@ -60,7 +60,7 @@ uuu |>
   ggplot() +
   geom_line(aes(x = lat, y = val), linewidth = 1.3) +
   scale_x_continuous("latitude", breaks = c(-90, -60, -30, 0, 30, 60, 90)) +
-  theme_minimal(base_size = 20) +
+  theme_minimal(base_size = 24) +
   theme(
     axis.title.y = element_blank(),
     panel.grid.minor.x = element_blank(),

### 2025-04-11 R/data-cube.R
new data cubes figure

@@ -98,7 +98,6 @@ eur_hex <- eur_hex * 6 + c(-8.8, 8.5)
 
 de = de * 6 + c(-8.5, 8.5)
 
-# pdf(here("media", "figures", "cubes.pdf"))
 png(
   filename = here("media", "figures", "cubes.png"),
   width = 19.5,

### 2025-04-12 paper.qmd
new cubes

@@
 
-The probability of impact for debris object $k$, $p_k$, is expressed as.
+We follow the same mathematical approach of [@patera2008].
+The probability, $p_k$, of a debris object $k$, impacting an aircraft over the
+area, $a$, is expressed as:
 
@@
 
-where $\rho({\vec{x}})$ is the probability density of an aircraft
-and $\sigma(\vec{x})$ is probability density of a debris object in the
-position $\vec{x}$.
+where $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
+$\sigma(\vec{x})$ is impact probability density of a debris object in the
+position $\vec{x}$ and $\varepsilon_k$ is the effective expose area.
+
+We can compute this probability for each hexagonal cell summing up for every
+group of same aircraft type (so with same $\varepsilon_k$).
 
@@
 Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
-circular orbits prior to reentry even if initially on an elliptical ones.
+circular orbits prior to reentry even if initially on elliptical ones.
 In fact drag at perigee will lower apogee and similarly drag at apogee will
@@
 
-So for example [@patera2008], if a building is located at latitude 30 degrees
+So for example if a building is located at latitude 30 degrees
 north (or south) and has an area of $\qty{1000}{m^2}$ and a piece of debris
@@
 
-$\num{0.046}$
+## Aircraft effective exposed area
 
-## Aircraft effective exposed area
-From Patera
+As from [@patera2008][^balistic_descent], we assume that reentring debris will
+descend vertically, which allows to express the effective expose area,
+$\varepsilon_i$, as
 
-> Even for a ballistic coefficient of 50 lbs/ft2, the flight path angle
-> of a reentering object at 40,000 feet is -86.6 degrees, which corresponds
-> to nearly vertical descent.
-> Since most debris objects have ballistic coefficients equal to or 
-> less than 50 lbs/ft2, the vertical descent assumption is a very
-> accurate approximation.
+$$
+\varepsilon_i = \frac{s_{a} a_{h} + s_{d_i} a_{v}}{s_{d_i}}
+$$
 
-The effective expose area, $\varepsilon_i$, is
-$$
-\varepsilon_i = \frac{s_{a} a_{h} + s_{d} a_{v}}{s_d}
-$$
-where $s_{a}$ is the average aircraft speed, $s_{d}$ is the average debris 
-object fall speed, $a_v$ area of the aircraft as viewed from above,
+where $s_{a}$ is the average aircraft speed, $s_{d_i}$ is the average fall speed
+of the $i$-th debris object, $a_v$ area of the aircraft as viewed from above,
 $a_h$ area of the aircraft as viewed from the from the front of the aircraft.
+
+In our calculations we will use
+
+* debris fall speed $s_{d} = \qty{145}{mph}$
+* aircraft speed, $s_{a}$, equal to its cruise speed
+* aircraft area as viewed from above, $a_v = W \times L$, where $W$ is the
+  wingspan, and $L$ the length
+* aircraft area as viewed from the front, $a_h = W \times H$, where $W$ is the
+  wingspan, and $H$ the height
+
+The values to use have been retrieved from [@performance_database;@skybrary;@doc8643].
+
+[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
+path angle of a reentering object at 40,000 feet is -86.6 degrees, which
+corresponds to nearly vertical descent.
+Since most debris objects have ballistic coefficients equal to or 
+less than 50 lbs/ft2, the vertical descent assumption is a very
+accurate approximation.*
+
+## Weighting function
+
+::: {#fig-weighting}
+
+```{r weighting}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/figures/weighting.png")
+```
+
+Weighting function derived from 25 years of reentries. 
+:::
 

### 2025-04-12 setup.qmd
new plot

@@
 
-This works takes NM CTFM profiles, resamples them and calculates number of
+This work takes NM CTFM profiles, resamples them and calculates the number of
 flight present per hour, `H`, of date, `D` in H3 hexagonal cells

### 2025-04-12 R/weighting_function.R
save weighting values

@@ -67,3 +67,5 @@ uuu |>
   )
 
 ggsave("media/figures/weighting.png")
+
+uuu |> write_csv(here::here("data", "weightings_2000-01-01_2025-01-01.csv"))

### 2025-04-13 R/prepare_density.R
prepare density

@@ -13,15 +13,36 @@ c(
   "data/trajectories_2024-12-05_resampled_30s_bbox_res_2.parquet",
   "data/trajectories_2024-08-01_resampled_30s_bbox_res_3.parquet",
   "data/trajectories_2024-12-05_resampled_30s_bbox_res_3.parquet",
-  NULL) |>
+  NULL
+) |>
   map(.f = read_parquet) |>
   bind_rows() |>
-  summarise(occupancy = n() * 30 / 3600,
-            .by = c(year, month, day, hour, h3_resolution, cell)) |>
+  summarise(
+    occupancy = n() * 30 / 3600,
+    .by = c(year, month, day, hour, h3_resolution, cell)
+  ) |>
   arrange(desc(occupancy)) |>
   write_parquet(here("data", "flight_density.parquet"), compression = "gzip")
 
-
+# group per aircraft type too
+c(
+  "data/trajectories_2024-08-01_resampled_30s_bbox_res_2.parquet",
+  "data/trajectories_2024-12-05_resampled_30s_bbox_res_2.parquet",
+  "data/trajectories_2024-08-01_resampled_30s_bbox_res_3.parquet",
+  "data/trajectories_2024-12-05_resampled_30s_bbox_res_3.parquet",
+  NULL
+) |>
+  map(.f = read_parquet) |>
+  bind_rows() |>
+  summarise(
+    occupancy = n() * 30 / 3600,
+    .by = c(year, month, day, hour, h3_resolution, cell, aircraft_type)
+  ) |>
+  arrange(occupancy) |>
+  write_parquet(
+    here("data", "flight_density_per_aircraft_type.parquet"),
+    compression = "gzip"
+  )
 
 # # Alternative (but far slower) computation of density
 # c(

### 2025-04-29 R/act_passengers.R
moved to aviodebris package

@@ -1,255 +1,0 @@
-# fmt: skip
-tribble(
-  ~icao, ~crew, ~pax_min, ~pax_typical, ~pax_max, ~accomodation,
-  "A124", 4,  88,  88,  88, "Flightcrew of 6 (later versions 4) and up to 88 passengers in upper deck and up to 150t payload.",
-  "A19N", 2, 120, 150, 160, "Flight crew of two and typical 120-150 passengers in two classes, max 160 in high density configuration.",
-  "A20N", 2, 150, 180, 194, "Flight crew of two and typical 150-180 passengers in two classes, max 194 in high density configuration",
-  "A21N", 2, 180, 220, 244, "Flight crew of two and typical 180-220 passengers in two classes, max 244 in high density configuration",
-  "A306", 2, 266, 266, 298, "Flightcrew of 2 and 266 passengers in typical seating configuration, up to 298 in single class configuration.",
-  "A310", 2, 220, 220, 247, "Flightcrew of 2 and 220 passengers in typical seating configuration, up to maximum of 247.",
-  "A318", 2, 107, 107, 117, "Flightcrew of two and typical 107 passengers in two classes, up to 117 in single class configuration.",
-  "A319", 2, 124, 124, 142, "Flightcrew of two and typical 124 passengers in two classes, up to 142 in high density configuration.",
-  "A320", 2, 150, 150, 180, "Flightcrew of two and typical 150 passengers in two classes, up to 180 in a one class layout.",
-  "A321", 2, 185, 185, 220, "Flightcrew of two and typical 185 passengers in two classes, up to 220 in all economy high density configuration.",
-  "A332", 2, 253, 253, 293, "Flightcrew of two and typical 253 passengers in three classes, up to 293 in two classes configuration.",
-  "A333", 2, 295, 295, 335, "Flightcrew of two and typical 295 passengers in three classes, up to 335 in high density configuration.",
-  "A338", 2, 220, 260, 406, "Flight crew of 2 and typical 220-260 passengers in three classes, max 406 in high density configuration.",
-  "A339", 2, 260, 300, 440, "Flight crew of 2 and typical 260-300 passengers in three classes, max 440 in high density configuration.",
[…]

### 2025-04-29 R/aircraft_types.R

@@ -1,32 +1,0 @@
-library(tidyverse)
-
-ac_types <- c(
-  "data/ectrl_acts.csv",
-  "data/sky_acts.csv",
-  "data/doc8643_acts.csv"
-) |>
-  purrr::map(.f = function(fn) {
-    # fmt: skip
-    cols <- cols(
-      icao              = col_character(),
-      name              = col_character(),
-      manufacturer      = col_character(),
-      type              = col_character(),
-      wtc               = col_character(),
-      recat_eu          = col_character(),
-      mtow_kg           = col_double(),
-      cruise_tas_kt     = col_double(),
-      cruise_mach       = col_double(),
-      cruise_range_nm   = col_double(),
-      cruise_ceiling_fl = col_double(),
-      wing_span_m       = col_double(),
-      length_m          = col_double(),
-      height_m          = col_double(),
-      pax               = col_double()
-    )
-    read_csv(fn, col_types = cols)
-  }) |>
-  bind_rows()
-
-ac_types |>
-  write_csv("data/aircraft_type_info.csv")

### 2025-04-29 R/scrape_aircraft_info.R
moved to aviodebris package

@@ -1,124 +1,0 @@
-# scrape aircraft dimensions
-library(conflicted)
-library(tidyverse)
-library(purrr)
-library(arrow)
-library(aviodebris)
-library(polite)
-
-# library(sf)
-# library(here)
-# library(rlang)
-
-conflict_prefer("filter", "dplyr", quiet = TRUE)
-
-# circa 15 secs
-trjs <- c(
-  "data/trajectories_2024-08-01_resampled_30s_bbox_res_2.parquet",
-  "data/trajectories_2024-12-05_resampled_30s_bbox_res_2.parquet",
-  "data/trajectories_2024-08-01_resampled_30s_bbox_res_3.parquet",
-  "data/trajectories_2024-12-05_resampled_30s_bbox_res_3.parquet",
-  NULL
-) |>
-  map(.f = read_parquet) |>
-  bind_rows()
-
-ac_types <- trjs |> pull(aircraft_type) |> unique() |> sort() |> setdiff("ZZZZ")
-
-
-# ---- EUROCONTROL Perfromance DB ----
-host <- "https://contentzone.eurocontrol.int/"
-session <- polite::bow(host, force = TRUE)
-scraper <- purrr::partial(scrape_aircraft_type_info_perfdb, session = session)
-
-# some aircraft types will be missing, i.e. helicopters
-all_acts <- ac_types |>
-  map(.f = scraper) |>
-  bind_rows()
-
-source("R/act_passengers.R")
-
-all_acts_missing <- all_acts |>
-  filter(if_all(c(-icao), is.na)) |>
-  pull(icao)
-
-all_acts <- all_acts |>
-  filter(!icao %in% all_acts_missing) |>
-  mutate(across(
-    c(
-      "mtow_kg",
-      "cruise_tas_kt",
-      "cruise_mach",
-      "cruise_range_nm",
-      "cruise_ceiling_fl"
-    ),
-    as.numeric
-  )) |>
-  left_join(act_all_pax) |>
-  mutate(pax = crew + pax_max) |>
-  select(-c("accomodation", "crew", "pax_min", "pax_typical", "pax_max")) |>
-  write_csv("data/ectrl_acts.csv")
-
-
-# ---- SkyBrary aircraft pages ----
-host <- "https://skybrary.aero/"
-session <- polite::bow(host, force = TRUE)
-scraper <- purrr::partial(scrape_aircraft_type_info_skybrary, session = session)
-
-
-# scrape only missing type from perf DB
[…]

### 2025-04-29 R/scrape_airport_passengers.R

@@ -1,59 +1,0 @@
-# To clean data
-library(tidyverse)
-library(lubridate)
-library(janitor)
-library(here)
-
-# To scrape data
-library(rvest)
-library(httr)
-library(polite)
-library(magrittr)
-
-
-url <- "https://en.wikipedia.org/wiki/List_of_the_busiest_airports_in_Europe#2024"
-url_bow <- polite::bow(url)
-
-
-eu_pass <- url_bow |>
-  polite::scrape() |>
-  rvest::html_nodes("table.wikitable") |>
-  rvest::html_table(header = TRUE, trim = TRUE, dec = ".") |>
-  simplify() |>
-  extract2(1) |>
-  clean_names() |>
-  slice(-1) |>
-  select(-2) |>
-  select(1:5) |>
-  mutate(
-    rank2024 = as.integer(rank2024),
-    passengers = str_remove(passengers, "\\[\\d*\\]$"),
-    passengers = str_remove_all(passengers, ","),
-    passengers = as.numeric(passengers),
-    NULL
-  ) |>
-  mutate(
-    year = 2024,
-    passengers = if_else(city_served == "Toulouse", 7844953, passengers),
-    passengers = if_else(
-      airport == "Brussels South Charleroi Airport",
-      10500000,
-      passengers
-    ),
-    # keep same number as 2023
-    passengers = if_else(
-      airport == "Bordeaux–Mérignac Airport",
-      6584194,
-      passengers
-    ),
-    passengers = if_else(
-      airport == "Beauvais–Tillé Airport",
-      5638955,
-      passengers
-    ),
-    NULL
-  ) |>
-  rename(rank = rank2024)
-
-eu_pass |>
-  write_csv(here("data", "airport_passengers.csv"))

### 2025-04-29 R/plot-cells-map.R
plot EUROONTROL area too

@@ -1,11 +1,56 @@
 # H3 hexagons covering NM area bbox
-
+library(tidyverse)
 library(aviodebris)
+library(pruatlas)
+library(ggplot2)
+library(sf)
+library(here)
+
+ms <- "BI|E.|L.|UD|UG|GM|UK|GC"
+fir_ctry <- country_fir(
+  firs = pruatlas::firs_nm_406,
+  icao_id = "BI|E.|L.|UD|UG|GM|UK|GC",
+  fl = 200,
+  merge = TRUE
+) |>
+  sf::st_transform(crs = sf::st_crs(3035))
+
+bbox <- fir_ctry |>
+  sf::st_convex_hull() |>
+  sf::st_buffer(1000) |>
+  sf::st_transform(crs = sf::st_crs(4326)) |>
+  sf::st_bbox()
+
 
 res <- 3
-g <- plot_hexes_map(res)
-ggsave(filename = here("figures", str_glue("cells_{res}.png")), plot = g)
+g <- plot_hexes_map(res) +
+  geom_sf(
+    data = fir_ctry,
+    fill = "blue",
+    colour = "blue",
+    linewidth = 0.6,
+    alpha = 0.1
+  ) +
+  ggplot2::coord_sf(crs = "ESRI:102013", datum = NA)
+
+ggsave(
+  filename = here("media", "figures", str_glue("cells_{res}.png")),
+  plot = g
+)
+
 
 res <- 2
-g <- plot_hexes_map(res)
-ggsave(filename = here("figures", str_glue("cells_{res}.png")), plot = g)
+g <- plot_hexes_map(res) +
+  geom_sf(
+    data = fir_ctry,
+    fill = "blue",
+    colour = "blue",
+    linewidth = 0.6,
+    alpha = 0.1
+  ) +
+  ggplot2::coord_sf(crs = "ESRI:102013", datum = NA)
+
+ggsave(
+  filename = here("media", "figures", str_glue("cells_{res}.png")),
+  plot = g
+)

### 2025-04-29 R/plot_weighting_function_inclination_i.R
plot U-shaped impact probability density  for debris of inclination `i`

@@ -1,0 +1,30 @@
+# plot U-shaped impact probability density for debris of inclination `i`
+
+library(tidyverse)
+library(aviodebris)
+
+# Earth equatorial radius
+rE <- 6378e3
+# altitude of satellite from (spherical) Earth surface, i.e. 500 km for LEO
+alt <- 550e3
+
+# inclination of Kosmos 482 Descent Craft (1972-023E, cat. nr. 6073)
+# see https://sattrackcam.blogspot.com/2025/04/kosmos-842-descent-craft-reentry.html
+inclination_deg = 51.7
+
+www <- latitude_weights(
+  delta_lat = 0.5,
+  altitude = rE + alt,
+  inclination_deg = inclination_deg
+)
+www |>
+  ggplot() +
+  geom_line(aes(x = lat, y = val), linewidth = 1.3) +
+  scale_x_continuous("latitude", breaks = c(-90, -60, -30, 0, 30, 60, 90)) +
+  theme_minimal(base_size = 24) +
+  theme(
+    axis.title.y = element_blank(),
+    panel.grid.minor.x = element_blank(),
+  )
+
+ggsave("media/figures/weighting_inclination_i.png")

### 2025-04-29 R/plot-weighting-latitude.R
refreshed figures

@@ -1,6 +1,7 @@
 library(tidyverse)
 library(aviodebris)
 
+
 weightings_half_degree_latitude |>
   ggplot() +
   geom_line(aes(x = lat, y = val), linewidth = 1.3) +
@@ -17,16 +18,36 @@ ggsave("media/figures/weighting.png")
 # plot density
 library(sf)
 library(h3jsr)
-bb <- bbox_nm()
-xlim <- c(bb["xmin"], bb["xmax"])
-ylim <- c(bb["ymin"], bb["ymax"])
+library(giscoR)
+library(scales)
+
+europe_hex <- function(res_h3, res_spatial = "20M") {
+  eur_hex_union <- aviodebris::bbox_nm() |>
+    aviodebris::bbox_of_hexes_for_bbox_at_res(resolution = res_h3)
+
+  gisco_get_countries(
+    resolution = "20",
+    epsg = "4326",
+    spatialtype = "RG",
+    region = c("Europe", "Africa", "Asia")
+  ) |>
+    mutate(res = res_spatial) |>
+    st_intersection(eur_hex_union) |>
+    select(res)
+}
+
 res <- 3L
 
 eur_hex <- bbox_nm() |> hexes_for_bbox_at_res(resolution = res)
 
-weightings_h3_resolution_3_hourly |>
+ddd <- weightings_h3_resolution_3_hourly |>
   filter(h3_resolution_3 %in% (eur_hex |> pull(h3_address))) |>
-  cell_to_polygon(simple = FALSE) |>
+  cell_to_polygon(simple = FALSE)
+
+MAX <- ddd |> pull("w") |> max()
+min <- ddd |> pull("w") |> min()
+
+ddd |>
   ggplot() +
   ggplot2::geom_sf(
     aes(fill = w),
@@ -36,4 +57,87 @@ weightings_h3_resolution_3_hourly |>
     title = "weightings on the spherical Earth",
     subtitle = "H3 hexes (res = 3)"
   )
+
+
+# draw hexagon for key shape in legend
+draw_key_hex <- function(data, params, size) {
+  # make fill inherit color if NA
+  if (is.na(data$fill)) data$fill <- data$col
+
+  # hexagon vertex coordinates
+  v <- list(
+    x = c(0.95, 0.725, 0.275, 0.05, 0.275, 0.725),
+    y = c(
+      0.5,
+      0.110288568297003,
+      0.110288568297003,
+      0.5,
+      0.889711431702997,
[…]

### 2025-04-30 setup.qmd
update setup

@@
 
-R and Python scripts are in the `R/` and `python/` directories respectively.
+R scripts are in the `R/` directory.
+
+You need duckdb 1.2.2 with `icu` extension and `h3` community extension
+installed (extension location hardcoded in `aviodebris::hexagonize`).
 
@@
 
-* H3 hexagons at resolution `R = 2`: 
-  - Average Area (km2): `86801.78`
-  - Average edge length (Km): `182.51`
 * H3 hexagons at resolution `R = 3`: 
@@
   - Average edge length (Km): `68.98`
-* bounding box approximating EUROCONTROL Area (decimal degrees):
-  - xmin = -25.488281
-  - ymin = 26.638253
-  - xmax = 45.407181
-  - ymax = 71.864780
-* bounding box containing the HEX at resolution 2 that match
-  the bounding box approximating EUROCONTROL Area:
-  - xmin = -27.78166
-  - xmax =  48.56037
-  - ymin =  24.90098
-  - ymax =  72.85065
+* bounding box approximating EUROCONTROL Area (decimal degrees) in the
+  `aviodebris` packages (enclosing FIRs of Member States)
 
@@
 #| echo: false
-knitr::include_graphics("media/figures/cells_2.png")
+knitr::include_graphics("media/figures/cells_3.png")
 ```
@@
 H3 **[cells]{colour="#ff0000"}** ([red]{colour="#ffffff" bg-colour="#ff0000"})
-at resolution 2 and
-their **[centers]{colour="#0000ff"}** ([blue]{colour="#ffffff" bg-colour="#0000ff"}) covering EUROCONTROL Area.
+at resolution 3 and
+their **[centers]{colour="#0000ff"}** ([blue]{colour="#ffffff" bg-colour="#0000ff"}) covering EUROCONTROL Area (([blue]{colour="#ffffff" bg-colour="#0000ff"}) polygon).
 
@@
 This is achieved via script `export_data.R`.
-The relevant trajectories are saved in
-`data-raw/trjs_{D}.parquet`, i.e. `data-raw/trjs_2024-08-01.parquet`.
+The relevant trajectories are saved in `data-raw/trjs/trjs_{D}.parquet`,
+i.e. `data-raw/trjs/trjs_2024-08-01.parquet`.
 
 ### Trajectory Resampling
-We load the trajectories and resample them at 30s using the Python `traffic` library.
-This is performed by the `resample_aug.py` and `resample_dec.py`
-scripts and is time consuming (40 min each).
-Execute as:
-
-> $ python resample_aug.py
-
-The result is in (52min 51s)
+We load the trajectories and resample them at 30s using helper function in the
+`aviodebris` package, 10min per day.  
+The result is in 
 `data/trajectories_2024-08-01_resampled_30s.parquet`
@@
 ### Fast H3 Labelling
-The script `prepare_pre_density.R` uses DuckDB for fast labelling
+The `hexagonize` function in `aviodebris` uses DuckDB (plus ICU extension
+and H3 community extension) for fast labelling
 of each lon/lat to the relevant H3 cell (at resolution `R`)
-and filtering for samples between bbox.
+and filtering for samples within EUROCONTROL bbox.
 Keep only point in the hex cells of the bbox and for day `D`.
-The result is saved in `trajectories_<YYYY-MM-DD>_resampled_30s_bbox_res_3.parquet` for date `D`.
+The result is saved in `trajectories_<YYYY-MM-DD>_resampled_30s_bbox_res_<R>.parquet` for date `D`.
 
@@
 
-The script `prepare_density.R` calculates the occupancy per hour `H` of `D` 
-for each cell at resolution `R`
-.
+The function `traffic_density_hourly` in the package `aviodebris` calculates
+the occupancy per hour `H` of `D` for each cell at resolution `R`.
+A file for each day is saved as
+`traffic_density_<YYYY-MM-DD>_res_<R>_hourly.parquet` for date `D`.
+The results are summarised per aircraft type.
+
+## Collision and Casuality Risk
+The function `collision_and_casuality_risk_expectation_hourly`
+calculates both collision and casuality risk on a hourly basis.

### 2025-04-30 paper.qmd
semi final?

--- 2025-04-06 paper.qmd
@@
 ---
-title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"
+title: "Preparing for Potential Closure of European Airspaces due to Reentering Space Objects"
 
@@
 # Abstract
-The risk of collision of re-entering space debris with aircraft is 
-increasing due to both increasing number of airborne flights and 
-re-entries. National and international authorities need to balance
-safety and economic impact of adopted measures.  
-We asses airspaces density as flown by commercial airlines
-distinguishing seasonal and hourly patterns.
-We base our hourly density estimation on (enhanced) flown trajectories as
-recorded by EUROCONTROL's \acr{NM} binned over Uber's H3
-gridding system. Impact risk probabilities will then calculated based on work
-from [@patera2008].
-This combined with a buffered area along the the path of re-entry,
-can provide risk impact probabilities for deciding on safety measure such as
-airspace closures and/or re-routings.
+The risk of collision between uncontrolled re-entering space debris and aircraft
+is  increasing due to both ever-growing number of airborne flights and 
+of orbiting bodies (and relevant re-entries). 
+National and international authorities need to balance safety and economic
+impact of adopted measures should a reentry event be announced, usually on short
+notice.  
+We employ hourly and daily spatial density of flight calculated from
+trajectories as recorded by EUROCONTROL's Network Manager and
+binned over Uber's H3 gridding system.  
+Impact risk probabilities are then calculated based on previous theoretical
+work but using these flights trajectories instead of proxies such as
+population density or ADS-B/Mode S message counts.
+These densities combined with a buffered area along the eventual path of
+re-entry, can provide risk impact probabilities to authorities as an aid
+for deciding on safety measures such as airspace closures and/or re-routings.
 
@@
 
-We base our hourly flight density estimation on trajectories flown
-by \acr{IFR} commercial aircraft as
-recorded by EUROCONTROL's \acr{NM}[^1] and overlapped on Uber's H3
-gridding system at different resolutions [@h3resolutions].
-@fig-cells shows cells for resolution $2$ which corresponds to hexagons
-of average area $\qty{86801.78}{km^2}$ and average edge length of 
-$\qty{182.51}{km}$, see also @tbl-resolutions for other resolutions.
+The ever-growing forecast number of flights, see @fig-forecast, and the modern
+race to commercialisation of space call for a reevaluation of the risk of
+collision and casuality due to uncontrolled debris reentries.
+Oversighting authorities need methodologies and tools to assess which 
+safety measures are the most adequate to minimize casualties and economic
+impact.
+
+Surging number of flights in Europe [@forecastUpdate2025] and in the
+world, but more importantly under development/deployment megaconstellations
+of satellites on \acr{LEO} such as already active Starlink and OneWeb or
+planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
+of orbiting objects with a short operational life some of which will 
+inevitably reenter Earth atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].
+
+This paper develops a practical and straightforward pipeline to rapidly
+assess the risk from uncontrolled reentry of space debris for the European
+region.
+Apart from the general risk assessment of potential reentries, we analyze
+a soon to happen reentry of debris from a 1972 failed USSR mission to Venus.
+
+::: {#fig-forecast}
+
+```{r forecast}
+#| out-width: "100%"
+#| echo: false
+#| fig-scap: "Forecast"
+knitr::include_graphics("media/figures/forecast_statfor.png")
+```
+
+EUROCONTROL 7-year forecast 2025-2031 (Spring 2025).
+:::
 
 ```{r}
-#| label: tbl-resolutions
-#| tbl-cap: H3 hexagons, average values
+#| label: tbl-megaconstellations
+#| tbl-cap: Megaconstellations (planned > 1000) as per filing to the \acr{ITU}
 #| tbl-cap-location: bottom
 
-library(tidyverse)
+library(tibble)
 library(gt)
-library(gtExtras)
 
-tribble(
-  ~resolution, ~hex_area, ~hex_edge,
-  2L , 86801.780398997, 182.5129565,
-  3L,  12393.434655088,  68.97922179,
-  4L,   1770.347654491,  26.07175968
-) |> 
+megas <- tibble::tribble(
+  ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
+  "Starlink", "US", 4714, 3521, 4714, 2018,
+  "Starlink2A", "US", 3689, 3070, 6720, 2022,
+  "Starlink2", "US", 0, 0, 30456, NA,
+  "OneWeb", "UK", 660, 635, 716, 2019,
+  "OneWeb2", "UK", 0, 0, 2304, NA,
+  "Kuiper", "US", 29, 27, 3232, 2023,
+  "StarShield", "US", 193, 126, 32, 2022,
+  "Xingwang", "CN", 50, 10, 996, 2021,
+  "Qianfan", "CN", 90, 28, 32, 2024,
+  "Guangwang", "CN", 0, 0, 12992, NA,
+  "Yinhe", "CN", 8, 7, 1000, 2020,
+  "Hanwha", "KR", 0, 0, 2000, NA,
+  "Lynk", "US", 10, 6, 2000, 2020,
+  "Astra", "US", 0, 0, 13620, NA,
+  "Telesat", "CA", 0, 0, 300, NA,
+  "HVNET", "US", 0, 0, 1440, NA,
+  "SpinLaunch", "US", 0, 0, 1190, NA,
+  "Globalstar3", "DE", 0, 0, 3080, NA,
+  "Honghu-3", "CN", 0, 0, 10000, NA,
+  "Semaphore", "FR", 0, 0, 116640, NA,
+  "E-Space", "US", 4, 0, 337323, 2022
+)
+
+# ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
+megas |> 
   gt() |> 
    cols_label(
-    resolution ~ "Resolution",
-    hex_area ~ md("Cell Area"),
-    hex_edge ~ md("Edge length")
+    constellation ~ "Constellation",
+    launched ~ "Launched",
+    active ~ "Active",
+    planned ~ "Planned",
+    first_launch ~ "First launch"
   ) |> 
-    gt::tab_header(
-    title = md("H3 Cell Statistics"),
-    subtitle = md("average hexagons")
+   sub_missing(
+    columns = "first_launch",
+    missing_text = "-"
   ) |> 
-  cols_units(
-    hex_area = "km^2",
-    hex_edge = "km"
-  ) |>
-  opt_align_table_header(align = "left")
+  opt_table_font(size = "10px")
+
 ```
+
+# Method
+
+We follow the same mathematical approach of 
+[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
+approximating flight density with population density [@patera2008] or
+aircraft transponder data [@wrightAirspaceClosuresDue2025], we use flown 
+trajectories.
+
+We base our hourly flight density estimation on flown \acr{IFR} trajectories as
+recorded by EUROCONTROL's \acr{NM}[^1] and binned on Uber's H3 gridding system
+[@uber-h3] for fast processing.  
+This work will use H3 cells at resolution 3 which corresponds to hexagons of
+average area $\qty{12393}{km^2}$ and average edge length of
+$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
 
@@
 ```{r cells-map}
-#| out-width: "100%"
+#| out-width: "120%"
 #| echo: false
-#| fig-scap: "H3 cells at resolution 2"
-knitr::include_graphics("media/figures/cells_2.png")
+#| fig-scap: "H3 cells at resolution 3"
+knitr::include_graphics("media/figures/cells_3.png")
 ```
 
-H3 cells (red) at resolution 2 and their centers (blue) covering EUROCONTROL Area.
+H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
+box of 43 EUROCONTROL Member States FIR's (blue shaded polygon).
+
 :::
 
-We characterise densities on an hourly basis for representative days
-in summer and winter.[^2]
+We characterise densities on an hourly basis for the days of the 2023 and 2024
+busiest \acr{IATA} summer ISO week.
 
-We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008; @ailorLargeConstellationDisposal2020] and
+We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
 provide a reproducible example for States, Agencies or relevant authorities
@@
 
-[^2]: We will consider representative days in \acr{IATA} summer 2024 
-and winter 2025 season.
+Next sections will develop the mathematcal background for the evaluation of
+collision and casuality risk of which traffic density is a key component.
 
-@fig-density-winter-res-3 provides a promising result in the hourly flight
-density calculation which at least shows
-expected outcomes such as cells at zero density for war or no-fly zones such as
-Ukraine, Libia or Syria.
+@fig-density-summer2024-res-3 shows hourly flight density for 2024-07-05.
+Expected outcomes such as cells at zero density for war or no-fly zones
+(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
+and East-bound over Türkiye) are clearly highlighted.
 
-::: {#fig-density-winter-res-3}
+::: {#fig-density-summer2024-res-3}
 
@@
 #| echo: false
-knitr::include_graphics("media/figures/flight_density_2024-08-01T12_3.png")
+knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
 ```
 
-Flight density in the EUROCONTROL Area on 2024-08-01T12 (H3 cells at resolution $3$).
+Flight density in the EUROCONTROL Area on 2024-07-05, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
 :::
 
-# Method
-
-The probability of impact for debris object $k$, $p_k$, is expressed as.
+## Impact Probability for Single Flight, Single Debris
+The probability, $p_k$, of a debris (point) object $k$, impacting an
+aircraft (point) object over the area, $a$, is expressed as:
 
 $$
-p_k = \varepsilon_k \int_a \rho(\vec{x}) \sigma(\vec{x}) da
-$$ {#eq-impact-probability}
+p_k = \varepsilon_k \int_a \rho(\vec{x}) \sigma_k(\vec{x}) da
+$$
 
-where $\rho({\vec{x}})$ is the probability density of an aircraft
-and $\sigma(\vec{x})$ is probability density of a debris object in the
-position $\vec{x}$.
+where $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
+$\sigma(\vec{x})$ is the impact probability density of a debris object $k$
+in the position $\vec{x}$ and $\varepsilon_k$ is the effective exposed area,
+i.e. the effective risk area for the $k^{th}$ debris object impacting
+the aircraft (see @sec-eea later on).
 
-## Impact probability density
+If we discretaze the area of interest $a$ using H3 cells,
+we have
+
+$$
+p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h)
+$$
+
+So for cell $h$ we have
+
+$$
+p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h)
+$$
+
+## Aircraft effective exposed area{#sec-eea}
+
+We assume the aircraft impact area to be much larger than the debris
+object area so as to neglet the debris impact area.
+We can also simplify the geometry of the aircraft and model it as
+a rectangular solid with dimesions of wingspan, length and height. 
+Now, if the aircraft were stationary, the area exposed to a debris impact would
+be the top view, while if the debris were stationary it would be the front view.
+In reality the aircraft is moving (manly) horizontally and the debris 
+vertically [^balistic_descent], which which allows to express the effective
+exposed area, $\varepsilon_k$, as
+
+$$
+\varepsilon_k = \frac{s_{a} a_{h} + s_{d_k} a_{v}}{s_{d_k}}
+$$
+
+where $s_{a}$ is the average aircraft speed, $s_{d_k}$ is the average fall speed
+of the $k$-th debris object, $a_T$ area of the aircraft as viewed from the top,
+$a_F$ area of the aircraft as viewed from the from the front of the aircraft.
+
+In our calculations we will use ($t$ subscript for aircraft type)
+
+* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$
+* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
+* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
+  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
+  type $t$
+* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
+  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
+  type $t$
+
+The effective exposed area now only depends on the aircraft type $t$ (dimensions
+and cruise speed) and can be rewritten as:
+
+$$
+\varepsilon_t = \frac{C_{t} F_{t} + S A_{t}}{S}
+$$
+
+The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].
+
+The impact  probability for cell $h$ and single aircraft of type $t$ becomes:
+
+$$
+p_{t,k}(h) = \varepsilon_t \rho_t(h) \sigma_k(h)
+$$
+
+[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
+path angle of a reentering object at 40,000 feet is -86.6 degrees, which
+corresponds to nearly vertical descent.
+Since most debris objects have ballistic coefficients equal to or 
+less than 50 lbs/ft2, the vertical descent assumption is a very
+accurate approximation.* [@patera2008]
+
+## Impact probability density for debris of inclination $i$
 
 Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
-circular orbits prior to reentry even if initially on an elliptical ones.
-In fact drag at perigee will lower apogee and similarly drag at apogee will
-lower perigee. But because drag at perigee is higher due to denser atmosphere,
-the apogee  will decrease faster than perigee and the orbit will then tend to be
-a circular one.
-The angular position of the reentry point from the ascending node, 
-see @sec-orbitals, is uniformly distributed, $\sigma(\theta) = 1 / (2\pi)$,
-due to uncertainties in atmospheric drag and the gravitational effects of
-the Equatorial bulge, hence we assume that the impact longitude, $\lambda$,
-is uniformely distributed $\sigma(\lambda) = 1/(2\pi)$.
+circular orbits prior to reentry even if initially on elliptical
+ones.[^circular-orbit]
+
+The angular position, $\theta$, of the reentry point from the ascending node, 
+see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
+$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
+and the gravitational effects of the Equatorial bulge.
+Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.
 
@@
 
-So for example [@patera2008], if a building is located at latitude 30 degrees
-north (or south) and has an area of $\qty{1000}{m^2}$ and a piece of debris
-reenters the Earth’s atmosphere from an orbit having an inclination of 45 degrees,
-the impact probability is given by
+where $R$ is the distance from the center of the Earth of the object potentially
+being impacted by the debris.
+
+So for example if a building is located at latitude 30 degrees and has an area 
+of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth’s atmosphere
+from an orbit having an inclination of 45 degrees, the impact probability is 
+given by
 
@@
 
-$\num{0.046}$
+The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
+$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic
+U shape.
 
-## Aircraft effective exposed area
-From Patera
+::: {#fig-weighting-inclination-i}
 
-> Even for a ballistic coefficient of 50 lbs/ft2, the flight path angle
-> of a reentering object at 40,000 feet is -86.6 degrees, which corresponds
-> to nearly vertical descent.
-> Since most debris objects have ballistic coefficients equal to or 
-> less than 50 lbs/ft2, the vertical descent assumption is a very
-> accurate approximation.
+```{r weighting-inclination-i}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/figures/weighting_inclination_i.png")
+```
 
-The effective expose area, $\varepsilon_i$, is
+Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of inclination $i$ and altitude $\qty{550}{km}$. 
+:::
+
+[^circular-orbit]: In fact drag at perigee will lower apogee and similarly drag
+at apogee will lower perigee. But because drag at perigee is higher due to
+denser atmosphere, the apogee  will decrease faster than perigee and the orbit
+will then tend to be a circular one.
+
+## Impact probability density for debris
+
+If we consider historical debris reentry data, we can numerically derive
+the debris impact probability as a weighting function by summing up and averaging
+all single densities:
+
+$$ \sigma(\varphi) = \sum_k \sigma(\varphi, i_k)$$
+
+From [@gcat] we extracted 1968 debris reentries during 25 calendar years from 
+1st Jan 2000 till 31st Dec 2024.
+The result of combining all the U-shaped densities can be seen in @fig-weighting.
+
+::: {#fig-weighting}
+
+```{r weighting}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/figures/weighting.png")
+```
+
+Weighting function derived from 25 years of reentries as an
+empirical approximation of $\sigma(\varphi)$. 
+:::
+
+Spatially combining $\sigma(\lambda,\varphi)$ for the H3 cells in the
+\acr{NM} area we obtain the weightings per cell (per hour) as in
+@fig-weighting-nm-area.
+
+::: {#fig-weighting-nm-area}
+
+```{r weighting-nm-area}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/figures/weighting_on_Earth.png")
+```
+
+Weightings for H3 cells in \acr{NM} area. 
+:::
+
+## Hourly impact probability per H3 cell
+By combining the hourly aircraft density (per aircraft type) with the relevant
+weightings per H3 cell, we can numerically compute the hourly impact probability
+for each hexagonal cell as:
+
 $$
-\varepsilon_i = \frac{s_{a} a_{h} + s_{d} a_{v}}{s_d}
+ E_h =  \sum_t W(h) \cdot \rho_{t}(h) \cdot \varepsilon_t
 $$
-where $s_{a}$ is the average aircraft speed, $s_{d}$ is the average debris 
-object fall speed, $a_v$ area of the aircraft as viewed from above,
-$a_h$ area of the aircraft as viewed from the from the front of the aircraft.
 
-# Data
+where $E_h$ is the collision expectation in the hexagon $h$,
+$W(h)$ is the median value of the weighting function over the area of hexagon $h$, $t$ is the \acr{ICAO} aircraft type,
+$\rho_{t}(h)$ is the traffic density for type $t$ in cell $h$ and
+$\varepsilon_T$ is effective exposed area for aircraft of type $t$.
+
+::: {#fig-collision}
+
+```{r collision}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/figures/cubes.png")
+```
+
+Collision risk probability. 
+:::
+
+## Use Case: KOSMOS 482 Descent Craft reentry
+
+In few days (at the time of writing this paper) around May 10, 2025, the
+KOSMOS 482 Descent Craft should reenter the Earth atmosphere [@kosmos482reentry].
+This is supposedly the lander module from a 1972 failed Soviet Venera mission
+[@SpaceReviewKosmos] of \qty{480}{kg} weight and circa \qty{1}{m} diameter.
+Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \qty{467}{degreesCelsius} (\qty{872}{degreesFarenheit})).
+So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
+
+# Data and Code for Reproducibility
+
+We have calculated collision and casuality risk probabilities for the two
+busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
+as logically depicted in @fig-data-densities is available in the companion
+GitHub repository of the paper at <https://github.com/espinielli/flight_density_space_debris>.
+We have also developed a companion R package `aviodebris` to heavylift the
+various data processing steps to produce the results in the paper;
+it is available at <https://github.com/espinielli/aviodebris>.
 

### 2025-04-30 R/plot-density-map.R

@@ -5,35 +5,27 @@ library(arrow)
 library(sf)
 library(h3jsr)
 library(ggplot2)
-library(h3jsr)
 library(magrittr)
 library(giscoR)
 library(grid)
+library(here)
 
 conflicts_prefer(dplyr::filter)
 
-
-density <- arrow::read_parquet("data/flight_density.parquet")
+date <- "2024-07-05" |> as_date()
+density <- arrow::read_parquet(
+  str_glue("data/traffic_density_{date}_res_3_hourly.parquet")
+)
 
 res <- 3
-YYYY <- 2024
-MM <- 8
-DD <- 1
-HH <- 10
 
 eur_hex <- bbox_nm() |>
   hexes_for_bbox_at_res(resolution = res)
 
 
 ddd <- density |>
-  filter(
-    year == YYYY,
-    month == MM,
-    day == DD,
-    hour == HH,
-    h3_resolution == res
-  ) |>
-  mutate(occupancy = mean(occupancy), .by = cell) |>
+  group_by(cell) |>
+  summarise(occupancy = sum(occupancy) / 24) |>
   filter(cell %in% (eur_hex |> pull(h3_address))) |>
   select(cell, everything()) |>
   cell_to_polygon(simple = FALSE)
@@ -155,13 +147,6 @@ ggplot() +
   theme_minimal() +
   coord_sf(crs = "ESRI:102013", datum = NA) +
   labs(
-    title = "Hourly Flight Density",
-    subtitle = str_glue(
-      "EUROCONTROL Area, {YYYY}-{MM}-{DD}, Hour = {HH} UTC",
-      MM = str_pad(MM, width = 2, side = "left", pad = "0"),
-      DD = str_pad(DD, width = 2, side = "left", pad = "0"),
-      HH = str_pad(HH, width = 2, side = "left", pad = "0")
-    ),
     caption = "Data: Network Manager, EUROCONTROL",
     NULL
   ) +
@@ -184,10 +169,14 @@ ggplot() +
       margin = margin(t = -5, r = 10, b = 0, l = 0, unit = "mm")
     ),
     legend.position = c(0.18, 0.8),
-    legend.title = element_text(size = 15),
-    legend.text = element_text(size = 13),
+    legend.title = element_text(size = 24),
+    legend.text = element_text(size = 24),
     NULL
   ) +
   guides(fill = guide_legend(override.aes = list(size = 3.0)))
 
[…]

### 2025-04-30 paper.qmd
sort of final structure draft

@@
 #| echo: false
-knitr::include_graphics("media/figures/cubes.png")
+knitr::include_graphics("media/example-image.png")
 ```
@@
 
-In few days (at the time of writing this paper) around May 10, 2025, the
-KOSMOS 482 Descent Craft should reenter the Earth atmosphere [@kosmos482reentry].
+In few days (at the time of writing this paper) at 6h May 10, 2025
+($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
+Earth's atmosphere [@kosmos482reentry].
 This is supposedly the lander module from a 1972 failed Soviet Venera mission
@@
 So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
+
+With what developed in this paper, we can derive the casuality risk and
+determine the maximum value using data for the busiest day in 
+the busiest 2024 week.
+
+Once the reentry track is known more or less accurately, the relevant
+H3 cells can be
+
+::: {#fig-kosmos}
+
+```{r kosmos}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/example-image.png")
+```
+
+Casuality for KOSMOS 482 Descent Craft reentry. 
+:::
 
@@
 
-# Conclusion / Outcome
+# Conclusion and Future work
 
@@
 
+Future work will explore the impact of megacostellations and the risk for
+European airspace.
+
 # Appendix {.appendix}
+# Orbital Elements {#sec-orbitals }
 
-## Orbital Elements {#sec-orbitals .appendix}
-We refer to @fig-orbital-elements to describe the orbital parameters of
-interest for the current work.
+:::{#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="t"}
 
-:::{#fig-orbital-elements layout="[40,-3,57]" fig-env="figure*"}
+```{r ellipse-parameters}
+#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$.
+(the semiparameter (a.k.a.
+_semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"
 
-![Keplerian elements: semi-major axis, $a$, the eccentricity, $e = \frac{2c}{2a}$.
-(the semiparameter (a.k.a.
-_semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as
-$p = b^2 / a = a (1 - e^2)$)](media/figures/classical_orbital_elements_2.png)
+#| echo: false
+knitr::include_graphics("media/figures/classical_orbital_elements_2.png")
+```
 
-![Keplerian elements: inclination, $i$, the right ascension of the ascending node, $\Omega$ , and the argument of periapsis, $\omega$.(Earth rotates around $\hat{K}$)](media/figures/classical_orbital_elements_4.png)
+```{r keplerian-elements}
+#| fig-cap: "Keplerian elements: inclination, $i$, the right ascension of the ascending node, $\\Omega$ , and the argument of periapsis, $\\omega$.(Earth rotates around $\\hat{K}$)"
+#| echo: false
+
+knitr::include_graphics("media/figures/classical_orbital_elements_4.png")
+```
 
@@
 
+We refer to @fig-orbital-elements to describe the orbital parameters of
+interest for the current work.
+
 The angular position, $\theta$, of the reentry point from the ascending node
@@
 
+These are used to derive numerically compute the weighting function,
+@fig-weighting-inclination-i:
+for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10001$,
+positions and calculate the corresponding cartesian position and the
+equivalent longitude and latitude.
+
 # References {.unnumbered}

### 2025-04-30 pipeline.R

@@ -1,0 +1,54 @@
+# pipeline fot the paper
+
+library(tidyverse)
+library(aviodebris)
+
+busy_2023 <- seq(as.Date("2023-07-03"), by = "day", length.out = 7)
+busy_2024 <- seq(as.Date("2024-07-01"), by = "day", length.out = 7)
+days <- c(
+  busy_2023,
+  as.Date("2023-09-01"),
+  busy_2024
+)
+
+# cherry peek days
+# days <- c(
+#   "2023-07-07",
+#   "2023-09-01",
+#   "2024-07-05"
+# )
+
+# exporting data 3min ca per day
+
+# 1. resample trajectories at 30s
+resample_30s <- purrr::partial(resample_traffic, interval = 30L)
+
+# fmt: skip
+{
+  # tictoc::tic()
+  days |> purrr::walk(.f = resample_30s)
+  # tictoc::toc()
+}
+
+# 2. assign H3 hex cells at resolution 3
+hexagonize_3_30s <- purrr::partial(
+  hexagonize_traffic,
+  resolution = 3L,
+  interval = 30L
+)
+days |> purrr::walk(hexagonize_3_30s)
+
+# 3. compute traffic density
+traffic_density_3_30s <- purrr::partial(
+  traffic_density_hourly,
+  resolution = 3L,
+  interval = 30L
+)
+days |> purrr::walk(traffic_density_3_30s)
+
+# 4. compute casuality
+casuality_3_30s_hourly <- purrr::partial(
+  collision_and_casuality_risk_expectation_hourly,
+  resolution = 3L
+)
+days |> purrr::walk(casuality_3_30s_hourly)

### 2025-04-30 paper.qmd
checked Quinten feedback

@@
 
-If we discretaze the area of interest $a$ using H3 cells,
+If we discretize the area of interest $a$ using H3 cells,
 we have
@@
 We can also simplify the geometry of the aircraft and model it as
-a rectangular solid with dimesions of wingspan, length and height. 
+a rectangular solid with dimesions of wingspan ($W$), length ($L$) and
+height ($H$). 
 Now, if the aircraft were stationary, the area exposed to a debris impact would
@@
 
-[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
-path angle of a reentering object at 40,000 feet is -86.6 degrees, which
-corresponds to nearly vertical descent.
+[^balistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
+the flight path angle of a reentering object at \qty{40000}{ft} is 
+\ang{-86.6}, which corresponds to nearly vertical descent.
 Since most debris objects have ballistic coefficients equal to or 
-less than 50 lbs/ft2, the vertical descent assumption is a very
+less than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
 accurate approximation.* [@patera2008]
@@
 
-If we consider historical debris reentry data, we can numerically derive
-the debris impact probability as a weighting function by summing up and averaging
-all single densities:
+If we consider historical uncontrolled debris reentry data, we can numerically
+derive the debris impact probability as a weighting function by summing up and
+averaging all single densities:
 
@@
 
-Future work will explore the impact of megacostellations and the risk for
-European airspace.
+Future work will explore the impact of megacostellations and flight growth
+forecast and what would happen in 20, 40, ..., 100 years in varios scenarios
+of uncontrolled  reentries and the risk for European airspace.
 
-# Appendix {.appendix}
-# Orbital Elements {#sec-orbitals }
+# Appendix: Orbital Elements {#sec-orbitals }
 

### 2025-05-01 paper.qmd
Pass through abstract + I Introduction

--- 2025-04-30 paper.qmd
@@
 ---
-title: "Preparing for Potential Closure of European Airspaces due to Reentering Space Objects"
+title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"
 
@@
       - ref: aiu
+    orcid: 0000-0003-2810-1404
+    email: rainer.koelle@eurocontrol.int
   - name: Quinten Goens
@@
       - ref: aiu
+    orcid: 0000-0003-2187-1924
+    email: quinten.goens@eurocontrol.int
 affiliations:
@@
 National and international authorities need to balance safety and economic
-impact of adopted measures should a reentry event be announced, usually on short
+impact of adopted measures should a re-entry event be announced, usually on short
 notice.  
-We employ hourly and daily spatial density of flight calculated from
+We employ hourly and daily spatial flight densities calculated from
 trajectories as recorded by EUROCONTROL's Network Manager and
-binned over Uber's H3 gridding system.  
+binned with the Uber H3 geospatial indexing system.  
 Impact risk probabilities are then calculated based on previous theoretical
-work but using these flights trajectories instead of proxies such as
-population density or ADS-B/Mode S message counts.
+work.
+This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
 These densities combined with a buffered area along the eventual path of
-re-entry, can provide risk impact probabilities to authorities as an aid
+re-entry, can provide mid-air impact probabilities to authorities as an aid
 for deciding on safety measures such as airspace closures and/or re-routings.
@@
 
-The ever-growing forecast number of flights, see @fig-forecast, and the modern
-race to commercialisation of space call for a reevaluation of the risk of
-collision and casuality due to uncontrolled debris reentries.
-Oversighting authorities need methodologies and tools to assess which 
+With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
+race to commercialise space call for a re-evaluation of the risk of
+collision and casualties due to uncontrolled debris re-entries.
+Overseeing authorities need methodologies and tools to assess which 
 safety measures are the most adequate to minimize casualties and economic
@@
 Surging number of flights in Europe [@forecastUpdate2025] and in the
-world, but more importantly under development/deployment megaconstellations
-of satellites on \acr{LEO} such as already active Starlink and OneWeb or
+world, but more importantly under development/deployment mega-constellations
+of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
 planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
-of orbiting objects with a short operational life some of which will 
-inevitably reenter Earth atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].
+of orbiting objects with a short operational life. Some of these will 
+inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].
 
 This paper develops a practical and straightforward pipeline to rapidly
-assess the risk from uncontrolled reentry of space debris for the European
+assess the risk from uncontrolled re-entry of space debris for the European
 region.
-Apart from the general risk assessment of potential reentries, we analyze
-a soon to happen reentry of debris from a 1972 failed USSR mission to Venus.
+Apart from the general risk assessment of potential re-entries, we analyse
+a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus.
 
@@
 Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
-circular orbits prior to reentry even if initially on elliptical
+circular orbits prior to re-entry even if initially on elliptical
 ones.[^circular-orbit]
 
-The angular position, $\theta$, of the reentry point from the ascending node, 
+The angular position, $\theta$, of the re-entry point from the ascending node, 
 see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
@@
 
-If we consider historical debris reentry data, we can numerically derive
+If we consider historical debris re-entry data, we can numerically derive
 the debris impact probability as a weighting function by summing up and averaging
@@
 
-From [@gcat] we extracted 1968 debris reentries during 25 calendar years from 
+From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
 1st Jan 2000 till 31st Dec 2024.
@@
 
-Weighting function derived from 25 years of reentries as an
+Weighting function derived from 25 years of re-entries as an
 empirical approximation of $\sigma(\varphi)$. 
@@
 
-## Use Case: KOSMOS 482 Descent Craft reentry
+## Use Case: KOSMOS 482 Descent Craft re-entry
 
@@
 ($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
-Earth's atmosphere [@kosmos482reentry].
+Earth's atmosphere [@kosmos482re-entry].
 This is supposedly the lander module from a 1972 failed Soviet Venera mission
@@
 
-Once the reentry track is known more or less accurately, the relevant
+Once the re-entry track is known more or less accurately, the relevant
 H3 cells can be
@@
 
-Casuality for KOSMOS 482 Descent Craft reentry. 
+Casuality for KOSMOS 482 Descent Craft re-entry. 
 :::
@@
 
-We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled reentry of space objects that could pose a safety concern on civilian aviation.
+We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
 The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.
@@
 
-The angular position, $\theta$, of the reentry point from the ascending node
+The angular position, $\theta$, of the re-entry point from the ascending node
 in the orbit plane is $\theta = \omega + \nu$.

### 2025-05-01 paper.qmd
Reviewed until C. Impact probability density for debris of inclination i

@@
 With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
-race to commercialise space call for a re-evaluation of the risk of
-collision and casualties due to uncontrolled debris re-entries.
+race to commercialise space it is clear a re-evaluation of the risks of
+collision and casualties due to uncontrolled debris re-entries is needed.
 Overseeing authorities need methodologies and tools to assess which 
-safety measures are the most adequate to minimize casualties and economic
+safety measures are the most adequate to minimise potential casualties and economic
 impact.
@@
 region.
-Apart from the general risk assessment of potential re-entries, we analyse
-a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus.
+Apart from the general risk assessment of potential re-entries, 
+a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus is analysed.
 
@@
     constellation ~ "Constellation",
+    state ~ "Country",
     launched ~ "Launched",
@@
 approximating flight density with population density [@patera2008] or
-aircraft transponder data [@wrightAirspaceClosuresDue2025], we use flown 
-trajectories.
+aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
+flight trajectories.
 
-We base our hourly flight density estimation on flown \acr{IFR} trajectories as
-recorded by EUROCONTROL's \acr{NM}[^1] and binned on Uber's H3 gridding system
-[@uber-h3] for fast processing.  
-This work will use H3 cells at resolution 3 which corresponds to hexagons of
-average area $\qty{12393}{km^2}$ and average edge length of
+The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
+recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
+[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
+This work will use H3 cells at resolution 3 which corresponds to hexagons of an
+average area $\qty{12393}{km^2}$ and an average edge length of
 $\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
@@
 H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
-box of 43 EUROCONTROL Member States FIR's (blue shaded polygon).
+box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).
 
@@
 
-We characterise densities on an hourly basis for the days of the 2023 and 2024
-busiest \acr{IATA} summer ISO week.
+The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.
 
-We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
-provide a reproducible example for States, Agencies or relevant authorities
+This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
+provides a reproducible example for States, Agencies or relevant authorities
 to build upon.
@@
 
-Next sections will develop the mathematcal background for the evaluation of
-collision and casuality risk of which traffic density is a key component.
+Next sections will develop the mathematical background for the evaluation of
+collision and casualties risk of which traffic density is a key component.
 
-@fig-density-summer2024-res-3 shows hourly flight density for 2024-07-05.
+@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
 Expected outcomes such as cells at zero density for war or no-fly zones
@@
 
-Flight density in the EUROCONTROL Area on 2024-07-05, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
+Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
 :::
 
-## Impact Probability for Single Flight, Single Debris
+## Impact probability for a single flight and single debris
 The probability, $p_k$, of a debris (point) object $k$, impacting an
@@
 
-If we discretaze the area of interest $a$ using H3 cells,
-we have
+Using H3 to discretise the area of interest $a$ using H3 cells gives
 
 $$
-p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h)
+p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h).
 $$
 
-So for cell $h$ we have
+Where for each cell $h$ the following applies
 
 $$
-p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h)
+p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h).
 $$
@@
 
-We assume the aircraft impact area to be much larger than the debris
-object area so as to neglet the debris impact area.
-We can also simplify the geometry of the aircraft and model it as
-a rectangular solid with dimesions of wingspan, length and height. 
+It is assumed the aircraft impact area to be much larger than the debris
+object area so as to neglect the debris impact area.
+The geometry of the aircraft is simplified and modelled as
+a rectangular cuboid with dimensions of wingspan, length and height. 
 Now, if the aircraft were stationary, the area exposed to a debris impact would
 be the top view, while if the debris were stationary it would be the front view.
-In reality the aircraft is moving (manly) horizontally and the debris 
+In reality an aircraft en-route is moving (mainly) horizontally and the debris 
 vertically [^balistic_descent], which which allows to express the effective
@@
 
-In our calculations we will use ($t$ subscript for aircraft type)
+In the calculations the following is assumed to be true ($t$ subscript for aircraft type)
 
@@
 [^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
-path angle of a reentering object at 40,000 feet is -86.6 degrees, which
+path angle of a re-entering object at 40,000 feet is -86.6 degrees, which
 corresponds to nearly vertical descent.

### 2025-05-02 paper.qmd
Update introduction

@@
 # Abstract
-The risk of collision between uncontrolled re-entering space debris and aircraft
+<!-- teaser --> The risk of collision between uncontrolled re-entering space debris and aircraft
 is  increasing due to both ever-growing number of airborne flights and 
 of orbiting bodies (and relevant re-entries). 
-National and international authorities need to balance safety and economic
+<!-- why important --> National and international authorities need to balance safety and economic
 impact of adopted measures should a re-entry event be announced, usually on short
-notice.  
-We employ hourly and daily spatial flight densities calculated from
+notice. <!-- no hard line breaks in abstracts -->
+<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
 trajectories as recorded by EUROCONTROL's Network Manager and
@@
 work. 
-This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
+<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
 These densities combined with a buffered area along the eventual path of
@@
 
+<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
+The quest for space is a driver for technological development.
+Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
+These developments helped to identify options for the wider use of space, benefits, and LALALALA.
+This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
+The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.
+
+<!-- zooming in on problem -->
 With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
-race to commercialise space it is clear a re-evaluation of the risks of
-collision and casualties due to uncontrolled debris re-entries is needed.
+race to commercialise space, there is a need to re-evaluate the risks of
+collision and casualties due to uncontrolled debris re-entries.
 Overseeing authorities need methodologies and tools to assess which 
@@
 
-This paper develops a practical and straightforward pipeline to rapidly
+<!-- now we have the risk / topic established -- next we need something to speak about a "high-level goal" or "what other work looked into to place this paper as a 'further development' -->
+
+...
+
+<!-- contribution of the paper -->
+This paper develops a practical and straightforward data-driven pipeline to rapidly
 assess the risk from uncontrolled re-entry of space debris for the European

### 2025-05-02 paper.qmd
organisational skeleton for conclusions

@@
 
+<!-- opening statement linking back to abstract/intro and movtivation for this work --------->
+Space debris re-entry is an increasing risk to aviation and the general public.
+
+<!-- what was shown in the paper ------------------------------------------------------------>
+This paper expands on previous work and ...
+
+<!-- benefits / strengths / weaknesses of the approach --------------------------------------->
+
+<!-- application of the results -------------------------------------------------------------->
 We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
@@
 
+<!-- future work ----------------------------------------------------------------------------->
 Future work will explore the impact of megacostellations and the risk for
 European airspace.
+
+<!-- if any final ~ highly inspirational thought / words ------------------------------------->
 

### 2025-05-02 paper.qmd
better wording

--- 2025-05-01 paper.qmd
@@
       - ref: aiu
-    orcid: 0000-0003-2810-1404
-    email: rainer.koelle@eurocontrol.int
   - name: Quinten Goens
@@
       - ref: aiu
-    orcid: 0000-0003-2187-1924
-    email: quinten.goens@eurocontrol.int
 affiliations:
@@
 National and international authorities need to balance safety and economic
-impact of adopted measures should a re-entry event be announced, usually on short
+impact of adopted measures should a reentry event be announced, usually on short
 notice.  
-We employ hourly and daily spatial flight densities calculated from
+We employ hourly and daily spatial density of flight calculated from
 trajectories as recorded by EUROCONTROL's Network Manager and
-binned with the Uber H3 geospatial indexing system.  
+binned over Uber's H3 gridding system.  
 Impact risk probabilities are then calculated based on previous theoretical
-work.
-This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
+work but using these flights trajectories instead of proxies such as
+population density or ADS-B/Mode S message counts.
 These densities combined with a buffered area along the eventual path of
-re-entry, can provide mid-air impact probabilities to authorities as an aid
+re-entry, can provide risk impact probabilities to authorities as an aid
 for deciding on safety measures such as airspace closures and/or re-routings.
@@
 
-With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
-race to commercialise space it is clear a re-evaluation of the risks of
-collision and casualties due to uncontrolled debris re-entries is needed.
-Overseeing authorities need methodologies and tools to assess which 
-safety measures are the most adequate to minimise potential casualties and economic
+The ever-growing forecast number of flights, see @fig-forecast, and the modern
+race to commercialisation of space call for a reevaluation of the risk of
+collision and casuality due to uncontrolled debris reentries.
+Oversighting authorities need methodologies and tools to assess which 
+safety measures are the most adequate to minimize casualties and economic
 impact.
@@
 Surging number of flights in Europe [@forecastUpdate2025] and in the
-world, but more importantly under development/deployment mega-constellations
-of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
+world, but more importantly under development/deployment megaconstellations
+of satellites on \acr{LEO} such as already active Starlink and OneWeb or
 planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
-of orbiting objects with a short operational life. Some of these will 
-inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].
+of orbiting objects with a short operational life some of which will 
+inevitably reenter Earth atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].
 
 This paper develops a practical and straightforward pipeline to rapidly
-assess the risk from uncontrolled re-entry of space debris for the European
+assess the risk from uncontrolled reentry of space debris for the European
 region.
-Apart from the general risk assessment of potential re-entries, 
-a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus is analysed.
+Apart from the general risk assessment of potential reentries, we analyze
+a soon to happen reentry of debris from a 1972 failed USSR mission to Venus.
 
@@
     constellation ~ "Constellation",
-    state ~ "Country",
     launched ~ "Launched",
@@
 approximating flight density with population density [@patera2008] or
-aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
-flight trajectories.
+aircraft transponder data [@wrightAirspaceClosuresDue2025], we use flown 
+trajectories.
 
-The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
-recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
-[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
-This work will use H3 cells at resolution 3 which corresponds to hexagons of an
-average area $\qty{12393}{km^2}$ and an average edge length of
+We base our hourly flight density estimation on flown \acr{IFR} trajectories as
+recorded by EUROCONTROL's \acr{NM}[^1] and binned on Uber's H3 gridding system
+[@uber-h3] for fast processing.  
+This work will use H3 cells at resolution 3 which corresponds to hexagons of
+average area $\qty{12393}{km^2}$ and average edge length of
 $\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
@@
 H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
-box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).
+box of 43 EUROCONTROL Member States FIR's (blue shaded polygon).
 
@@
 
-The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.
+We characterise densities on an hourly basis for the days of the 2023 and 2024
+busiest \acr{IATA} summer ISO week.
 
-This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
-provides a reproducible example for States, Agencies or relevant authorities
+We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
+provide a reproducible example for States, Agencies or relevant authorities
 to build upon.
@@
 
-Next sections will develop the mathematical background for the evaluation of
-collision and casualties risk of which traffic density is a key component.
+Next sections will develop the mathematcal background for the evaluation of
+collision and casuality risk of which traffic density is a key component.
 
-@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
+@fig-density-summer2024-res-3 shows hourly flight density for 2024-07-05.
 Expected outcomes such as cells at zero density for war or no-fly zones
@@
 
-Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
+Flight density in the EUROCONTROL Area on 2024-07-05, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
 :::
 
-## Impact probability for a single flight and single debris
+## Impact Probability for Single Flight, Single Debris
 The probability, $p_k$, of a debris (point) object $k$, impacting an
@@
 
-Using H3 to discretise the area of interest $a$ using H3 cells gives
+If we discretize the area of interest $a$ using H3 cells,
+we have
 
 $$
-p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h).
+p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h)
 $$
 
-Where for each cell $h$ the following applies
+So for cell $h$ we have
 
 $$
-p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h).
+p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h)
 $$
@@
 
-It is assumed the aircraft impact area to be much larger than the debris
-object area so as to neglect the debris impact area.
-The geometry of the aircraft is simplified and modelled as
-a rectangular cuboid with dimensions of wingspan, length and height. 
+We assume the aircraft impact area to be much larger than the debris
+object area so as to neglet the debris impact area.
+We can also simplify the geometry of the aircraft and model it as
+a rectangular solid with dimesions of wingspan ($W$), length ($L$) and
+height ($H$). 
 Now, if the aircraft were stationary, the area exposed to a debris impact would
 be the top view, while if the debris were stationary it would be the front view.
-In reality an aircraft en-route is moving (mainly) horizontally and the debris 
+In reality the aircraft is moving (manly) horizontally and the debris 
 vertically [^balistic_descent], which which allows to express the effective
@@
 
-In the calculations the following is assumed to be true ($t$ subscript for aircraft type)
+In our calculations we will use ($t$ subscript for aircraft type)
 
@@
 
-[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
-path angle of a re-entering object at 40,000 feet is -86.6 degrees, which
-corresponds to nearly vertical descent.
+[^balistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
+the flight path angle of a reentering object at \qty{40000}{ft} is 
+\ang{-86.6}, which corresponds to nearly vertical descent.
 Since most debris objects have ballistic coefficients equal to or 
-less than 50 lbs/ft2, the vertical descent assumption is a very
+less than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
 accurate approximation.* [@patera2008]
@@
 Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
-circular orbits prior to re-entry even if initially on elliptical
+circular orbits prior to reentry even if initially on elliptical
 ones.[^circular-orbit]
 
-The angular position, $\theta$, of the re-entry point from the ascending node, 
+The angular position, $\theta$, of the reentry point from the ascending node, 
 see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
@@
 
-If we consider historical debris re-entry data, we can numerically derive
-the debris impact probability as a weighting function by summing up and averaging
-all single densities:
+If we consider historical uncontrolled debris reentry data, we can numerically
+derive the debris impact probability as a weighting function by summing up and
+averaging all single densities:
 
@@
 
-From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
+From [@gcat] we extracted 1968 debris reentries during 25 calendar years from 
 1st Jan 2000 till 31st Dec 2024.
@@
 
-Weighting function derived from 25 years of re-entries as an
+Weighting function derived from 25 years of reentries as an
 empirical approximation of $\sigma(\varphi)$. 
@@
 ## Hourly impact probability per H3 cell
-By combining the hourly aircraft density (per aircraft type) with the relevant
-weightings per H3 cell, we can numerically compute the hourly impact probability
-for each hexagonal cell as:
+
+When combining the hourly ($H$) aircraft density (per aircraft type) with the
+relevant weightings per H3 cell, $h$, we can numerically compute the hourly
+impact probability for each hexagonal cell as:
 
 $$
- E_h =  \sum_t W(h) \cdot \rho_{t}(h) \cdot \varepsilon_t
+ E_{h, H} =  \sum_t W(h) \cdot \rho_{t,H}(h) \cdot \varepsilon_t
 $$
 
-where $E_h$ is the collision expectation in the hexagon $h$,
-$W(h)$ is the median value of the weighting function over the area of hexagon $h$, $t$ is the \acr{ICAO} aircraft type,
-$\rho_{t}(h)$ is the traffic density for type $t$ in cell $h$ and
-$\varepsilon_T$ is effective exposed area for aircraft of type $t$.
+where $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
+$W(h)$ is the median value of the weighting function over the area of hexagon
+$h$,
+$\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
+cell $h$ at hout $H$ and
+$\varepsilon_t$ is effective exposed area for aircraft of type $t$.
 
@@
 #| echo: false
-knitr::include_graphics("media/example-image.png")
+knitr::include_graphics(here::here("media", "figures", "collision_expectation_2024-07-05_3.png"))
 ```
 
-Collision risk probability. 
+Averaged hourly collision expectation on 2024-07-05. 
 :::
 
-## Use Case: KOSMOS 482 Descent Craft re-entry
+## Use Case: KOSMOS 482 Descent Craft reentry
 
-In few days (at the time of writing this paper) at 6h May 10, 2025
+In few days (at the time of writing of this paper) at 6h May 10, 2025
 ($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
-Earth's atmosphere [@kosmos482re-entry].
+Earth's atmosphere [@kosmos482reentry].
 This is supposedly the lander module from a 1972 failed Soviet Venera mission
-[@SpaceReviewKosmos] of \qty{480}{kg} weight and circa \qty{1}{m} diameter.
-Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \qty{467}{degreesCelsius} (\qty{872}{degreesFarenheit})).
-So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
+[@SpaceReviewKosmos] of \qty{480}{kg} weight and approximately
+\qty{1}{m} diameter.
+Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
+it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
 
-With what developed in this paper, we can derive the casuality risk and
-determine the maximum value using data for the busiest day in 
-the busiest 2024 week.
+Knowing that the debris inclination is $\ang{51.7}$,
+see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a
+proxy for the traffic density on the day of reentry.
 
-Once the re-entry track is known more or less accurately, the relevant
-H3 cells can be
+Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
+can be identified to retrieve the values of collision expectancy.
 
@@
 
-Casuality for KOSMOS 482 Descent Craft re-entry. 
+Casuality for KOSMOS 482 Descent Craft reentry. 
 :::
@@
 
-We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
+We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled reentry of space objects that could pose a safety concern on civilian aviation.
 The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.
 
-Future work will explore the impact of megacostellations and the risk for
-European airspace.
+Future work will explore the impact of megacostellations and flight growth
+forecast and what would happen in 20, 40, ..., 100 years in varios scenarios
+of uncontrolled  reentries and the risk for European airspace.
 
-# Appendix {.appendix}
-# Orbital Elements {#sec-orbitals }
+# Appendix: Orbital Elements {#sec-orbitals}
 
@@
 
-The angular position, $\theta$, of the re-entry point from the ascending node
+The angular position, $\theta$, of the reentry point from the ascending node
 in the orbit plane is $\theta = \omega + \nu$.

### 2025-05-02 paper.qmd

--- 2025-05-02 paper.qmd
@@
 
-It is assumed the aircraft impact area to be much larger than the debris
-object area so as to neglect the debris impact area.
-The geometry of the aircraft is simplified and modelled as
-a rectangular cuboid with dimensions of wingspan, length and height. 
+We assume the aircraft impact area to be much larger than the debris
+object area so as to neglet the debris impact area.
+We can also simplify the geometry of the aircraft and model it as
+a rectangular solid with dimesions of wingspan ($W$), length ($L$) and
+height ($H$). 
 Now, if the aircraft were stationary, the area exposed to a debris impact would
@@
 #| echo: false
-knitr::include_graphics("media/example-image.png")
+knitr::include_graphics(here::here("media", "figures", "collision_expectation_2024-07-05_3.png"))
 ```
 
-Collision risk probability. 
+Averaged hourly collision expectation on 2024-07-05. 
 :::
@@
 
-In few days (at the time of writing this paper) at 6h May 10, 2025
+In few days (at the time of writing of this paper) at 6h May 10, 2025
 ($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
@@
 This is supposedly the lander module from a 1972 failed Soviet Venera mission
-[@SpaceReviewKosmos] of \qty{480}{kg} weight and circa \qty{1}{m} diameter.
-Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \qty{467}{degreesCelsius} (\qty{872}{degreesFarenheit})).
-So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
+[@SpaceReviewKosmos] of $\qty{480}{kg}$ weight and approximately
+$\qty{1}{m}$ diameter.
+Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
+it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
 
-With what developed in this paper, we can derive the casuality risk and
-determine the maximum value using data for the busiest day in 
-the busiest 2024 week.
+Knowing that the debris inclination is $\ang{51.7}$,
+see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a
+proxy for the traffic density on the day of reentry.
 
-Once the re-entry track is known more or less accurately, the relevant
-H3 cells can be
+Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
+can be identified to retrieve the values of collision expectancy.
+It can then be decided whether it is advisable to close the airspace and
+traffic around the area afftected.
 
@@
 
-<!-- future work ----------------------------------------------------------------------------->
-Future work will explore the impact of megacostellations and the risk for
-European airspace.
+Future work will explore the impact of megacostellations and flight growth
+forecast and what would happen in 20, 40, ..., 100 years in varios scenarios
+of uncontrolled  reentries and the risk for European airspace.
 
-<!-- if any final ~ highly inspirational thought / words ------------------------------------->
+# Appendix: Orbital Elements {#sec-orbitals}
 
-# Appendix {.appendix}
-# Orbital Elements {#sec-orbitals }
-
-:::{#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="t"}
+:::{#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
 

### 2025-05-03 paper.qmd
Update of introduction

@@
 # Abstract
-<!-- teaser --> The risk of collision between uncontrolled re-entering space debris and aircraft
+<!-- teaser --> The recent decades observed an increasing number of space activities and associated launches. 
+The risk of collision between uncontrolled re-entering space debris and aircraft
 is  increasing due to both ever-growing number of airborne flights and 
@@
 Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
-These developments helped to identify options for the wider use of space, benefits, and LALALALA.
+These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
 This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
@@
 collision and casualties due to uncontrolled debris re-entries.
-Overseeing authorities need methodologies and tools to assess which 
+The increasing risk landscape poses challenges in terms of oversight and risk management.
+There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
 safety measures are the most adequate to minimise potential casualties and economic
@@
 
-Surging number of flights in Europe [@forecastUpdate2025] and in the
-world, but more importantly under development/deployment mega-constellations
-of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
-planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
-of orbiting objects with a short operational life. Some of these will 
-inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].
+The increase in space operations does not exist in a vacuum. 
+The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
+On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
+such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
+Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].
 
@@
 
-...
+Aviation safety risk management is a well-established domain in air transportation. 
+Despite the increasing need for situational awareness of space objects and their re-entry, the classical risk avoidance paradigm revolves around restricting the airspace.
+While the probability of a strike is low, it can result in catastrophic consequences. 
+The growing number of space operations requires NEW - ADVANCED TOOLS. PREVIOUS WORK DID A,B, and MAYBE C - WE DO SOMETHING ON TOP.
 

### 2025-05-03 paper.qmd
some updates to method section

@@
 
-We follow the same mathematical approach of 
-[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
-approximating flight density with population density [@patera2008] or
+<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
+The modelling and tracking of uncontrolled re-entries poses challenges.
+Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
+Common approaches employ a mix of handingling the uncertainty of object or debris movement and exposed area, including temporal uncertainty windows.
+
+      <!-- is there more to say and provide context? -->
+
+<!-- APPROACH / METHOD --------------------------------------------------- -->
+<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
+<!-- as a minimum, we should say something about the WHAT -->
+In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
+In that respect, a robust determination of flight density is a key aspect.
+
+We follow the same mathematical approach of [@patera2008; @wrightAirspaceClosuresDue2025].
+However, this paper augments the approximation of risk exposure. 
+Instead of approximating flight density with population density [@patera2008] or
 aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
@@
 
+This work focusses on the European region.
 The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
-recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
-[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
-This work will use H3 cells at resolution 3 which corresponds to hexagons of an
-average area $\qty{12393}{km^2}$ and an average edge length of
+recorded by EUROCONTROL's \acr{NM}[^1].
+To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
+The reduction of the spatial complexity enables fast processing.
+For this work, we employ a H3-cell-resolution of 3. 
+This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
 $\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
+
+<!-- CAN WE MOTIVATE WHY resolution 3??? -- e.g. is this an area used in other works, or does it do something? -->
 
@@
 
-The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.
+To build an upper threshold for flight exposure, this study uses the busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.   <!-- does ISO do here anything? --> 
+The flight densities are characterised on an hourly basis for all days in these weeks.
 
-This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
-provides a reproducible example for States, Agencies or relevant authorities
-to build upon.
+<!-- next sentence feels like an orphan ... do we need this here? ------>
+This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.
 
-[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States
-and 2 Comprehensive Member States covering
-almost all \acr{ECAC} States.
+[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States.
 
-Next sections will develop the mathematical background for the evaluation of
-collision and casualties risk of which traffic density is a key component.
+<!-- next sentence feels like an orphan ... do we need this here? ------>
+Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.
 
 @fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
-Expected outcomes such as cells at zero density for war or no-fly zones
-(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
-and East-bound over Türkiye) are clearly highlighted.
+The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
+For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
+(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
+and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.
 

### 2025-05-03 paper.qmd
Update paper.qmd

@@
 
-where $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
-$W(h)$ is the median value of the weighting function over the area of hexagon
+where 
+* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
+* $W(h)$ is the median value of the weighting function over the area of hexagon
 $h$,
-$\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
-cell $h$ at hout $H$ and
-$\varepsilon_t$ is effective exposed area for aircraft of type $t$.
+* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
+cell $h$ at hour $H$, and
+* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.
+
+@fig-collision depicts the determined hourly collision expectation.
+As identified in @fig-density-summer2024-res-3 areas of higher traffic density observe a higher collision expectation.
+<!-- WHAT ELSE CAN WE SAY ABOUT THIS PLOT? -->
 
@@
 
-## Use Case: KOSMOS 482 Descent Craft re-entry
+# Application Use Case: KOSMOS 482 Descent Craft re-entry
+<!-- I would make this a separate section in the paper -->
 
@@
 Earth's atmosphere [@kosmos482re-entry].
-This is supposedly the lander module from a 1972 failed Soviet Venera mission
-[@SpaceReviewKosmos] of $\qty{480}{kg}$ weight and approximately
-$\qty{1}{m}$ diameter.
+This is supposedly the lander module from a 1972 failed Soviet Venera mission [@SpaceReviewKosmos]
+The module is reported to weigh about $\qty{480}{kg}$ and has a diameter of approximately $\qty{1}{m}$.
 Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
@@
 
-Knowing that the debris inclination is $\ang{51.7}$,
-see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a
-proxy for the traffic density on the day of reentry.
+Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of reentry.
 
@@
 It can then be decided whether it is advisable to close the airspace and
-traffic around the area afftected.
+traffic around the area affected.
 
@@
 <!-- opening statement linking back to abstract/intro and movtivation for this work --------->
-Space debris re-entry is an increasing risk to aviation and the general public.
+Space objects and debris re-entry is an increasing risk to aviation and the general public.
+The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
+Dependent on the mission, objects re-entering the atmosphere can take on differnt forms, shapes, and sizes.
+On top, a higher number of these entries may be uncontrolled.
+With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accomodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
 
 <!-- what was shown in the paper ------------------------------------------------------------>
-This paper expands on previous work and ...
+This paper expands on previous work.
+With a focus on Europe, we establish an upper bound for the traffic density based on actual flights.
+
+... SOME MORE STUFF DONE IN THE PAPER ...
+
+The findings are applied to a use-case: the imminent re-entry of the lander module of an earlier Venus mission.
+The maximum collision expectation for the busiest day in 2024 is determined.
 
 <!-- benefits / strengths / weaknesses of the approach --------------------------------------->
+
+The approach presented in this paper builds on previous research.
+Our approach utilises H3-tesselation to reduce the spatial complexity.
+This cell indexing approach reduces the computation significantly.
+
+        WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
+
+<!-- future work ---------------------------------------------------------------------------->
+
+Future work will address the shortcomings. <!-- or can we say something about the next concrete steps to remedy some weaknesses -->
 
@@
 
-Future work will explore the impact of megacostellations and flight growth
-forecast and what would happen in 20, 40, ..., 100 years in varios scenarios
-of uncontrolled  reentries and the risk for European airspace.
+As part of this research and to inform the decision-makers it is planned to explore the impact of megacostellations and flight growth forecast and what would happen in the mid- to long-term in varios scenarios of uncontrolled re-entries and the risk for European airspace.
 

### 2025-05-04 paper.qmd
method - impact probability - some editorial tweaks

@@
 # Abstract
-<!-- teaser --> The recent decades observed an increasing number of space activities and associated launches. 
+<!-- teaser --> The recent decades observed an increasing number of space activities resulting in higher numbers of re-entries at the end of the mission lifespan or orbital decay. 
 The risk of collision between uncontrolled re-entering space debris and aircraft
@@
 Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
-Common approaches employ a mix of handingling the uncertainty of object or debris movement and exposed area, including temporal uncertainty windows.
+Common approaches employ a mix of handling the uncertainty of object or debris movement, the associated exposed area, including temporal uncertainty windows.
 
@@
 ## Impact probability for a single flight and single debris
-The probability, $p_k$, of a debris (point) object $k$, impacting an
+
+Safety and collision risk awareness is directly linked to the probability of a re-entering space object colliding with a transversing flight.
+This probability, $p_k$, of a debris (point) object $k$, impacting an
 aircraft (point) object over the area, $a$, is expressed as:
@@
 
-where $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
-$\sigma(\vec{x})$ is the impact probability density of a debris object $k$
-in the position $\vec{x}$ and $\varepsilon_k$ is the effective exposed area,
-i.e. the effective risk area for the $k^{th}$ debris object impacting
+where 
+* $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
+* $\sigma(\vec{x})$ is the impact probability density of a debris object $k$
+in the position $\vec{x}$, and 
+* $\varepsilon_k$ is the effective exposed area, i.e., the effective risk area for the $k^{th}$ debris object impacting
 the aircraft (see @sec-eea later on).
 
-Using H3 to discretise the area of interest $a$ using H3 cells gives
+Our approach builds on the discretisation of the geo-spatial risk area.
+For this purpose we use a hexagon-based tesselation pattern.
+Uber's H3 method ensures non-overlapping hexagons, with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
+Using H3 to discretise the area of interest $a$ gives
 
@@
 
-Where for each cell $h$ the following applies
+The impact probability $p_{k}(h)$ for each cell $h$ is determined by the following
 

### 2025-05-04 paper.qmd
editorial changes to effective expose area and rewritten impact probability

@@
 
-We assume the aircraft impact area to be much larger than the debris
-object area so as to neglet the debris impact area.
-We can also simplify the geometry of the aircraft and model it as
-a rectangular solid with dimesions of wingspan ($W$), length ($L$) and
-height ($H$). 
-Now, if the aircraft were stationary, the area exposed to a debris impact would
-be the top view, while if the debris were stationary it would be the front view.
-In reality an aircraft en-route is moving (mainly) horizontally and the debris 
-vertically [^balistic_descent], which which allows to express the effective
-exposed area, $\varepsilon_k$, as
+For the actual collision encounter between a re-entering object and an aircraft, we assume the aircraft impact area to be much larger than the debris object area.
+Thus, the area spanned by the debris object is smaller than the aircraft dimensions. 
+This assumption holds for practically all commercial aircraft within the scope of this study, and we can neglect the debris impact area.
+Accordingly, we can also simplify the geometry of the aircraft and model it as a rectangular solid with dimesions of wingspan ($W$), length ($L$) and height ($H$).
+
+Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations. 
+<!-- TERMINOLOGY? rectangular solid -- impact --> top plane, front plane? Does the literture use "view"  -->
+1.) if the aircraft were stationary, the area exposed to the debris impact would be the top view, and
+2.) if the debris were stationary, it would be the front view.
+
+Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
+The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
+At the same time, the movement of the re-entered debris object at regular en-route flight levels will be vertically [^balistic_descent].
+This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as
 
@@
 
-where $s_{a}$ is the average aircraft speed, $s_{d_k}$ is the average fall speed
-of the $k$-th debris object, $a_T$ area of the aircraft as viewed from the top,
-$a_F$ area of the aircraft as viewed from the from the front of the aircraft.
+where 
+* $s_{a}$ is the average aircraft speed, 
+* $s_{d_k}$ is the average fall speed of the $k$-th debris object, 
+* $a_T$ area of the aircraft as viewed from the top, and 
+* $a_F$ area of the aircraft as viewed from the from the front of the aircraft.
 
+This can be further instantiated for specific aircraft types.
 In the calculations the following is assumed to be true ($t$ subscript for aircraft type)
@@
 
-The effective exposed area now only depends on the aircraft type $t$ (dimensions
+The effective exposed area now only depends on the aircraft type $t$ (i.e., its dimensions
 and cruise speed) and can be rewritten as:
@@
 
+These aircraft properties are well-known. 
+While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
 The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].
 
-The impact probability for cell $h$ and single aircraft of type $t$ becomes:
+The impact probability for cell $h$ under consideration of a specific aircraft of type $t$ becomes:
 

### 2025-05-04 paper.qmd
reference for velocity of debris

--- 2025-05-03 paper.qmd
@@
 # Abstract
-<!-- teaser --> The recent decades observed an increasing number of space activities and associated launches. 
-The risk of collision between uncontrolled re-entering space debris and aircraft
+<!-- teaser --> The risk of collision between uncontrolled re-entering space debris and aircraft
 is  increasing due to both ever-growing number of airborne flights and 
@@
 Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
-These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
+These developments helped to identify options for the wider use of space, benefits, and LALALALA.
 This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
@@
 collision and casualties due to uncontrolled debris re-entries.
-The increasing risk landscape poses challenges in terms of oversight and risk management.
-There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
+Overseeing authorities need methodologies and tools to assess which 
 safety measures are the most adequate to minimise potential casualties and economic
@@
 
-The increase in space operations does not exist in a vacuum. 
-The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
-On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
-such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
-Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].
+Surging number of flights in Europe [@forecastUpdate2025] and in the
+world, but more importantly under development/deployment mega-constellations
+of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
+planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
+of orbiting objects with a short operational life. Some of these will 
+inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].
 
@@
 
-Aviation safety risk management is a well-established domain in air transportation. 
-Despite the increasing need for situational awareness of space objects and their re-entry, the classical risk avoidance paradigm revolves around restricting the airspace.
-While the probability of a strike is low, it can result in catastrophic consequences. 
-The growing number of space operations requires NEW - ADVANCED TOOLS. PREVIOUS WORK DID A,B, and MAYBE C - WE DO SOMETHING ON TOP.
+...
 
@@
 
-<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
-The modelling and tracking of uncontrolled re-entries poses challenges.
-Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
-Common approaches employ a mix of handingling the uncertainty of object or debris movement and exposed area, including temporal uncertainty windows.
-
-      <!-- is there more to say and provide context? -->
-
-<!-- APPROACH / METHOD --------------------------------------------------- -->
-<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
-<!-- as a minimum, we should say something about the WHAT -->
-In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
-In that respect, a robust determination of flight density is a key aspect.
-
-We follow the same mathematical approach of [@patera2008; @wrightAirspaceClosuresDue2025].
-However, this paper augments the approximation of risk exposure. 
-Instead of approximating flight density with population density [@patera2008] or
+We follow the same mathematical approach of 
+[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
+approximating flight density with population density [@patera2008] or
 aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
@@
 
-This work focusses on the European region.
 The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
-recorded by EUROCONTROL's \acr{NM}[^1].
-To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
-The reduction of the spatial complexity enables fast processing.
-For this work, we employ a H3-cell-resolution of 3. 
-This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
+recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
+[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
+This work will use H3 cells at resolution 3 which corresponds to hexagons of an
+average area $\qty{12393}{km^2}$ and an average edge length of
 $\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
-
-<!-- CAN WE MOTIVATE WHY resolution 3??? -- e.g. is this an area used in other works, or does it do something? -->
 
@@
 
-To build an upper threshold for flight exposure, this study uses the busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.   <!-- does ISO do here anything? --> 
-The flight densities are characterised on an hourly basis for all days in these weeks.
+The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.
 
-<!-- next sentence feels like an orphan ... do we need this here? ------>
-This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.
+This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
+provides a reproducible example for States, Agencies or relevant authorities
+to build upon.
 
-[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States.
+[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States
+and 2 Comprehensive Member States covering
+almost all \acr{ECAC} States.
 
-<!-- next sentence feels like an orphan ... do we need this here? ------>
-Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.
+Next sections will develop the mathematical background for the evaluation of
+collision and casualties risk of which traffic density is a key component.
 
 @fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
-The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
-For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
-(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
-and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.
+Expected outcomes such as cells at zero density for war or no-fly zones
+(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
+and East-bound over Türkiye) are clearly highlighted.
 
@@
 * constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
+  [[pag 8 @patera2008]]
 * constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
@@
 
-The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].
+The values $W_t$, $H_t$ and $L_t$ have been collected from [@performance_database;@skybrary;@doc8643]; we used \qty{1000}{m^2} for 
+$\varepsilon_t$ in case of missing aircraft type.
 
@@
 
-where 
-* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
-* $W(h)$ is the median value of the weighting function over the area of hexagon
+where $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
+$W(h)$ is the median value of the weighting function over the area of hexagon
 $h$,
-* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
-cell $h$ at hour $H$, and
-* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.
-
-@fig-collision depicts the determined hourly collision expectation.
-As identified in @fig-density-summer2024-res-3 areas of higher traffic density observe a higher collision expectation.
-<!-- WHAT ELSE CAN WE SAY ABOUT THIS PLOT? -->
+$\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
+cell $h$ at hout $H$ and
+$\varepsilon_t$ is effective exposed area for aircraft of type $t$.
 
@@
 
-# Application Use Case: KOSMOS 482 Descent Craft re-entry
-<!-- I would make this a separate section in the paper -->
+## Use Case: KOSMOS 482 Descent Craft re-entry
 
@@
 Earth's atmosphere [@kosmos482re-entry].
-This is supposedly the lander module from a 1972 failed Soviet Venera mission [@SpaceReviewKosmos]
-The module is reported to weigh about $\qty{480}{kg}$ and has a diameter of approximately $\qty{1}{m}$.
+This is supposedly the lander module from a 1972 failed Soviet Venera mission
+[@SpaceReviewKosmos] of $\qty{480}{kg}$ weight and approximately
+$\qty{1}{m}$ diameter.
 Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
@@
 
-Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of reentry.
+Knowing that the debris inclination is $\ang{51.7}$,
+see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a
+proxy for the traffic density on the day of reentry.
 
@@
 It can then be decided whether it is advisable to close the airspace and
-traffic around the area affected.
+traffic around the area afftected.
 
@@
 <!-- opening statement linking back to abstract/intro and movtivation for this work --------->
-Space objects and debris re-entry is an increasing risk to aviation and the general public.
-The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
-Dependent on the mission, objects re-entering the atmosphere can take on differnt forms, shapes, and sizes.
-On top, a higher number of these entries may be uncontrolled.
-With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accomodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
+Space debris re-entry is an increasing risk to aviation and the general public.
 
 <!-- what was shown in the paper ------------------------------------------------------------>
-This paper expands on previous work.
-With a focus on Europe, we establish an upper bound for the traffic density based on actual flights.
-
-... SOME MORE STUFF DONE IN THE PAPER ...
-
-The findings are applied to a use-case: the imminent re-entry of the lander module of an earlier Venus mission.
-The maximum collision expectation for the busiest day in 2024 is determined.
+This paper expands on previous work and ...
 
 <!-- benefits / strengths / weaknesses of the approach --------------------------------------->
-
-The approach presented in this paper builds on previous research.
-Our approach utilises H3-tesselation to reduce the spatial complexity.
-This cell indexing approach reduces the computation significantly.
-
-        WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
-
-<!-- future work ---------------------------------------------------------------------------->
-
-Future work will address the shortcomings. <!-- or can we say something about the next concrete steps to remedy some weaknesses -->
 
@@
 
-As part of this research and to inform the decision-makers it is planned to explore the impact of megacostellations and flight growth forecast and what would happen in the mid- to long-term in varios scenarios of uncontrolled re-entries and the risk for European airspace.
+Future work will explore the impact of megacostellations and flight growth
+forecast and what would happen in 20, 40, ..., 100 years in varios scenarios
+of uncontrolled  reentries and the risk for European airspace.
 

### 2025-05-04 paper.qmd

--- 2025-05-04 paper.qmd
@@
 While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
-The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].
+The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used \qty{1000}{m^2} for 
+$\varepsilon_t$ in case of missing aircraft type.
 

### 2025-05-04 python/resample_aug.py
remove useless files; resampling implemented in R

@@ -1,8 +1,0 @@
-from traffic.core import Traffic
-
-(Traffic.from_file('data-raw/trjs_2024-08-01.parquet')
-  .resample(rule = "30s")
-  .eval()
-  .to_parquet('data/trajectories_2024-08-01_resampled_30s.parquet',
-              index = False)
-)

### 2025-05-04 python/resample_dec.py

@@ -1,8 +1,0 @@
-from traffic.core import Traffic
-
-(Traffic.from_file('data-raw/trjs_2024-12-05.parquet')
-  .resample(rule = "30s")
-  .eval()
-  .to_parquet('data/trajectories_2024-12-05_resampled_30s.parquet',
-              index = False)
-)

### 2025-05-04 R/plot-collision-risk.R
plot collision risk

@@ -1,0 +1,154 @@
+library(conflicted)
+library(tidyverse)
+library(aviodebris)
+library(arrow)
+library(sf)
+library(h3jsr)
+library(ggplot2)
+library(magrittr)
+library(giscoR)
+library(grid)
+library(here)
+library(latex2exp)
+
+conflicts_prefer(dplyr::filter)
+
+date <- "2024-07-05" |> as_date()
+density <- arrow::read_parquet(
+  str_glue("data/collision_and_casuality_{date}_res_3_hourly.parquet")
+)
+
+resolution <- 3
+
+eur_hex <- bbox_nm() |>
+  hexes_for_bbox_at_res(resolution = resolution)
+
+
+ddd <- density |>
+  group_by(cell) |>
+  summarise(collision_expectation = sum(collision_expectation) / 24) |>
+  ungroup() |>
+  filter(cell %in% (eur_hex |> pull(h3_address))) |>
+  cell_to_polygon(simple = FALSE)
+
+MAX <- ddd |> pull("collision_expectation") |> max()
+min <- ddd |> pull("collision_expectation") |> min()
+# min <- 0
+
+eur_hex_union <- bbox_nm() |>
+  bbox_of_hexes_for_bbox_at_res(resolution = resolution)
+
+gisco_RG <- gisco_get_countries(
+  resolution = "20",
+  epsg = "4326",
+  spatialtype = "RG",
+  region = c("Europe", "Africa", "Asia")
+) |>
+  mutate(res = "20M")
+
+
+EUR_res20 <- gisco_RG |>
+  st_intersection(eur_hex_union) |>
+  select(res)
+
+# draw hexagon for key shape in legend
+draw_key_hex <- function(data, params, size) {
+  # make fill inherit color if NA
+  if (is.na(data$fill)) data$fill <- data$col
+
+  # hexagon vertex coordinates
+  v <- list(
+    x = c(0.95, 0.725, 0.275, 0.05, 0.275, 0.725),
+    y = c(
+      0.5,
+      0.110288568297003,
+      0.110288568297003,
+      0.5,
+      0.889711431702997,
+      0.889711431702997
+    )
+  )
+  # hexagon grob
+  grid::polygonGrob(
+    v$x,
+    v$y,
+    gp = grid::gpar(col = data$colour, fill = alpha(data$fill, data$alpha))
+  )
+}
+
+
+ggplot() +
+  geom_sf(
+    data = eur_hex,
+    fill = "#7DF9FF",
+    colour = NA,
[…]

### 2025-05-04 R/plot-density-map.R
regenerate plot

@@ -17,10 +17,10 @@ density <- arrow::read_parquet(
   str_glue("data/traffic_density_{date}_res_3_hourly.parquet")
 )
 
-res <- 3
+resolution <- 3
 
 eur_hex <- bbox_nm() |>
-  hexes_for_bbox_at_res(resolution = res)
+  hexes_for_bbox_at_res(resolution = resolution)
 
 
 ddd <- density |>
@@ -36,7 +36,7 @@ min <- 0
 
 
 eur_hex_union <- bbox_nm() |>
-  bbox_of_hexes_for_bbox_at_res(resolution = res)
+  bbox_of_hexes_for_bbox_at_res(resolution = resolution)
 
 gisco_RG <- gisco_get_countries(
   resolution = "20",
@@ -46,9 +46,8 @@ gisco_RG <- gisco_get_countries(
 ) |>
   mutate(res = "20M")
 
-gisco <- gisco_RG
 
-EUR_res20 <- gisco |>
+EUR_res20 <- gisco_RG |>
   st_intersection(eur_hex_union) |>
   select(res)
 
@@ -74,30 +73,6 @@ draw_key_hex <- function(data, params, size) {
     v$x,
     v$y,
     gp = grid::gpar(col = data$colour, fill = alpha(data$fill, data$alpha))
-  )
-}
-
-"%||%" <- function(a, b) {
-  if (!is.null(a)) a else b
-}
-
-# Copy & Paste from ggplot2::draw_key_rect
-draw_key_cust <- function(data, params, size) {
-  # make fill inherit color if NA
-  if (is.na(data$fill)) data$fill <- data$col
-
-  grid::rectGrob(
-    gp = grid::gpar(
-      col = NA,
-      fill = alpha(
-        data$fill %||%
-          data$colour %||%
-          "grey20",
-        data$alpha
-      ),
-      lty = data$linetype %||%
-        1
-    )
   )
 }
 
@@ -178,5 +153,5 @@ ggplot() +
 ggsave(here(
   "media",
   "figures",
-  str_glue("flight_density_{date}_{res}.png")
+  str_glue("flight_density_{date}_{resolution}.png")
 ))

### 2025-05-04 R/plot-weighting-latitude.R
clean plot code and regenerate figure

@@ -36,9 +36,9 @@ europe_hex <- function(res_h3, res_spatial = "20M") {
     select(res)
 }
 
-res <- 3L
+resolution <- 3L
 
-eur_hex <- bbox_nm() |> hexes_for_bbox_at_res(resolution = res)
+eur_hex <- bbox_nm() |> hexes_for_bbox_at_res(resolution = resolution)
 
 ddd <- weightings_h3_resolution_3_hourly |>
   filter(h3_resolution_3 %in% (eur_hex |> pull(h3_address))) |>
@@ -55,7 +55,7 @@ ddd |>
   ) +
   labs(
     title = "weightings on the spherical Earth",
-    subtitle = "H3 hexes (res = 3)"
+    subtitle = "H3 hexes (resolution = 3)"
   )
 
 

### 2025-05-05 paper.qmd
Went over the spellcheck to fix many typos

--- 2025-05-04 paper.qmd
@@
 title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"
-
 filters:
@@
   - abstract-section
-
 format:
@@
       - text: |
+          \let\qty\SI
           % NOTE: make siunitx's \qty work without clashing with the one in physics
-          \let\qty\SI
   ieee-html: 
@@
       file: media/siunitx.html
-
 author:
@@
     country: Belgium
-
 bibliography: media/references.bib
-
 acronyms:
@@
     - ./media/acronyms.yml
+editor: 
+  markdown: 
+    wrap: 72
+---
+```{r}
+```
 
----
+```{r}
+```
 
 # Abstract
-<!-- teaser --> The recent decades observed an increasing number of space activities resulting in higher numbers of re-entries at the end of the mission lifespan or orbital decay. 
-The risk of collision between uncontrolled re-entering space debris and aircraft
-is  increasing due to both ever-growing number of airborne flights and 
-of orbiting bodies (and relevant re-entries). 
-<!-- why important --> National and international authorities need to balance safety and economic
-impact of adopted measures should a re-entry event be announced, usually on short
-notice. <!-- no hard line breaks in abstracts -->
-<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
-trajectories as recorded by EUROCONTROL's Network Manager and
-binned with the Uber H3 geospatial indexing system.  
-Impact risk probabilities are then calculated based on previous theoretical
-work. 
-<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
-These densities combined with a buffered area along the eventual path of
-re-entry, can provide mid-air impact probabilities to authorities as an aid
-for deciding on safety measures such as airspace closures and/or re-routings.
+
+<!-- teaser -->
+
+The recent decades observed an increasing number of space activities
+resulting in higher numbers of re-entries at the end of the mission
+lifespan or orbital decay. The risk of collision between uncontrolled
+re-entering space debris and aircraft is increasing due to both
+ever-growing number of airborne flights and of orbiting bodies (and
+relevant re-entries). <!-- why important --> National and international
+authorities need to balance safety and economic impact of adopted
+measures should a re-entry event be announced, usually on short notice.
+<!-- no hard line breaks in abstracts -->
+<!-- approach / experiments --> We employ hourly and daily spatial
+flight densities calculated from trajectories as recorded by
+EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
+indexing system.\
+Impact risk probabilities are then calculated based on previous
+theoretical work. <!-- results / conclusions --> This work enhances the
+model by using actual flown flight trajectories as a basis for
+time-varying flight densities rather than relying on proxies such as the
+population density or ADS-B/Mode S message counts. These densities
+combined with a buffered area along the eventual path of re-entry, can
+provide mid-air impact probabilities to authorities as an aid for
+deciding on safety measures such as airspace closures and/or
+re-routings.
 
@@
 
-<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
-The quest for space is a driver for technological development.
-Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
-These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
-This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
-The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.
+<!-- teaser opening -->
+
+<!-- might need a rewrite - first jotting down -->
+
+The quest for space is a driver for technological development. Key
+milestones include the first human travel to space and surrounding the
+globe, the first and now many succeeding successful landings on the
+moon, and regular operations of space stations in the orbit. These
+developments helped to identify options for the wider use of space,
+benefits, and its commercialisation. This is showcased by an increasing
+number of activities aimed to generate value through the exploration of
+space ranging from scientific understanding and experimentation,
+sustainable production of goods in a non-gravity environment, services,
+and, more recently, space tourism. The increased number of launches,
+however, also increased the number of artefacts
+<!-- falling down and --> re-entering the atmosphere.
 
 <!-- zooming in on problem -->
-With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
-race to commercialise space, there is a need to re-evaluate the risks of
-collision and casualties due to uncontrolled debris re-entries.
-The increasing risk landscape poses challenges in terms of oversight and risk management.
-There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
-safety measures are the most adequate to minimise potential casualties and economic
-impact.
 
-The increase in space operations does not exist in a vacuum. 
-The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
-On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
-such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
-Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].
+With forecasts predicting a steadily growing number of flights, see
+@fig-forecast, and the modern race to commercialise space, there is a
+need to re-evaluate the risks of collision and casualties due to
+uncontrolled debris re-entries. The increasing risk landscape poses
+challenges in terms of oversight and risk management. There is a growing
+demand by overseeing authorities for new methodologies and tools to
+assess which safety measures are the most adequate to minimise potential
+casualties and economic impact.
+
+The increase in space operations does not exist in a vacuum. The
+continual growth of air traffic - despite temporal bumps - is evidenced
+by the surging number of flights in Europe [@forecastUpdate2025] and
+across the world. On-going development/deployment activities of
+mega-constellations of satellites in Low Earth Orbit \acr{LEO}, such as
+already active Starlink and OneWeb or planned ones (e.g. Kuiper,
+Starshield, Xingwang) will add thousand of orbiting objects with a short
+operational life. Some of these will inevitably re-enter the Earth's
+atmosphere in an uncontrolled manner, c.f. [@tbl-megaconstellations;
+@megaconstellations].
 
@@
 
-Aviation safety risk management is a well-established domain in air transportation. 
-Despite the increasing need for situational awareness of space objects and their re-entry, the classical risk avoidance paradigm revolves around restricting the airspace.
-While the probability of a strike is low, it can result in catastrophic consequences. 
-The growing number of space operations requires NEW - ADVANCED TOOLS. PREVIOUS WORK DID A,B, and MAYBE C - WE DO SOMETHING ON TOP.
+Aviation safety risk management is a well-established domain in air
+transportation. Despite the increasing need for situational awareness of
+space objects and their re-entry, the classical risk avoidance paradigm
+revolves around restricting the airspace. While the probability of a
+strike is low, it can result in catastrophic consequences. The growing
+number of space operations requires NEW - ADVANCED TOOLS. PREVIOUS WORK
+DID A,B, and MAYBE C - WE DO SOMETHING ON TOP.
 
 <!-- contribution of the paper -->
-This paper develops a practical and straightforward data-driven pipeline to rapidly
-assess the risk from uncontrolled re-entry of space debris for the European
-region.
-Apart from the general risk assessment of potential re-entries, 
-a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus is analysed.
+
+This paper develops a practical and straightforward data-driven pipeline
+to rapidly assess the risk from uncontrolled re-entry of space debris
+for the European region. Apart from the general risk assessment of
+potential re-entries, a soon to occur re-entry of debris from a 1972
+failed USSR mission to Venus is analysed.
 
 ::: {#fig-forecast}
-
 ```{r forecast}
@@
 #| label: tbl-megaconstellations
-#| tbl-cap: Megaconstellations (planned > 1000) as per filing to the \acr{ITU}
+#| tbl-cap: Mega-constellations (planned > 1000) as per filing to the \acr{ITU}
 #| tbl-cap-location: bottom
@@
 <!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
+
 The modelling and tracking of uncontrolled re-entries poses challenges.
-Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
-Common approaches employ a mix of handling the uncertainty of object or debris movement, the associated exposed area, including temporal uncertainty windows.
+Actual and historic data is sparse, and the size, shape, and attitude or
+rotation of re-entering objects increase the level of uncertainty for
+modelling. Common approaches employ a mix of handling the uncertainty of
+object or debris movement, the associated exposed area, including
+temporal uncertainty windows.
 
+```         
   <!-- is there more to say and provide context? -->
+```
 
 <!-- APPROACH / METHOD --------------------------------------------------- -->
+
 <!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
+
 <!-- as a minimum, we should say something about the WHAT -->
-In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
-In that respect, a robust determination of flight density is a key aspect.
 
-We follow the same mathematical approach of [@patera2008; @wrightAirspaceClosuresDue2025].
-However, this paper augments the approximation of risk exposure. 
-Instead of approximating flight density with population density [@patera2008] or
-aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
-flight trajectories.
+In general, the method for estimating the risk to commercial aircraft
+from reentering debris objects revolves around estimating the exposure
+of a single flight and the population of flights to the re-entering
+object. In that respect, a robust determination of flight density is a
+key aspect.
 
-This work focusses on the European region.
-The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
-recorded by EUROCONTROL's \acr{NM}[^1].
-To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
-The reduction of the spatial complexity enables fast processing.
-For this work, we employ a H3-cell-resolution of 3. 
-This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
+We follow the same mathematical approach of [@patera2008;
+@wrightAirspaceClosuresDue2025]. However, this paper augments the
+approximation of risk exposure. Instead of approximating flight density
+with population density [@patera2008] or aircraft transponder data
+[@wrightAirspaceClosuresDue2025], we use actual flown flight
+trajectories.
+
+This work focusses on the European region. The hourly flight density
+estimation is based on on flown \acr{IFR} trajectories as recorded by
+EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
+binned using Uber's H3 geospatial indexing system [@uber-h3]. The
+reduction of the spatial complexity enables fast processing. For this
+work, we employ a H3-cell-resolution of 3. This corresponds to hexagons
+of an average area $\qty{12393}{km^2}$ and an average edge length of
 $\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
+
+[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's
+    42 Member States and 2 Comprehensive Member States covering almost
+    all \acr{ECAC} States.
 
@@
 ::: {#fig-cells}
-
 ```{r cells-map}
@@
 
-H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
-box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).
-
+H3 cells (red) at resolution 3 and their centres (blue) covering the
+bounding box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded
+polygon).
 :::
 
-To build an upper threshold for flight exposure, this study uses the busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.   <!-- does ISO do here anything? --> 
-The flight densities are characterised on an hourly basis for all days in these weeks.
+To build an upper threshold for flight exposure, this study uses the
+busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
+<!-- does ISO do here anything? --> The flight densities are
+characterised on an hourly basis for all days in these weeks.
 
 <!-- next sentence feels like an orphan ... do we need this here? ------>
-This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.
 
-[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States.
+This paper expands the works of [@wrightAirspaceClosuresDue2025;
+@patera2008] and provides a reproducible example for States, Agencies or
+relevant authorities to build upon.
 
 <!-- next sentence feels like an orphan ... do we need this here? ------>
-Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.
 
-@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
-The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
-For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
-(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
-and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.
+Next sections will develop the mathematical background for the
+evaluation of collision and casualties risk of which traffic density is
+a key component.
+
+@fig-density-summer2024-res-3 shows hourly flight density for the 5th of
+July 2024. The chart shows the expected outcomes in terms of major
+traffic flows, and areas of high or low traffic. For example, cells with
+a density of zero are clearly visible for areas of war or no-fly zones
+(Ukraine, Libya or Syria). The major traffic flows (North Atlantic,
+Canaries and East-bound over Türkiye) are clearly highlighted showing
+densities of more than 10.
 
 ::: {#fig-density-summer2024-res-3}
-
 ```{r density-winter-res-3}
@@
 
-Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
+Flight density in the EUROCONTROL Area on the 5th of July 2024, the
+busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
 :::
@@
 
-Safety and collision risk awareness is directly linked to the probability of a re-entering space object colliding with a transversing flight.
-This probability, $p_k$, of a debris (point) object $k$, impacting an
-aircraft (point) object over the area, $a$, is expressed as:
+Safety and collision risk awareness is directly linked to the
+probability of a re-entering space object colliding with a transversing
+flight. This probability, $p_k$, of a debris (point) object $k$,
+impacting an aircraft (point) object over the area, $a$, is expressed
+as:
 
@@
 
-where 
-* $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
-* $\sigma(\vec{x})$ is the impact probability density of a debris object $k$
-in the position $\vec{x}$, and 
-* $\varepsilon_k$ is the effective exposed area, i.e., the effective risk area for the $k^{th}$ debris object impacting
-the aircraft (see @sec-eea later on).
+where \* $\rho({\vec{x}})$ is the probability density of an aircraft in
+$\vec{x}$, \* $\sigma(\vec{x})$ is the impact probability density of a
+debris object $k$ in the position $\vec{x}$, and \* $\varepsilon_k$ is
+the effective exposed area, i.e., the effective risk area for the
+$k^{th}$ debris object impacting the aircraft (see @sec-eea later on).
 
-Our approach builds on the discretisation of the geo-spatial risk area.
-For this purpose we use a hexagon-based tesselation pattern.
-Uber's H3 method ensures non-overlapping hexagons, with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
-Using H3 to discretise the area of interest $a$ gives
+Our approach builds on the discretisation of the geospatial risk area.
+For this purpose we use a hexagon-based tessellation pattern. Uber's H3
+method ensures non-overlapping hexagons, with an equidistant
+relationship between all neighbours, and enables an efficient indexing
+of the tessel-cells. Using H3 to discretise the area of interest $a$
+gives
 
@@
 
-The impact probability $p_{k}(h)$ for each cell $h$ is determined by the following
+The impact probability $p_{k}(h)$ for each cell $h$ is determined by the
+following
 
@@
 
-## Aircraft effective exposed area{#sec-eea}
+## Aircraft effective exposed area {#sec-eea}
 
-For the actual collision encounter between a re-entering object and an aircraft, we assume the aircraft impact area to be much larger than the debris object area.
-Thus, the area spanned by the debris object is smaller than the aircraft dimensions. 
-This assumption holds for practically all commercial aircraft within the scope of this study, and we can neglect the debris impact area.
-Accordingly, we can also simplify the geometry of the aircraft and model it as a rectangular solid with dimesions of wingspan ($W$), length ($L$) and height ($H$).
+For the actual collision encounter between a re-entering object and an
+aircraft, we assume the aircraft impact area to be much larger than the
+debris object area. Thus, the area spanned by the debris object is
+smaller than the aircraft dimensions. This assumption holds for
+practically all commercial aircraft within the scope of this study, and
+we can neglect the debris impact area. Accordingly, we can also simplify
+the geometry of the aircraft and model it as a rectangular solid with
+dimensions of wingspan ($W$), length ($L$) and height ($H$).
 
-Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations. 
-<!-- TERMINOLOGY? rectangular solid -- impact --> top plane, front plane? Does the literture use "view"  -->
-1.) if the aircraft were stationary, the area exposed to the debris impact would be the top view, and
-2.) if the debris were stationary, it would be the front view.
+Regarding the aircraft-debris encounter geometry, this yields to two
+conceptual considerations.
+<!-- TERMINOLOGY? rectangular solid -- impact --> top plane, front
+plane? Does the literature use "view" --\> 1.) if the aircraft were
+stationary, the area exposed to the debris impact would be the top view,
+and 2.) if the debris were stationary, it would be the front view.
 
-Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
-The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
-At the same time, the movement of the re-entered debris object at regular en-route flight levels will be vertically [^balistic_descent].
-This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as
+Our focus is on commercial aircraft in flight, and - implicitly -
+operating at higher altitudes. The en-route portion of a flight is
+characterised by the aircraft in straight and level flight, moving
+(mainly) horizontally. At the same time, the movement of the re-entered
+debris object at regular en-route flight levels will be vertically [^2].
+This encounter geometry allows to express the effective exposed area,
+$\varepsilon_k$, as
+
+[^2]: *Even for a ballistic coefficient of* \qty{50}{lbs/ft^2}, the
+    flight path angle of a reentering object at \qty{40000}{ft} is
+    \ang{-86.6}, which corresponds to nearly vertical descent. Since
+    most debris objects have ballistic coefficients equal to or less
+    than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
+    accurate approximation. [@patera2008]
 
@@
 
-where 
-* $s_{a}$ is the average aircraft speed, 
-* $s_{d_k}$ is the average fall speed of the $k$-th debris object, 
-* $a_T$ area of the aircraft as viewed from the top, and 
-* $a_F$ area of the aircraft as viewed from the from the front of the aircraft.
+where \* $s_{a}$ is the average aircraft speed, \* $s_{d_k}$ is the
+average fall speed of the $k$-th debris object, \* $a_T$ area of the
+aircraft as viewed from the top, and \* $a_F$ area of the aircraft as
+viewed from the from the front of the aircraft.
 
-This can be further instantiated for specific aircraft types.
-In the calculations the following is assumed to be true ($t$ subscript for aircraft type)
+This can be further instantiated for specific aircraft types. In the
+calculations the following is assumed to be true ($t$ subscript for
+aircraft type)
 
-* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$
-* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
-* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
-  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
-  type $t$
-* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
-  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
-  type $t$
+-   constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ [[page 8
+    @patera2008]\]
+-   constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed,
+-   aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$,
+    where $W_t$ is the wingspan, and $L_t$ the length for the specific
+    aircraft type $t$
+-   aircraft area as viewed from the front,
+    $a_F = W_t \times H_t = F_t$, where $W_t$ is the wingspan, and $H_t$
+    the height for the specific aircraft type $t$
 
-The effective exposed area now only depends on the aircraft type $t$ (i.e., its dimensions
-and cruise speed) and can be rewritten as:
+The effective exposed area now only depends on the aircraft type $t$
+(i.e., its dimensions and cruise speed) and can be rewritten as:
 
@@
 
-These aircraft properties are well-known. 
-While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
-The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].
+These aircraft properties are well-known. While there does not exist a
+single openly available date base for these aircraft parameters, we
+collected the information from a series of sources and integrated them
+into a lookup dataset. The values $W_t$, $H_t$ and $L_t$ have been
+retrieved from [@performance_database; @skybrary; @doc8643]; we used
+\qty{1000}{m^2} for $\varepsilon_t$ in case of missing aircraft type.
 
-The impact probability for cell $h$ under consideration of a specific aircraft of type $t$ becomes:
+The impact probability for cell $h$ under consideration of a specific
+aircraft of type $t$ becomes:
 
@@
 
-[^balistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
-the flight path angle of a reentering object at \qty{40000}{ft} is 
-\ang{-86.6}, which corresponds to nearly vertical descent.
-Since most debris objects have ballistic coefficients equal to or 
-less than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
-accurate approximation.* [@patera2008]
-
 ## Impact probability density for debris of inclination $i$
 
-Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
-circular orbits prior to re-entry even if initially on elliptical
-ones.[^circular-orbit]
+Debris that reenter the Earth's atmosphere via drag-induced decay will
+fly circular orbits prior to re-entry even if initially on elliptical
+ones.[^3]
 
-The angular position, $\theta$, of the re-entry point from the ascending node, 
-see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
-$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
-and the gravitational effects of the Equatorial bulge.
-Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.
+[^3]: In fact drag at perigee will lower apogee and similarly drag at
+    apogee will lower perigee. But because drag at perigee is higher due
+    to denser atmosphere, the apogee will decrease faster than perigee
+    and the orbit will then tend to be a circular one.
 
-The impact probability density, $\sigma(\varphi, i)$,
-at latitude $\varphi$ of a reentring object whose orbit has inclination $i$
-(see @fig-orbital-elements) is given by [[Equation 15 in @patera2008]]:
+The angular position, $\theta$, of the re-entry point from the ascending
+node, see @sec-orbitals, is uniformly distributed for uncontrolled
+reentries, $\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in
+atmospheric drag and the gravitational effects of the Equatorial bulge.
+Hence we assume that the impact longitude, $\lambda$, is uniformly
+distributed.
+
+The impact probability density, $\sigma(\varphi, i)$, at latitude
+$\varphi$ of a reentering object whose orbit has inclination $i$ (see
+@fig-orbital-elements) is given by [[Equation 15 in @patera2008]\]:
 
@@
 
-where $R$ is the distance from the center of the Earth of the object potentially
-being impacted by the debris.
+where $R$ is the distance from the centre of the Earth of the object
+potentially being impacted by the debris.
 
-So for example if a building is located at latitude 30 degrees and has an area 
-of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth’s atmosphere
-from an orbit having an inclination of 45 degrees, the impact probability is 
-given by
+So for example if a building is located at latitude 30 degrees and has
+an area of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth's
+atmosphere from an orbit having an inclination of 45 degrees, the impact
+probability is given by
 
@@
 
-where we used the equatorial value of $\qty{6378}{km}$, for the Earth's radius,
-$R$, as the distance of the building from the center of the planet.
+where we used the equatorial value of $\qty{6378}{km}$, for the Earth's
+radius, $R$, as the distance of the building from the centre of the
+planet.
 
-The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
-$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic
-U shape.
+The impact probability density $\sigma(\varphi, i)$ for an inclination
+$i$ of $\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes
+a characteristic U shape.
 
 ::: {#fig-weighting-inclination-i}
-
 ```{r weighting-inclination-i}
@@
 
-Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of inclination $i$ and altitude $\qty{550}{km}$. 
+Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of
+inclination $i$ and altitude $\qty{550}{km}$.
 :::
-
-[^circular-orbit]: In fact drag at perigee will lower apogee and similarly drag
-at apogee will lower perigee. But because drag at perigee is higher due to
-denser atmosphere, the apogee  will decrease faster than perigee and the orbit
-will then tend to be a circular one.
 
@@
 
-If we consider historical uncontrolled debris reentry data, we can numerically
-derive the debris impact probability as a weighting function by summing up and
-averaging all single densities:
+If we consider historical uncontrolled debris reentry data, we can
+numerically derive the debris impact probability as a weighting function
+by summing up and averaging all single densities:
 
@@
 
-From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
-1st Jan 2000 till 31st Dec 2024.
-The result of combining all the U-shaped densities can be seen in @fig-weighting.
+From [@gcat] we extracted 1968 debris re-entries during 25 calendar
+years from 1st Jan 2000 till 31st Dec 2024. The result of combining all
+the U-shaped densities can be seen in @fig-weighting.
 
 ::: {#fig-weighting}
-
 ```{r weighting}
@@
 
-Weighting function derived from 25 years of re-entries as an
-empirical approximation of $\sigma(\varphi)$. 
+Weighting function derived from 25 years of re-entries as an empirical
+approximation of $\sigma(\varphi)$.
 :::
@@
 ::: {#fig-weighting-nm-area}
-
 ```{r weighting-nm-area}
@@
 
-When combining the hourly ($H$) aircraft density (per aircraft type) with the
-relevant weightings per H3 cell, $h$, we can numerically compute the hourly
-impact probability for each hexagonal cell as:
+When combining the hourly ($H$) aircraft density (per aircraft type)
+with the relevant weightings per H3 cell, $h$, we can numerically
+compute the hourly impact probability for each hexagonal cell as:
 
@@
 
-where 
-* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
-* $W(h)$ is the median value of the weighting function over the area of hexagon
-$h$,
-* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
-cell $h$ at hour $H$, and
-* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.
+where \* $E_{h,H}$ is the collision expectation in the hexagon $h$ at
+hour $H$, \* $W(h)$ is the median value of the weighting function over
+the area of hexagon $h$, \* $\rho_{t,H}(h)$ is the traffic density for
+the \acr{ICAO} aircraft type $t$ in cell $h$ at hour $H$, and \*
+$\varepsilon_t$ is effective exposed area for aircraft of type $t$.
 
-@fig-collision depicts the determined hourly collision expectation.
-As identified in @fig-density-summer2024-res-3 areas of higher traffic density observe a higher collision expectation.
+@fig-collision depicts the determined hourly collision expectation. As
+identified in @fig-density-summer2024-res-3 areas of higher traffic
+density observe a higher collision expectation.
 <!-- WHAT ELSE CAN WE SAY ABOUT THIS PLOT? -->
@@
 ::: {#fig-collision}
-
 ```{r collision}
@@
 # Application Use Case: KOSMOS 482 Descent Craft re-entry
+
 <!-- I would make this a separate section in the paper -->
@@
 ($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
-Earth's atmosphere [@kosmos482re-entry].
-This is supposedly the lander module from a 1972 failed Soviet Venera mission [@SpaceReviewKosmos]
-The module is reported to weigh about $\qty{480}{kg}$ and has a diameter of approximately $\qty{1}{m}$.
-Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
-it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
+Earth's atmosphere [@kosmos482re-entry]. This is supposedly the lander
+module from a 1972 failed Soviet Venera mission [@SpaceReviewKosmos] The
+module is reported to weigh about $\qty{480}{kg}$ and has a diameter of
+approximately $\qty{1}{m}$. Being supposed to survive Venus's atmosphere
+which is much denser and hotter than Earth's atmosphere (surface
+pressure 93 times that of Earth and temperatures reaching \ang{467}C
+(\ang{872}F)), it could re-enter Earth intact and not break into smaller
+fragments some of which could burn up.
 
-Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of reentry.
+Knowing that the debris inclination is $\ang{51.7}$, see
+@fig-weighting-inclination-i, we can derive the maximum collision
+expectation using data for the busiest day in the busiest 2024 week as a
+proxy for the traffic density on the day of reentry.
 
-Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
-can be identified to retrieve the values of collision expectancy.
-It can then be decided whether it is advisable to close the airspace and
-traffic around the area affected.
+Once the reentry track is known, the H3 cells around \qty{35}{km} on
+each side can be identified to retrieve the values of collision
+expectancy. It can then be decided whether it is advisable to close the
+airspace and traffic around the area affected.
 
 ::: {#fig-kosmos}
-
 ```{r kosmos}
@@
 
-We have calculated collision and casuality risk probabilities for the two
-busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
-as logically depicted in @fig-data-densities is available in the companion
-GitHub repository of the paper at <https://github.com/espinielli/flight_density_space_debris>.
-We have also developed a companion R package `aviodebris` to heavylift the
-various data processing steps to produce the results in the paper;
-it is available at <https://github.com/espinielli/aviodebris>.
+We have calculated collision and casuality risk probabilities for the
+two busiest weeks in 2023 and 2024. The relevant dataset represents a
+data cube as logically depicted in @fig-data-densities is available in
+the companion GitHub repository of the paper at
+<https://github.com/espinielli/flight_density_space_debris>. We have
+also developed a companion R package `aviodebris` to heavy lift the
+various data processing steps to produce the results in the paper; it is
+available at <https://github.com/espinielli/aviodebris>.
 
 ::: {#fig-data-densities}
-
 ```{r densities}
@@
 
-Data cube of hourly flight densities data over H3 hexes (in Germany as an example) for multiple dates, $d_1 \cdots d_n$. 
+Data cube of hourly flight densities data over H3 hexes (in Germany as
+an example) for multiple dates, $d_1 \cdots d_n$.
 :::
@@
 
-<!-- opening statement linking back to abstract/intro and movtivation for this work --------->
-Space objects and debris re-entry is an increasing risk to aviation and the general public.
-The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
-Dependent on the mission, objects re-entering the atmosphere can take on differnt forms, shapes, and sizes.
-On top, a higher number of these entries may be uncontrolled.
-With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accomodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
+<!-- opening statement linking back to abstract/intro and motivation for this work --------->
+
+Space objects and debris re-entry is an increasing risk to aviation and
+the general public. The higher pace of commercial space operations will
+see an increase of launches and subsequent re-entries. Dependent on the
+mission, objects re-entering the atmosphere can take on different forms,
+shapes, and sizes. On top, a higher number of these entries may be
+uncontrolled. With the continual growth of air traffic, there is a need
+to enhance aviation safety risk awareness and respond in a more tailored
+fashion to accommodate the associated uncertainties, reduce the impact
+on aviation, and limit the risk to the travelling public.
 
 <!-- what was shown in the paper ------------------------------------------------------------>
-This paper expands on previous work.
-With a focus on Europe, we establish an upper bound for the traffic density based on actual flights.
+
+This paper expands on previous work. With a focus on Europe, we
+establish an upper bound for the traffic density based on actual
+flights.
 
@@
 
-The findings are applied to a use-case: the imminent re-entry of the lander module of an earlier Venus mission.
-The maximum collision expectation for the busiest day in 2024 is determined.
+The findings are applied to a use-case: the imminent re-entry of the
+lander module of an earlier Venus mission. The maximum collision
+expectation for the busiest day in 2024 is determined.
 
@@
 
-The approach presented in this paper builds on previous research.
-Our approach utilises H3-tesselation to reduce the spatial complexity.
-This cell indexing approach reduces the computation significantly.
+The approach presented in this paper builds on previous research. Our
+approach utilises H3-tessellation to reduce the spatial complexity. This
+cell indexing approach reduces the computation significantly.
 
+```         
     WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
+```
 
@@
 
-Future work will address the shortcomings. <!-- or can we say something about the next concrete steps to remedy some weaknesses -->
+Future work will address the shortcomings.
+<!-- or can we say something about the next concrete steps to remedy some weaknesses -->
 
 <!-- application of the results -------------------------------------------------------------->
-We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
-The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.
 
-As part of this research and to inform the decision-makers it is planned to explore the impact of megacostellations and flight growth forecast and what would happen in the mid- to long-term in varios scenarios of uncontrolled re-entries and the risk for European airspace.
+We show how National and/or International authorities could prepare for
+the more and more likely and frequent occurrence of uncontrolled
+re-entry of space objects that could pose a safety concern on civilian
+aviation. The techniques we implemented on realistic data and sound
+algorithms can contribute to the bag of tools in the hands of technical
+and political entities aiming at guaranteeing the utmost safety and
+minimal economic disruption to civil aviation in Europe.
+
+As part of this research and to inform the decision-makers it is planned
+to explore the impact of mega-constellations and flight growth forecast
+and what would happen in the mid- to long-term in various scenarios of
+uncontrolled re-entries and the risk for European airspace.
 
@@
 
-:::{#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
-
+::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
 ```{r ellipse-parameters}
 #| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$.
-(the semiparameter (a.k.a.
+(the semi-parameter (a.k.a.
 _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"
@@
 
-Definition of Keplerian elements.
-The orbital elements describing the trajectory of orbiting object are illustrated in association with an inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$.
-The $I$-axis, $\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$, is perpendicular to the $I$-axis in the equatorial plane n the sense of Earth's rotation with the $K$-axis pointing up.
-The semi-major axis, $a$, the eccentricity, $e$, the inclination, $i$, the right ascension of the ascending node, $\Omega$ , and the argument of perigee, $\omega$, constitute, together with the mean anomaly at the initial position, $M_0$, a set of six constants of motion that are preserved in the absence of perturbation.
-True anomaly, $\nu$, specifies the location of an orbiting object within the orbit.
-It is the angle between perigee and the orbiting object’s position vector measured in the direction of the orbiting object’s motion.
-Figures from [@sellers2003understanding], respectively Figure 5-2 and Figure 5-9.
-
+Definition of Keplerian elements. The orbital elements describing the
+trajectory of orbiting object are illustrated in association with an
+inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$. The $I$-axis,
+$\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$,
+is perpendicular to the $I$-axis in the equatorial plane n the sense of
+Earth's rotation with the $K$-axis pointing up. The semi-major axis,
+$a$, the eccentricity, $e$, the inclination, $i$, the right ascension of
+the ascending node, $\Omega$ , and the argument of perigee, $\omega$,
+constitute, together with the mean anomaly at the initial position,
+$M_0$, a set of six constants of motion that are preserved in the
+absence of perturbation. True anomaly, $\nu$, specifies the location of
+an orbiting object within the orbit. It is the angle between perigee and
+the orbiting object's position vector measured in the direction of the
+orbiting object's motion. Figures from [@sellers2003understanding],
+respectively Figure 5-2 and Figure 5-9.
 :::
@@
 
-The angular position, $\theta$, of the re-entry point from the ascending node
-in the orbit plane is $\theta = \omega + \nu$.
-It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
+The angular position, $\theta$, of the re-entry point from the ascending
+node in the orbit plane is $\theta = \omega + \nu$. It is related to the
+latitude, $\varphi$, via the Cartesian coordinate $k$,
 $k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or
@@
 These are used to derive numerically compute the weighting function,
-@fig-weighting-inclination-i:
-for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10001$,
-positions and calculate the corresponding cartesian position and the
-equivalent longitude and latitude.
+@fig-weighting-inclination-i: for debris with inclination $i$ we
+simulate an orbit of $N$, i.e. $10001$, positions and calculate the
+corresponding Cartesian position and the equivalent longitude and
+latitude.
 

### 2025-05-05 paper.qmd
most final for words...still need to fix pipeline

--- 2025-05-04 paper.qmd
@@
   - abstract-section
+
+crossref:
+  eq-title: Equation     # (default is "")
 
@@
 The quest for space is a driver for technological development.
-Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
+Key milestones include the first human travel to space and circumnavigating the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
 These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
-This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
+This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable production of goods in a non-gravity environment, services, and recently leisure.
 The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.
@@
 <!-- as a minimum, we should say something about the WHAT -->
-In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
+In general, the method for estimating the risk to commercial aircraft from re-entering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
 In that respect, a robust determination of flight density is a key aspect.
@@
 This work focusses on the European region.
-The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
+The hourly flight density estimation is based on flown \acr{IFR} trajectories as
 recorded by EUROCONTROL's \acr{NM}[^1].
@@
 
-H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
+H3 cells (red) at resolution 3 and their centres (blue) covering the bounding
 box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).
@@
 
-[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States.
+[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States, see blue polygon in @fig-cells.
 
 <!-- next sentence feels like an orphan ... do we need this here? ------>
-Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.
+Next sections will develop the mathematical background for the evaluation of collision expectation of which traffic density is a key component.
 
@@
 For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
-(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
+(Ukraine, Libya or Syria). The major traffic flows (North Atlantic, Canaries
 and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.
@@
 #| out-width: "110%"
+#| width: "9cm"
 #| echo: false
@@
 Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
+The black hexagon in the South of UK marks the densest cell at 24.56 flights/hour, just around the airport of London Heathrow.
 :::
@@
 
-Safety and collision risk awareness is directly linked to the probability of a re-entering space object colliding with a transversing flight.
+Safety and collision risk awareness is directly linked to the probability of a re-entering space object colliding with a traversing flight.
 This probability, $p_k$, of a debris (point) object $k$, impacting an
@@
 where
+
 * $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
@@
 
-Our approach builds on the discretisation of the geo-spatial risk area.
-For this purpose we use a hexagon-based tesselation pattern.
-Uber's H3 method ensures non-overlapping hexagons, with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
+Our approach builds on the discretisation of the geospatial risk area.
+For this purpose we use a hexagon-based tessellation pattern.
+Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
 Using H3 to discretise the area of interest $a$ gives
@@
 
-The impact probability $p_{k}(h)$ for each cell $h$ is determined by the following
+The impact probability $p_{k}(h)$ for each cell $h$ is determined by the following expression
 
@@
 p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h).
-$$
+$$ {#eq-impact-k-debris-cell}
 
@@
 This assumption holds for practically all commercial aircraft within the scope of this study, and we can neglect the debris impact area.
-Accordingly, we can also simplify the geometry of the aircraft and model it as a rectangular solid with dimesions of wingspan ($W$), length ($L$) and height ($H$).
+Accordingly, we can also simplify the geometry of the aircraft and model it as a rectangular solid with dimensions of wingspan ($W$), length ($L$) and height ($H$).
 
-Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations. 
-<!-- TERMINOLOGY? rectangular solid -- impact --> top plane, front plane? Does the literture use "view"  -->
-1.) if the aircraft were stationary, the area exposed to the debris impact would be the top view, and
-2.) if the debris were stationary, it would be the front view.
+Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
+<!-- TERMINOLOGY? rectangular solid -- impact top plane, front plane? Does the literature use "view"  -->
+
+1. if the aircraft were stationary, the area exposed to the debris impact would be the top face of the rectangular solid, and
+2. if the debris were stationary, it would be the front face.
 
@@
 The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
-At the same time, the movement of the re-entered debris object at regular en-route flight levels will be vertically [^balistic_descent].
+At the same time, the falling of the re-entering debris object at regular en-route flight levels will be vertical[^ballistic_descent].
 This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as
@@
 \varepsilon_k = \frac{s_{a} a_{h} + s_{d_k} a_{v}}{s_{d_k}}
-$$
+$$ {#eq-effective-exposed-area}
 
 where
+
 * $s_{a}$ is the average aircraft speed, 
@@
 * constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
-  [[pag 8 @patera2008]]
-* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
+  ([we adopt this value from pag 7-8 in @patera2008])
+* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, under the
+  simplification that the majority of the flight time spent cruising, 
 * aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
@@
 \varepsilon_t = \frac{C_{t} F_{t} + S A_{t}}{S}
-$$
+$$ {#eq-effective-exposed-area-simplified}
 
@@
 While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
-The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used \qty{1000}{m^2} for 
-$\varepsilon_t$ in case of missing aircraft type.
+The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries.
 
@@
 p_{t,k}(h) = \varepsilon_t \rho_t(h) \sigma_k(h)
-$$
+$$ {#eq-impact-k}
 
-[^balistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
-the flight path angle of a reentering object at \qty{40000}{ft} is 
+[^ballistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
+the flight path angle of a re-entering object at \qty{40000}{ft} is 
 \ang{-86.6}, which corresponds to nearly vertical descent.
@@
 
-Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
+Debris that re-enter the Earth’s atmosphere via drag-induced decay will fly
 circular orbits prior to re-entry even if initially on elliptical
@@
 
-The angular position, $\theta$, of the re-entry point from the ascending node, 
-see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
-$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
-and the gravitational effects of the Equatorial bulge.
-Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.
+The angular position, $\theta = \omega + \nu$, of the re-entry point from the
+ascending node, see @sec-orbitals, is uniformly distributed for uncontrolled
+re-entries, $\sigma(\theta) = 1 / (2\pi)$.
+This is due to uncertainties in atmospheric drag (among other things solar
+activity can expand the atmosphere) and the gravitational effects of the
+Equatorial bulge.
+Hence we assume that the impact longitude, $\lambda$, is uniformly distributed.
 
 The impact probability density, $\sigma(\varphi, i)$,
-at latitude $\varphi$ of a reentring object whose orbit has inclination $i$
-(see @fig-orbital-elements) is given by [[Equation 15 in @patera2008]]:
+at latitude $\varphi$ of a re-entering object whose orbit has inclination $i$
+(see @fig-orbital-elements) is given by ([Equation 15 in @patera2008]):
 
@@
 
-where $R$ is the distance from the center of the Earth of the object potentially
+where $R$ is the distance from the centre of the Earth of the object potentially
 being impacted by the debris.
@@
 So for example if a building is located at latitude 30 degrees and has an area 
-of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth’s atmosphere
+of $\qty{1000}{m^2}$ and a piece of debris re-enters the Earth’s atmosphere
 from an orbit having an inclination of 45 degrees, the impact probability is 
@@
 where we used the equatorial value of $\qty{6378}{km}$, for the Earth's radius,
-$R$, as the distance of the building from the center of the planet.
+$R$, as the distance of the building from the centre of the planet.
 
 The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
-$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic
-U shape.
+$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
+This density has been derived by simulating a full circular orbit at
+\qty{550}{km} height with 10000  positions and calculating the relevant longitudes and latitudes at half degree resolution.
 
@@
 at apogee will lower perigee. But because drag at perigee is higher due to
-denser atmosphere, the apogee  will decrease faster than perigee and the orbit
-will then tend to be a circular one.
+denser atmosphere (the perigee is closest to Earth), the apogee  will decrease faster than perigee and the orbit will then tend to become a circular one, see figure about apogee/perigee tracking in [@kosmos482reentry] for empirical evidence.
 
@@
 
-If we consider historical uncontrolled debris reentry data, we can numerically
+If we consider historical uncontrolled debris re-entry data, we can numerically
 derive the debris impact probability as a weighting function by summing up and
@@
 
-Spatially combining $\sigma(\lambda,\varphi)$ for the H3 cells in the
-\acr{NM} area we obtain the weightings per cell (per hour) as in
+Spatially projecting $\sigma(\lambda,\varphi)$ on the H3 cells in the
+\acr{NM} area we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
 @fig-weighting-nm-area.
@@
 
-Weightings for H3 cells in \acr{NM} area. 
+Hourly weightings for H3 cells, $W_H(h)$, in \acr{NM} area. 
 :::
+
+Then @eq-impact-k becomes
+
+$$
+p_t(h) = \varepsilon_t \rho_t(h) \sigma(h).
+$$ {#eq-impact-cell}
 
@@
 When combining the hourly ($H$) aircraft density (per aircraft type) with the
-relevant weightings per H3 cell, $h$, we can numerically compute the hourly
-impact probability for each hexagonal cell as:
+relevant weightings per $h$ cell, $W_H(h)$, we can numerically compute the
+hourly impact probability for each hexagonal cell as:
 
 $$
- E_{h, H} =  \sum_t W(h) \cdot \rho_{t,H}(h) \cdot \varepsilon_t
+ E_{h, H} =  \sum_t W_H(h) \cdot \rho_{t,H}(h) \cdot \varepsilon_t
 $$
@@
 where
+
 * $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
@@
 
-# Application Use Case: KOSMOS 482 Descent Craft re-entry
+# Use Case: KOSMOS 482 Descent Craft re-entry
 <!-- I would make this a separate section in the paper -->
 
-In few days (at the time of writing of this paper) at 6h May 10, 2025
-($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
-Earth's atmosphere [@kosmos482re-entry].
-This is supposedly the lander module from a 1972 failed Soviet Venera mission [@SpaceReviewKosmos]
-The module is reported to weigh about $\qty{480}{kg}$ and has a diameter of approximately $\qty{1}{m}$.
-Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
+In few days (at the time of writing of this paper) at 23:06 May 10, 2025
+($\pm 1.8$ days), the KOSMOS 482 Descent Craft should re-enter the
+Earth's atmosphere [@kosmos482reentry].
+Supposedly the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
+it is reported to weigh about $\qty{480}{kg}$ and to have a diameter of approximately $\qty{1}{m}$.
+Being built to survive Venus's atmosphere which is much denser and hotter than Earth's one (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
 it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
 
-Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of reentry.
+Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
+The densest cell, `83194afffffffff` (area \qty{11131585027}{m^2}), at 24.56 flights/hour has centroid latitude
+of \ang{50.94369} so within the latitude range of $[\ang{-51.7}, \ang{51.7}]$.
+Using an average value for effective exposed area of \qty{1000}{m^2},
+we can quickly calculate the collision expectation as
 
-Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
-can be identified to retrieve the values of collision expectancy.
-It can then be decided whether it is advisable to close the airspace and
+$$
+\begin{aligned}
+E &= w_h \cdot \sigma_h \cdot \bar{\varepsilon} / area_h \\
+  &= 10^{-6} \cdot 24.56 \cdot 1000 / 11131585027 = 2.2\cdot 10^{-12}
+\end{aligned}
+$$
+
+The risk is very low.
+
+Eventually once the re-entry track is known, the H3 cells around \qty{35}{km} on
+each side  can be identified to retrieve the values of collision expectancy.
+It can then be decided whether it is advisable to close the airspace and re-route
 traffic around the area affected.
-
-::: {#fig-kosmos}
-
-```{r kosmos}
-#| out-width: "100%"
-#| echo: false
-knitr::include_graphics("media/example-image.png")
-```
-
-Casuality for KOSMOS 482 Descent Craft re-entry. 
-:::
 
@@
 
-We have calculated collision and casuality risk probabilities for the two
+We have calculated collision and casualty risk probabilities for the two
 busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
 as logically depicted in @fig-data-densities is available in the companion
-GitHub repository of the paper at <https://github.com/espinielli/flight_density_space_debris>.
-We have also developed a companion R package `aviodebris` to heavylift the
+GitHub repository of the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
+We have also developed a companion R package `aviodebris` to heavy-lift the
 various data processing steps to produce the results in the paper;
-it is available at <https://github.com/espinielli/aviodebris>.
+it is available at <https://github.com/euctrl-pru/aviodebris>.
 
@@
 
-<!-- opening statement linking back to abstract/intro and movtivation for this work --------->
+<!-- opening statement linking back to abstract/intro and motivation for this work --------->
 Space objects and debris re-entry is an increasing risk to aviation and the general public.
 The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
-Dependent on the mission, objects re-entering the atmosphere can take on differnt forms, shapes, and sizes.
+Dependent on the mission, objects re-entering the atmosphere can take on different forms, shapes, and sizes.
 On top, a higher number of these entries may be uncontrolled.
-With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accomodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
+With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accommodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
 
@@
 The approach presented in this paper builds on previous research.
-Our approach utilises H3-tesselation to reduce the spatial complexity.
+Our approach utilises H3-tessellation to reduce the spatial complexity.
 This cell indexing approach reduces the computation significantly.
@@
 
-As part of this research and to inform the decision-makers it is planned to explore the impact of megacostellations and flight growth forecast and what would happen in the mid- to long-term in varios scenarios of uncontrolled re-entries and the risk for European airspace.
+As part of this research and to inform the decision-makers it is planned to explore the impact of megaconstellations and flight growth forecast and what would happen in the mid- to long-term in various scenarios of uncontrolled re-entries and the risk for European airspace.
 
@@
 
-$$\sin(\varphi) = \sin(\theta)\sin(i)$$ {#eq-spheric-coords}
+$$
+\sin(\varphi) = \sin(\theta)\sin(i)
+$$ {#eq-spheric-coords}
 
@@
 for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10001$,
-positions and calculate the corresponding cartesian position and the
+positions and calculate the corresponding Cartesian position and the
 equivalent longitude and latitude.

### 2025-05-05 R/pipeline.R

@@ -1,4 +1,4 @@
-# pipeline fot the paper
+# pipeline for the paper
 
 library(tidyverse)
 library(aviodebris)

### 2025-05-05 R/plot-density-map.R

@@ -101,7 +101,6 @@ ggplot() +
     aes(fill = occupancy),
     alpha = 0.8,
     key_glyph = draw_key_hex
-    # key_glyph = draw_key_cust
   ) +
   scale_fill_viridis_c(
     trans = "log",
@@ -117,7 +116,14 @@ ggplot() +
       nrow = 1
     )
   ) +
-  geom_sf(data = EUR_res20, fill = NA, linewidth = 0.65, colour = "white") +
+  # highlight MAX
+  geom_sf(
+    data = ddd |> mutate(MAX = max(occupancy)) |> filter(occupancy == MAX),
+    fill = "black",
+    alpha = 0.8,
+    key_glyph = draw_key_hex
+  ) +
+  geom_sf(data = EUR_res20, fill = NA, linewidth = 0.45, colour = "white") +
   # annotate() +
   theme_minimal() +
   coord_sf(crs = "ESRI:102013", datum = NA) +
@@ -150,8 +156,13 @@ ggplot() +
   ) +
   guides(fill = guide_legend(override.aes = list(size = 3.0)))
 
-ggsave(here(
-  "media",
-  "figures",
-  str_glue("flight_density_{date}_{resolution}.png")
-))
+ggsave(
+  # width = 80,
+  # units = "mm",
+  dpi = 600,
+  here(
+    "media",
+    "figures",
+    str_glue("flight_density_{date}_{resolution}.png")
+  )
+)

### 2025-05-05 paper.qmd

@@
   - abstract-section
-
 crossref:
@@
 <!-- zooming in on problem -->
-With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
-race to commercialise space, there is a need to re-evaluate the risks of
-collision and casualties due to uncontrolled debris re-entries.
-The increasing risk landscape poses challenges in terms of oversight and risk management.
-There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
-safety measures are the most adequate to minimise potential casualties and economic
-impact.
 
-The increase in space operations does not exist in a vacuum. 
-The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
-On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
-such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
-Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].
+With forecasts predicting a steadily growing number of flights, see
+@fig-forecast, and the modern race to commercialise space, there is a
+need to re-evaluate the risks of collision and casualties due to
+uncontrolled debris re-entries. The increasing risk landscape poses
+challenges in terms of oversight and risk management. There is a growing
+demand by overseeing authorities for new methodologies and tools to
+assess which safety measures are the most adequate to minimise potential
+casualties and economic impact.
+
+The increase in space operations does not exist in a vacuum. The
+continual growth of air traffic - despite temporal bumps - is evidenced
+by the surging number of flights in Europe [@forecastUpdate2025] and
+across the world. On-going development/deployment activities of
+mega-constellations of satellites in Low Earth Orbit \acr{LEO}, such as
+already active Starlink and OneWeb or planned ones (e.g. Kuiper,
+Starshield, Xingwang) will add thousand of orbiting objects with a short
+operational life. Some of these will inevitably re-enter the Earth's
+atmosphere in an uncontrolled manner, c.f. [@tbl-megaconstellations;
+@megaconstellations].
 
@@
 <!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
+
 The modelling and tracking of uncontrolled re-entries poses challenges.
-Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
-Common approaches employ a mix of handling the uncertainty of object or debris movement, the associated exposed area, including temporal uncertainty windows.
+Actual and historic data is sparse, and the size, shape, and attitude or
+rotation of re-entering objects increase the level of uncertainty for
+modelling. Common approaches employ a mix of handling the uncertainty of
+object or debris movement, the associated exposed area, including
+temporal uncertainty windows.
 
+```         
   <!-- is there more to say and provide context? -->
+```
 
 <!-- APPROACH / METHOD --------------------------------------------------- -->
+
 <!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
+
 <!-- as a minimum, we should say something about the WHAT -->
-In general, the method for estimating the risk to commercial aircraft from re-entering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
-In that respect, a robust determination of flight density is a key aspect.
+In general, the method for estimating the risk to commercial aircraft
+from reentering debris objects revolves around estimating the exposure
+of a single flight and the population of flights to the re-entering
+object. In that respect, a robust determination of flight density is a
+key aspect.
 
-We follow the same mathematical approach of [@patera2008; @wrightAirspaceClosuresDue2025].
-However, this paper augments the approximation of risk exposure. 
-Instead of approximating flight density with population density [@patera2008] or
-aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
-flight trajectories.
+We follow the same mathematical approach of [@patera2008;
+@wrightAirspaceClosuresDue2025]. However, this paper augments the
+approximation of risk exposure. Instead of approximating flight density
+with population density [@patera2008] or aircraft transponder data
+[@wrightAirspaceClosuresDue2025], we use actual flown flight
+trajectories.
 
-This work focusses on the European region.
-The hourly flight density estimation is based on flown \acr{IFR} trajectories as
-recorded by EUROCONTROL's \acr{NM}[^1].
-To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
-The reduction of the spatial complexity enables fast processing.
-For this work, we employ a H3-cell-resolution of 3. 
-This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
+This work focusses on the European region. The hourly flight density
+estimation is based on on flown \acr{IFR} trajectories as recorded by
+EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
+binned using Uber's H3 geospatial indexing system [@uber-h3]. The
+reduction of the spatial complexity enables fast processing. For this
+work, we employ a H3-cell-resolution of 3. This corresponds to hexagons
+of an average area $\qty{12393}{km^2}$ and an average edge length of
 $\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
@@
 
-H3 cells (red) at resolution 3 and their centres (blue) covering the bounding
-box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).
-
+H3 cells (red) at resolution 3 and their centres (blue) covering the
+bounding box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded
+polygon).
 :::
 
-To build an upper threshold for flight exposure, this study uses the busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.   <!-- does ISO do here anything? --> 
-The flight densities are characterised on an hourly basis for all days in these weeks.
+To build an upper threshold for flight exposure, this study uses the
+busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
+<!-- does ISO do here anything? --> The flight densities are
+characterised on an hourly basis for all days in these weeks.
 
@@
 <!-- next sentence feels like an orphan ... do we need this here? ------>
-Next sections will develop the mathematical background for the evaluation of collision expectation of which traffic density is a key component.
+Next sections will develop the mathematical background for the
+evaluation of collision and casualties risk of which traffic density is
+a key component.
 
-@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
-The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
-For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
-(Ukraine, Libya or Syria). The major traffic flows (North Atlantic, Canaries
-and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.
+@fig-density-summer2024-res-3 shows hourly flight density for the 5th of
+July 2024. The chart shows the expected outcomes in terms of major
+traffic flows, and areas of high or low traffic. For example, cells with
+a density of zero are clearly visible for areas of war or no-fly zones
+(Ukraine, Libya or Syria). The major traffic flows (North Atlantic,
+Canaries and East-bound over Türkiye) are clearly highlighted showing
+densities of more than 10.
 
 ::: {#fig-density-summer2024-res-3}
-
 ```{r density-winter-res-3}
@@
 
-## Aircraft effective exposed area{#sec-eea}
+## Aircraft effective exposed area {#sec-eea}
 
@@
 
-This can be further instantiated for specific aircraft types.
-In the calculations the following is assumed to be true ($t$ subscript for aircraft type)
+This can be further instantiated for specific aircraft types. In the
+calculations the following is assumed to be true ($t$ subscript for
+aircraft type)
 
@@
 
-The effective exposed area now only depends on the aircraft type $t$ (i.e., its dimensions
-and cruise speed) and can be rewritten as:
+The effective exposed area now only depends on the aircraft type $t$
+(i.e., its dimensions and cruise speed) and can be rewritten as:
 
@@
 
-The impact probability for cell $h$ under consideration of a specific aircraft of type $t$ becomes:
+The impact probability for cell $h$ under consideration of a specific
+aircraft of type $t$ becomes:
 
@@
 
-$$\begin{aligned}
+$$
+\begin{aligned}
 P &= \Delta A \times \sigma(\varphi, i) = 1000 \times \sigma\Bigl(\frac{\pi}{6}, \frac{\pi}{4}\Bigl) \\
   &= 1000 \times \num{2.490756e-15} = \num{2.490756e-12}
-\end{aligned}$$
+\end{aligned}
+$$
 
@@
 ::: {#fig-weighting-inclination-i}
-
 ```{r weighting-inclination-i}
@@
 
-Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of inclination $i$ and altitude $\qty{550}{km}$. 
+Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of
+inclination $i$ and altitude $\qty{550}{km}$.
 :::
@@
 ::: {#fig-weighting}
-
 ```{r weighting}
@@
 
-@fig-collision depicts the determined hourly collision expectation.
-As identified in @fig-density-summer2024-res-3 areas of higher traffic density observe a higher collision expectation.
+@fig-collision depicts the determined hourly collision expectation. As
+identified in @fig-density-summer2024-res-3 areas of higher traffic
+density observe a higher collision expectation.
 <!-- WHAT ELSE CAN WE SAY ABOUT THIS PLOT? -->
@@
 ::: {#fig-data-densities}
-
 ```{r densities}
@@
 <!-- what was shown in the paper ------------------------------------------------------------>
-This paper expands on previous work.
-With a focus on Europe, we establish an upper bound for the traffic density based on actual flights.
+
+This paper expands on previous work. With a focus on Europe, we
+establish an upper bound for the traffic density based on actual
+flights.
 
@@
 
+```         
     WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
+```
 
@@
 
-Future work will address the shortcomings. <!-- or can we say something about the next concrete steps to remedy some weaknesses -->
+Future work will address the shortcomings.
+<!-- or can we say something about the next concrete steps to remedy some weaknesses -->
 
 <!-- application of the results -------------------------------------------------------------->
-We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
-The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.
 

### 2025-05-05 paper.qmd
some tweaks to abstract/intro

@@
 
-The recent decades observed an increasing number of space activities
-resulting in higher numbers of re-entries at the end of the mission
+The recent decades observed an increase in the number of space activities.
+This resulted in higher numbers of re-entries at the end of the respective mission
 lifespan or orbital decay. The risk of collision between uncontrolled
-re-entering space debris and aircraft is increasing due to both
-ever-growing number of airborne flights and of orbiting bodies (and
-relevant re-entries). <!-- why important --> National and international
+re-entering space debris and aircraft is increasing due to the continual 
+growth of flights and of orbiting bodies and their relevant re-entries). 
+<!-- why important --> National and international
 authorities need to balance safety and economic impact of adopted
-measures should a re-entry event be announced, usually on short notice.
-<!-- no hard line breaks in abstracts -->
-<!-- approach / experiments --> We employ hourly and daily spatial
-flight densities calculated from trajectories as recorded by
+measures. This is particularly challenging should a re-entry event be announced on short notice.
+Thus, there is a need for effective tools to reduce the uncertainty and increase the safety awareness of uncontrolled re-entries. 
+<!-- approach / experiments --> To determine the risk exposure, this paper follows a data-driven approach.
+We employ hourly and daily spatial flight densities calculated from trajectories as recorded by
 EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
-indexing system.\
-Impact risk probabilities are then calculated based on previous
-theoretical work. <!-- results / conclusions --> This work enhances the
+indexing system. Impact risk probabilities are then calculated based on previous
+theoretical work. 
+<!-- results / conclusions --> This work enhances the
 model by using actual flown flight trajectories as a basis for
@@
 re-routings.
+The findings and the approach are demonstrated on the basis of an expected re-entry from an earlier space mission.\
+       IS IT WORTH TO STATE THE DETERMINED RISK?  \
+The approach in this paper enables a computing time efficient mechanism to support the re-entry related decision-making.
 
@@
 <!-- teaser opening --><!-- might need a rewrite - first jotting down -->
-The quest for space is a driver for technological development.
+The quest for space is a driver for technological development and exploitation.
 Key milestones include the first human travel to space and circumnavigating the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
-These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
-This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable production of goods in a non-gravity environment, services, and recently leisure.
+These developments helped to identify options for the wider use of space, associated technological or operational benefits, and its commercialisation.
+This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable production of goods in a non-gravity environment, space-asset based services, and recently even leisure.
 The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.  
+While ascent operations are typically well controlled and coordinated, a high number of re-entries ensue based on lifespan, decay, or side-effects, such as debris, failed components, etc.
 
@@
 uncontrolled debris re-entries. The increasing risk landscape poses
-challenges in terms of oversight and risk management. There is a growing
-demand by overseeing authorities for new methodologies and tools to
+challenges in terms of oversight and risk management. The operational and 
+economic demand by air transport requires a tradeoff with large-scale
+airspace and flow management measures to address the inherent safety risk.
+There is a growing demand by overseeing authorities for new methodologies and tools to
 assess which safety measures are the most adequate to minimise potential
@@
 across the world. On-going development/deployment activities of
-mega-constellations of satellites in Low Earth Orbit \acr{LEO}, such as
+mega-constellations of satellites in \acr{LEO}, such as the
 already active Starlink and OneWeb or planned ones (e.g. Kuiper,
-Starshield, Xingwang) will add thousand of orbiting objects with a short
+Starshield, Xingwang), will add thousand of orbiting objects with a short
 operational life. Some of these will inevitably re-enter the Earth's
@@
 @megaconstellations].
+These numbers and occurence rates will pose challenges to the steadily increasing air traffic.
 
@@
 space objects and their re-entry, the classical risk avoidance paradigm
-revolves around restricting the airspace. While the probability of a
-strike is low, it can result in catastrophic consequences. The growing
-number of space operations requires NEW - ADVANCED TOOLS. PREVIOUS WORK
-DID A,B, and MAYBE C - WE DO SOMETHING ON TOP.
+revolves around restricting the airspace. 
+Such airspace closures impact the flow of air traffic, as the operating schedule will get distorted due to delays or re-routings circumnavigating the exposed volume of airspace.
+While the probability of a strike is low, it can result in catastrophic consequences. 
+The growing number of space operations requires therefore tools to help improve the monitoring and trajectory prediction of uncontrolled re-entries, techniques to help reduce the uncertainty in terms of the dimension and duration of required airspace measures, including computationally efficient and fast methods to determine the risk exposure.
 

### 2025-05-05 paper.qmd
Update paper.qmd

@@
 
-The modelling and tracking of uncontrolled re-entries poses challenges.
+The modelling and tracking of uncontrolled re-entries poses challenges in terms of safety awareness and operational risk.
 Actual and historic data is sparse, and the size, shape, and attitude or
 rotation of re-entering objects increase the level of uncertainty for
-modelling. Common approaches employ a mix of handling the uncertainty of
-object or debris movement, the associated exposed area, including
+modelling. To support the decision-making, common approaches employ a mix of handling the uncertainty of
+object or debris movement and the associated exposed area for encounter, including
 temporal uncertainty windows.
-
+Thus, computationally efficient methods to assess and continously revalidate the collision risk of uncontrolled re-entries and aircraft will help to fine-tune the decision-making and geo-spactial and temporal constraints on aviation.
 ```         

### 2025-05-05 paper.qmd
Fixed some typos and made one sentence passive

@@
 challenges in terms of oversight and risk management. The operational and 
-economic demand by air transport requires a tradeoff with large-scale
+economic demand by air transport requires a trade-off with large-scale
 airspace and flow management measures to address the inherent safety risk.
@@
 @megaconstellations].
-These numbers and occurence rates will pose challenges to the steadily increasing air traffic.
+These numbers and occurrence rates will pose challenges to the steadily increasing air traffic.
 
@@
 temporal uncertainty windows.
-Thus, computationally efficient methods to assess and continously revalidate the collision risk of uncontrolled re-entries and aircraft will help to fine-tune the decision-making and geo-spactial and temporal constraints on aviation.
-```         
+Thus, computationally efficient methods to assess and continuously revalidate the collision risk of uncontrolled re-entries and aircraft will help to fine-tune the decision-making and geospatial and temporal constraints on aviation.
+
 <!-- is there more to say and provide context? -->
-```
 
@@
 
-If we consider historical uncontrolled debris re-entry data, we can numerically
+Considering historical uncontrolled debris re-entry data, we can numerically
 derive the debris impact probability as a weighting function by summing up and

### 2025-05-05 paper.qmd
brain-melting nodes and equator plane intersections

--- 2025-05-05 paper.qmd
@@
 In general, the method for estimating the risk to commercial aircraft
-from reentering debris objects revolves around estimating the exposure
+from re-entering debris objects revolves around estimating the exposure
 of a single flight and the population of flights to the re-entering
-object. In that respect, a robust determination of flight density is a
+object. In that respect, a robust determination of the flight density for encounter airspace volumes is a
 key aspect.
@@
 [@wrightAirspaceClosuresDue2025], we use actual flown flight
-trajectories.
+trajectories. This provides a more realistic figure of the number of affected air traffic.
+Thus, flight density is directly related to the number of flight operations per volume of airspace.
 
@@
 
-To build an upper threshold for flight exposure, this study uses the
+To build an upper threshold for the flight exposure, this study uses the
 busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
+This ensures a reasonable value, as peak airspace utilisation provides a handle on the air traffic operations and interactions of the different aircraft as part of wider *'traffic flows'*.
 <!-- does ISO do here anything? --> The flight densities are
-characterised on an hourly basis for all days in these weeks.
+characterised on an hourly basis for all days in these weeks per H3-cell.
 
@@
 <!-- next sentence feels like an orphan ... do we need this here? ------>
-Next sections will develop the mathematical background for the
-evaluation of collision and casualties risk of which traffic density is
+The next sections will develop the mathematical background for the
+evaluation of the collision and casualty risk for which air traffic density is
 a key component.
 
-@fig-density-summer2024-res-3 shows hourly flight density for the 5th of
-July 2024. The chart shows the expected outcomes in terms of major
+@fig-density-summer2024-res-3 shows the hourly flight density for the 5th of
+July 2024. The chart depicts the expected outcomes in terms of major
 traffic flows, and areas of high or low traffic. For example, cells with
 a density of zero are clearly visible for areas of war or no-fly zones
-(Ukraine, Libya or Syria). The major traffic flows (North Atlantic,
+(Ukraine, Libya or Syria). At the same time, the major traffic flows (North Atlantic,
 Canaries and East-bound over Türkiye) are clearly highlighted showing
@@
 Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
+While this paper focusses on the determination of the risk exposure per cell, the approach would allow to identify adjacent cells to account for the inherent uncertainty and risk appetite of decision-makers.
+
 Using H3 to discretise the area of interest $a$ gives
@@
 Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
-<!-- TERMINOLOGY? rectangular solid -- impact top plane, front plane? Does the literature use "view"  -->
 
@@
 
-Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
+Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes (i.e., flight levels).
 The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
-At the same time, the falling of the re-entering debris object at regular en-route flight levels will be vertical[^ballistic_descent].
+At the same time, the trajectory of a falling re-entering debris object at regular en-route flight levels will be a vertical decent movement[^ballistic_descent].
 This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as
@@
 
-These aircraft properties are well-known. 
+These aircraft properties are well-known design parameters. 
 While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
-The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries.
+The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries or unknown aircraft types.
 
@@
 ones.[^circular-orbit]
-
-The angular position, $\theta = \omega + \nu$, of the re-entry point from the
+A characteristic point for the determination of the re-entry position is the *'ascending node'*.
+This denotes the intersection point of the object crossing the plane of Earth's equator if moving from the southern hemnisphere to the nothern one.
+The respective angular position, $\theta = \omega + \nu$, of the re-entry point from the
 ascending node, see @sec-orbitals, is uniformly distributed for uncontrolled

### 2025-05-05 paper.qmd
a few edits/typos

@@
 A characteristic point for the determination of the re-entry position is the *'ascending node'*.
-This denotes the intersection point of the object crossing the plane of Earth's equator if moving from the southern hemnisphere to the nothern one.
+This denotes the intersection point of the object crossing the plane of Earth's equator if moving from the southern hemnisphere to the northern one.
 The respective angular position, $\theta = \omega + \nu$, of the re-entry point from the
@@
 re-entries, $\sigma(\theta) = 1 / (2\pi)$.
-This is due to uncertainties in atmospheric drag (among other things solar
-activity can expand the atmosphere) and the gravitational effects of the
+This is due to uncertainties in atmospheric drag and the gravitational effects of the
 Equatorial bulge.
-Hence we assume that the impact longitude, $\lambda$, is uniformly distributed.
+The height of the atmosphere is not a constant across the globe. Next to the gravitational effects, the atmosphere may also expand due to solar activity.
+Given these uncertainties, we assume that the impact longitude, $\lambda$, is uniformly distributed.
 
@@
 Spatially projecting $\sigma(\lambda,\varphi)$ on the H3 cells in the
-\acr{NM} area we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
+\acr{NM} area, we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
 @fig-weighting-nm-area.

### 2025-05-05 paper.qmd
W(h) -> W_H(h)

--- 2025-05-05 paper.qmd
@@
 In general, the method for estimating the risk to commercial aircraft
-from re-entering debris objects revolves around estimating the exposure
+from reentering debris objects revolves around estimating the exposure
 of a single flight and the population of flights to the re-entering
-object. In that respect, a robust determination of the flight density for encounter airspace volumes is a
+object. In that respect, a robust determination of flight density is a
 key aspect.
@@
 [@wrightAirspaceClosuresDue2025], we use actual flown flight
-trajectories. This provides a more realistic figure of the number of affected air traffic.
-Thus, flight density is directly related to the number of flight operations per volume of airspace.
+trajectories.
 
@@
 
-To build an upper threshold for the flight exposure, this study uses the
+To build an upper threshold for flight exposure, this study uses the
 busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
-This ensures a reasonable value, as peak airspace utilisation provides a handle on the air traffic operations and interactions of the different aircraft as part of wider *'traffic flows'*.
 <!-- does ISO do here anything? --> The flight densities are
-characterised on an hourly basis for all days in these weeks per H3-cell.
+characterised on an hourly basis for all days in these weeks.
 
@@
 <!-- next sentence feels like an orphan ... do we need this here? ------>
-The next sections will develop the mathematical background for the
-evaluation of the collision and casualty risk for which air traffic density is
+Next sections will develop the mathematical background for the
+evaluation of collision and casualties risk of which traffic density is
 a key component.
 
-@fig-density-summer2024-res-3 shows the hourly flight density for the 5th of
-July 2024. The chart depicts the expected outcomes in terms of major
+@fig-density-summer2024-res-3 shows hourly flight density for the 5th of
+July 2024. The chart shows the expected outcomes in terms of major
 traffic flows, and areas of high or low traffic. For example, cells with
 a density of zero are clearly visible for areas of war or no-fly zones
-(Ukraine, Libya or Syria). At the same time, the major traffic flows (North Atlantic,
+(Ukraine, Libya or Syria). The major traffic flows (North Atlantic,
 Canaries and East-bound over Türkiye) are clearly highlighted showing
@@
 Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
-While this paper focusses on the determination of the risk exposure per cell, the approach would allow to identify adjacent cells to account for the inherent uncertainty and risk appetite of decision-makers.
-
 Using H3 to discretise the area of interest $a$ gives
@@
 Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
+<!-- TERMINOLOGY? rectangular solid -- impact top plane, front plane? Does the literature use "view"  -->
 
@@
 
-Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes (i.e., flight levels).
+Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
 The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
-At the same time, the trajectory of a falling re-entering debris object at regular en-route flight levels will be a vertical decent movement[^ballistic_descent].
+At the same time, the falling of the re-entering debris object at regular en-route flight levels will be vertical[^ballistic_descent].
 This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as
@@
 
-These aircraft properties are well-known design parameters. 
+These aircraft properties are well-known. 
 While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
-The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries or unknown aircraft types.
+The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries.
 
@@
 ones.[^circular-orbit]
-A characteristic point for the determination of the re-entry position is the *'ascending node'*.
-This denotes the intersection point of the object crossing the plane of Earth's equator if moving from the southern hemnisphere to the nothern one.
-The respective angular position, $\theta = \omega + \nu$, of the re-entry point from the
+
+The angular position, $\theta = \omega + \nu$, of the re-entry point from the
 ascending node, see @sec-orbitals, is uniformly distributed for uncontrolled
@@
 * $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
-* $W(h)$ is the median value of the weighting function over the area of hexagon
+* $W_H(h)$ is the median value of the weighting function over the area of hexagon
 $h$,

### 2025-05-05 R/pipeline.R
new wording

@@ -47,8 +47,8 @@ traffic_density_3_30s <- purrr::partial(
 days |> purrr::walk(traffic_density_3_30s)
 
 # 4. compute casuality
-casuality_3_30s_hourly <- purrr::partial(
-  collision_and_casuality_risk_expectation_hourly,
+casualty_3_30s_hourly <- purrr::partial(
+  collision_and_casualty_risk_expectation_hourly,
   resolution = 3L
 )
-days |> purrr::walk(casuality_3_30s_hourly)
+days |> purrr::walk(casualty_3_30s_hourly)

### 2025-05-05 R/plot-collision-risk.R

@@ -26,7 +26,7 @@ eur_hex <- bbox_nm() |>
 
 ddd <- density |>
   group_by(cell) |>
-  summarise(collision_expectation = sum(collision_expectation) / 24) |>
+  summarise(collision_expectation = sum(collision_expectation)) |>
   ungroup() |>
   filter(cell %in% (eur_hex |> pull(h3_address))) |>
   cell_to_polygon(simple = FALSE)

### 2025-05-05 paper.qmd
merge

--- 2025-05-05 paper.qmd
@@
 $$
-\varepsilon_k = \frac{s_{a} a_{h} + s_{d_k} a_{v}}{s_{d_k}}
+\varepsilon_k = \frac{s_{a} a_{F} + s_{d_k} a_{T}}{s_{d_k}}
 $$ {#eq-effective-exposed-area}
@@
   simplification that the majority of the flight time spent cruising, 
-* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
+* aircraft area as viewed from the top, $a_T = W_t \times L_t = T_t$, 
   where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
@@
 $$
-\varepsilon_t = \frac{C_{t} F_{t} + S A_{t}}{S}
+\varepsilon_t = \frac{C_{t} F_{t} + S T_{t}}{S}
 $$ {#eq-effective-exposed-area-simplified}
@@
 While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
-The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries or unknown aircraft types.
+The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{500}{m^2}$
+and $pax = 7$ for missing entries which usually are due to smaller aircraft not listed in the above sources.
 
@@
 * $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
-* $W(h)$ is the median value of the weighting function over the area of hexagon
+* $W(h)$ is the mean value of the weighting function over the area of hexagon
 $h$,
@@
 Earth's atmosphere [@kosmos482reentry].
-Supposedly the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
+This is the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
 it is reported to weigh about $\qty{480}{kg}$ and to have a diameter of approximately $\qty{1}{m}$.
@@
 $a$, the eccentricity, $e$, the inclination, $i$, the right ascension of
-the ascending node, $\Omega$ , and the argument of perigee, $\omega$,
+the ascending node, $\Omega$, and the argument of perigee, $\omega$,
 constitute, together with the mean anomaly at the initial position,
@@
 The angular position, $\theta$, of the re-entry point from the ascending
-node in the orbit plane is $\theta = \omega + \nu$. It is related to the
+node[^ascending-node] in the orbit plane is $\theta = \omega + \nu$. It is related to the
 latitude, $\varphi$, via the Cartesian coordinate $k$,
@@
 
+[^ascending-node]: For an earth satellite, the ascending node is the point in its orbit where a satellite crosses the equatorial plane going from south to north.
+
 # References {.unnumbered}

### 2025-05-05 paper.qmd
almost there...

@@
 re-entering space debris and aircraft is increasing due to the continual 
-growth of flights and of orbiting bodies and their relevant re-entries). 
+growth of flights and of orbiting bodies and their relevant re-entries. 
 <!-- why important --> National and international
@@
 The next sections will develop the mathematical background for the
-evaluation of the collision and casualty risk for which air traffic density is
+evaluation of the collision risk for which air traffic density is
 a key component.
@@
 For this purpose we use a hexagon-based tessellation pattern.
-Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
+Uber's H3 method ensures non-overlapping hexagons (and 12 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
 While this paper focusses on the determination of the risk exposure per cell, the approach would allow to identify adjacent cells to account for the inherent uncertainty and risk appetite of decision-makers.
@@
 The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
-At the same time, the trajectory of a falling re-entering debris object at regular en-route flight levels will be a vertical decent movement[^ballistic_descent].
+At the same time, the trajectory of a falling re-entering debris object at regular en-route flight levels will be a vertical descent movement[^ballistic_descent].
 This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as
@@
 * constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, under the
-  simplification that the majority of the flight time spent cruising, 
+  simplification that the majority of the flight time is spent cruising, 
 * aircraft area as viewed from the top, $a_T = W_t \times L_t = T_t$, 
@@
 
-So for example if a building is located at latitude 30 degrees and has an area 
+So for example if a building is located at latitude \ang{30} and has an area 
 of $\qty{1000}{m^2}$ and a piece of debris re-enters the Earth’s atmosphere
-from an orbit having an inclination of 45 degrees, the impact probability is 
+from an orbit having an inclination of \ang{45}, the impact probability is 
 given by
@@
 
-$$ \sigma(\varphi) = \sum_k \sigma(\varphi, i_k)$$
+$$ \sigma(\varphi) = \frac{1}{N} \sum_{k = 1}^N \sigma(\varphi, i_k)$$
 
@@
 Spatially projecting $\sigma(\lambda,\varphi)$ on the H3 cells in the
-\acr{NM} area, we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
+\acr{NM} area, i.e. taking the mean of all the half degree weighting values
+and spreading in longitude,
+we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
 @fig-weighting-nm-area.
@@
 
+::: {#fig-kosmos-cell}
+```{r kosmos-cell}
+#| out-width: "45%"
+#| echo: false
+knitr::include_graphics(here::here("media", "figures", "cell_london.png"))
+```
+
+Cell with highest collision expectation due to KOSMO debris re-entry based on 2024-07-05 traffic.
+:::
+
 Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
-The densest cell, `83194afffffffff` (area \qty{11131585027}{m^2}), at 24.56 flights/hour has centroid latitude
-of \ang{50.94369} so within the latitude range of $[\ang{-51.7}, \ang{51.7}]$.
-Using an average value for effective exposed area of \qty{1000}{m^2},
-we can quickly calculate the collision expectation as
-
-$$
-\begin{aligned}
-E &= w_h \cdot \sigma_h \cdot \bar{\varepsilon} / area_h \\
-  &= 10^{-6} \cdot 24.56 \cdot 1000 / 11131585027 = 2.2\cdot 10^{-12}
-\end{aligned}
-$$
+Using the approach in this study and the weightings for this specific debris,
+the cell with the highest impact risk is cell, `83194afffffffff` (located South
+of London, cf. @fig-kosmos-cell), with a value of $1.89 \cdot  10^{-10}$.
 
@@
 
-We have calculated collision and casualty risk probabilities for the two
+We have calculated collision risk probabilities for the two
 busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
-as logically depicted in @fig-data-densities is available in the companion
-GitHub repository of the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
-We have also developed a companion R package `aviodebris` to heavy-lift the
-various data processing steps to produce the results in the paper;
-it is available at <https://github.com/euctrl-pru/aviodebris>.
+as logically depicted in @fig-data-densities.
 
@@
 ```{r densities}
-#| out-width: "100%"
+#| out-width: "80%"
 #| echo: false
@@
 
-# Conclusion and Future work
+Code and sample data is available in the companion GitHub repository of
+the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
+We have also developed an R package `aviodebris` to heavy-lift the
+various data processing steps to produce the results in the paper;
+it is available at <https://github.com/euctrl-pru/aviodebris>.
 
-<!-- opening statement linking back to abstract/intro and motivation for this work --------->
-Space objects and debris re-entry is an increasing risk to aviation and the general public.
-The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
-Dependent on the mission, objects re-entering the atmosphere can take on different forms, shapes, and sizes.
-On top, a higher number of these entries may be uncontrolled.
-With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accommodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
-
-<!-- what was shown in the paper ------------------------------------------------------------>
-
-This paper expands on previous work. With a focus on Europe, we
-establish an upper bound for the traffic density based on actual
-flights.
-
-... SOME MORE STUFF DONE IN THE PAPER ...
-
-The findings are applied to a use-case: the imminent re-entry of the
-lander module of an earlier Venus mission. The maximum collision
-expectation for the busiest day in 2024 is determined.
-
-<!-- benefits / strengths / weaknesses of the approach --------------------------------------->
-
-The approach presented in this paper builds on previous research.
-Our approach utilises H3-tessellation to reduce the spatial complexity.
-This cell indexing approach reduces the computation significantly.
-
-```         
-    WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
-```
-
-<!-- future work ---------------------------------------------------------------------------->
-
-Future work will address the shortcomings.
-<!-- or can we say something about the next concrete steps to remedy some weaknesses -->
-
-<!-- application of the results -------------------------------------------------------------->
-
-As part of this research and to inform the decision-makers it is planned to explore the impact of megaconstellations and flight growth forecast and what would happen in the mid- to long-term in various scenarios of uncontrolled re-entries and the risk for European airspace.
-
-# Appendix: Orbital Elements {#sec-orbitals}
-
-::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
+::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="htb"}
 ```{r ellipse-parameters}
@@
 
+# Conclusion and Future work
+
+<!-- opening statement linking back to abstract/intro and motivation for this work --------->
+Space objects and debris re-entry is an increasing risk to aviation and the general public.
+The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
+Dependent on the mission, objects re-entering the atmosphere can take on different forms, shapes, and sizes.
+On top, a higher number of these entries may be uncontrolled.
+With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accommodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
+
+<!-- what was shown in the paper ------------------------------------------------------------>
+
+This paper expands on previous work. With a focus on Europe, we
+establish an upper bound for the traffic density based on actual
+flights.
+In particular we derive impact risk probability for the European airspace
+using flown trajectories.
+
+The findings are applied to a use-case: the imminent re-entry of the
+lander module of an earlier Venus mission. The maximum collision
+expectation for the busiest day in 2024 is determined.
+
+<!-- benefits / strengths / weaknesses of the approach --------------------------------------->
+
+The approach presented in this paper builds on previous research.
+Our approach utilises H3-tessellation to reduce the spatial complexity.
+This cell indexing approach reduces the computation significantly.
+
+<!-- future work ---------------------------------------------------------------------------->
+
+Future work will address the derivation of casuality expectation using
+seats information for aircraft types in the traffic.
+<!-- or can we say something about the next concrete steps to remedy some weaknesses -->
+
+<!-- application of the results -------------------------------------------------------------->
+
+As part of this research and to inform the decision-makers it is planned to explore the impact of megaconstellations and flight growth forecast and what would happen in the mid- to long-term in various scenarios of uncontrolled re-entries and the risk for European airspace.
+
+# Appendix: Orbital Elements {#sec-orbitals}
+
 We refer to @fig-orbital-elements to describe the orbital parameters of

### 2025-05-05 R/plot-collision-risk-kosmos.R

@@ -1,0 +1,270 @@
+library(conflicted)
+library(tidyverse)
+library(aviodebris)
+library(arrow)
+library(sf)
+library(h3jsr)
+library(ggplot2)
+library(magrittr)
+library(giscoR)
+library(grid)
+library(here)
+library(latex2exp)
+
+conflicts_prefer(dplyr::filter)
+
+resolution <- 3
+date <- "2024-07-05" |> as_date()
+
+# longitudes at half degrees
+lons <- seq(0L, 719L) / 2
+hs <- 24
+
+# Earth equatorial radius
+rE <- 6378e3
+# altitude of satellite from (spherical) Earth surface, i.e. 500 km for LEO
+alt <- 550e3
+
+# inclination of Kosmos 482 Descent Craft (1972-023E, cat. nr. 6073)
+# see https://sattrackcam.blogspot.com/2025/04/kosmos-842-descent-craft-reentry.html
+inclination_deg = 51.7
+
+weightings_h3_resolution_3_hourly_kosmos <- latitude_weights(
+  delta_lat = 0.5,
+  altitude = rE + alt,
+  inclination_deg = inclination_deg
+) |>
+  mutate(lon = list(lons)) |>
+  unnest(lon) |>
+  mutate(
+    # spread by longitude
+    val = val / length(lons),
+    # spread by hour
+    val = val / hs,
+    NULL
+  ) |>
+  relocate(lon, .before = lat) |>
+  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
+  point_to_cell(res = resolution, simple = FALSE) |>
+  # integrate (sum) over the ones in each cell and divide by the cell area
+  # (take the mean of all the values at half degree)
+  summarize(
+    w = mean(val),
+    .by = h3_resolution_3
+  ) |>
+  as_tibble()
+
+
+collision <- here::here(
+  "data",
+  stringr::str_glue(
+    "traffic_density_{date}_res_{resolution}_hourly.parquet"
+  )
+) |>
+  arrow::read_parquet() |>
+  dplyr::filter(.data$h3_resolution == resolution) |>
+  dplyr::left_join(
+    weightings_h3_resolution_3_hourly_kosmos,
+    by = c("cell" = stringr::str_glue("h3_resolution_{resolution}"))
+  ) |>
+  # per cell per hour per aircraft type
+  dplyr::group_by(
+    .data$year,
+    .data$month,
+    .data$day,
+    .data$hour,
[…]

### 2025-05-05 R/plot-collision-risk.R

@@ -104,6 +104,7 @@ ggplot() +
   ) +
   scale_fill_viridis_c(
     transform = "log10",
+    # transform = "pseudo_log",
     breaks = seq(min, MAX, length.out = 4) |>
       format(digits = 3) |>
       as.numeric(),

### 2025-05-05 R/plot-weighting-latitude-kosmos.R

@@ -1,0 +1,169 @@
+library(tidyverse)
+library(aviodebris)
+
+
+resolution <- 3
+date <- "2024-07-05" |> as_date()
+
+# longitudes at half degrees
+lons <- seq(0L, 719L) / 2
+hs <- 24
+
+# Earth equatorial radius
+rE <- 6378e3
+# altitude of satellite from (spherical) Earth surface, i.e. 500 km for LEO
+alt <- 550e3
+
+# inclination of Kosmos 482 Descent Craft (1972-023E, cat. nr. 6073)
+# see https://sattrackcam.blogspot.com/2025/04/kosmos-842-descent-craft-reentry.html
+inclination_deg = 51.7
+
+weightings_h3_resolution_3_daily <- latitude_weights(
+  delta_lat = 0.5,
+  altitude = rE + alt,
+  inclination_deg = inclination_deg
+) |>
+  mutate(lon = list(lons)) |>
+  unnest(lon) |>
+  mutate(
+    # spread by longitude
+    val = val / length(lons),
+    # spead by hour
+    # val = val / hs,
+    NULL
+  ) |>
+  relocate(lon, .before = lat) |>
+  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
+  point_to_cell(res = resolution, simple = FALSE) |>
+  # integrate (sum) over the ones in each cell and divide by the cell area
+  # (take the mean of all the values at half degree)
+  reframe(
+    w = mean(val),
+    .by = h3_resolution_3
+  ) |>
+  as_tibble()
+
+# plot density
+library(sf)
+library(h3jsr)
+library(giscoR)
+library(scales)
+
+europe_hex <- function(res_h3, res_spatial = "20M") {
+  eur_hex_union <- aviodebris::bbox_nm() |>
+    aviodebris::bbox_of_hexes_for_bbox_at_res(resolution = res_h3)
+
+  gisco_get_countries(
+    resolution = "20",
+    epsg = "4326",
+    spatialtype = "RG",
+    region = c("Europe", "Africa", "Asia")
+  ) |>
+    mutate(res = res_spatial) |>
+    st_intersection(eur_hex_union) |>
+    select(res)
+}
+
+eur_hex <- bbox_nm() |> hexes_for_bbox_at_res(resolution = resolution)
+
+ddd <- weightings_h3_resolution_3_daily |>
[…]

### 2025-05-05 paper.qmd
final submission

@@
   - id: es
-    name: Enrico Spinielli
+    name: Author 1
     affiliations:
-      - ref: aiu
-    orcid: 0000-0001-8584-9131
-    email: enrico.spinielli@eurocontrol.int
-    attributes:
-      corresponding: true
-  - name: Rainer Koelle
+      - ref: fake
+    # name: Enrico Spinielli
+    # affiliations:
+    #   - ref: aiu
+    # orcid: 0000-0001-8584-9131
+    # email: enrico.spinielli@eurocontrol.int
+    # attributes:
+    #   corresponding: true
+  - name: Author 2
     affiliations:
-      - ref: aiu
-    orcid: 0000-0003-2810-1404
-    email: rainer.koelle@eurocontrol.int
-  - name: Quinten Goens
+      - ref: fake
+  # - name: Rainer Koelle
+  #   affiliations: 
+  #     - ref: aiu
+  #   orcid: 0000-0003-2810-1404
+  #   email: rainer.koelle@eurocontrol.int
+  - name: Author 3
     affiliations:
-      - ref: aiu
-    orcid: 0000-0003-2187-1924
-    email: quinten.goens@eurocontrol.int
+      - ref: fake
+  # - name: Quinten Goens
+  #   affiliations: 
+  #     - ref: aiu
+  #   orcid: 0000-0003-2187-1924
+  #   email: quinten.goens@eurocontrol.int
 affiliations:
@@
     country: Belgium
+  - id: fake
+    name: ACME
+    department: Best Tricks
+    city: Los Angeles
+    country: USA
 bibliography: media/references.bib
@@
 The recent decades observed an increase in the number of space activities.
-This resulted in higher numbers of re-entries at the end of the respective mission
-lifespan or orbital decay. The risk of collision between uncontrolled
+This resulted in higher numbers of re-entries at the end of the respective 
+mission lifespan or orbital decay. The risk of collision between uncontrolled
 re-entering space debris and aircraft is increasing due to the continual 
@@
 <!-- approach / experiments --> To determine the risk exposure, this paper follows a data-driven approach.
-We employ hourly and daily spatial flight densities calculated from trajectories as recorded by
-EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
-indexing system. Impact risk probabilities are then calculated based on previous
-theoretical work. 
-<!-- results / conclusions --> This work enhances the
-model by using actual flown flight trajectories as a basis for
-time-varying flight densities rather than relying on proxies such as the
-population density or ADS-B/Mode S message counts. These densities
-combined with a buffered area along the eventual path of re-entry, can
-provide mid-air impact probabilities to authorities as an aid for
-deciding on safety measures such as airspace closures and/or
+We employ hourly and daily spatial flight densities calculated from trajectories
+as recorded by EUROCONTROL's Network Manager and binned with the Uber H3
+geospatial indexing system.
+Impact risk probabilities are then calculated based on previous theoretical
+work. 
+<!-- results / conclusions --> This work enhances the previou model by using
+actual flown flight trajectories as a basis for time-varying flight densities
+rather than relying on proxies such as the population density or ADS-B/Mode S
+message counts. These densities combined with a buffered area along the eventual path of re-entry, can provide mid-air impact probabilities to authorities
+as an aid for deciding on safety measures such as airspace closures and/or
 re-routings.
 The findings and the approach are demonstrated on the basis of an expected re-entry from an earlier space mission.\
-       IS IT WORTH TO STATE THE DETERMINED RISK?  \
 The approach in this paper enables a computing time efficient mechanism to support the re-entry related decision-making.
@@
 airspace and flow management measures to address the inherent safety risk.
-There is a growing demand by overseeing authorities for new methodologies and tools to
-assess which safety measures are the most adequate to minimise potential
-casualties and economic impact.
+There is a growing demand by overseeing authorities for methodologies and
+tools to assess which safety measures are the most adequate to minimise
+potential casualties and economic impact.
 
@@
 already active Starlink and OneWeb or planned ones (e.g. Kuiper,
-Starshield, Xingwang), will add thousand of orbiting objects with a short
+Starshield, Xingwang), will add thousand of orbiting objects,
+c.f. [@tbl-megaconstellations;@megaconstellations] with a short
 operational life. Some of these will inevitably re-enter the Earth's
-atmosphere in an uncontrolled manner, c.f. [@tbl-megaconstellations;
-@megaconstellations].
+atmosphere in an uncontrolled manner.
 These numbers and occurrence rates will pose challenges to the steadily increasing air traffic.
@@
 revolves around restricting the airspace. 
-Such airspace closures impact the flow of air traffic, as the operating schedule will get distorted due to delays or re-routings circumnavigating the exposed volume of airspace.
-While the probability of a strike is low, it can result in catastrophic consequences. 
-The growing number of space operations requires therefore tools to help improve the monitoring and trajectory prediction of uncontrolled re-entries, techniques to help reduce the uncertainty in terms of the dimension and duration of required airspace measures, including computationally efficient and fast methods to determine the risk exposure.
+Such airspace closures impact the flow of air traffic, as the operating schedule
+will get distorted due to delays or re-routings circumnavigating the exposed volume of airspace.
+While the probability of a strike is low, it can result in catastrophic 
+consequences. 
+The growing number of space operations requires therefore tools to help improve
+the monitoring and impact of uncontrolled re-entries, techniques to help reduce
+the uncertainty in terms of the dimension and duration of required airspace
+measures, including computationally efficient and fast methods to determine the risk exposure.
 
@@
 
-The modelling and tracking of uncontrolled re-entries poses challenges in terms of safety awareness and operational risk.
+The modelling and tracking of uncontrolled re-entries poses challenges in terms
+of safety awareness and operational risk.
 Actual and historic data is sparse, and the size, shape, and attitude or
 rotation of re-entering objects increase the level of uncertainty for
-modelling. To support the decision-making, common approaches employ a mix of handling the uncertainty of
-object or debris movement and the associated exposed area for encounter, including
-temporal uncertainty windows.
-Thus, computationally efficient methods to assess and continuously revalidate the collision risk of uncontrolled re-entries and aircraft will help to fine-tune the decision-making and geospatial and temporal constraints on aviation.
+modelling. To support the decision-making, common approaches employ a mix of
+handling the uncertainty of
+object or debris movement and the associated exposed area for encounter,
+including temporal uncertainty windows.
+Thus, computationally efficient methods to assess and continuously revalidate
+the collision risk of uncontrolled re-entries and aircraft will help to
+fine-tune the decision-making and geospatial and temporal constraints on
+aviation.
 
@@
 ```{r cells-map}
-#| out-width: "120%"
+#| out-width: "110%"
 #| echo: false
@@
 ```{r density-winter-res-3}
-#| out-width: "110%"
+#| out-width: "105%"
 #| width: "9cm"
@@
 
-The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
-$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
-This density has been derived by simulating a full circular orbit at
-\qty{550}{km} height with 10000  positions and calculating the relevant longitudes and latitudes at half degree resolution.
-
 ::: {#fig-weighting-inclination-i}
 ```{r weighting-inclination-i}
-#| out-width: "100%"
+#| out-width: "90%"
 #| echo: false
@@
 :::
+
+The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
+$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
+This density has been derived by simulating a full circular orbit as detailed
+in @#sec-orbitals.
 
@@
 
-From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
-1st Jan 2000 till 31st Dec 2024.
-The result of combining all the U-shaped densities can be seen in @fig-weighting.
-
 ::: {#fig-weighting}
@@
 :::
+
+From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
+1st Jan 2000 till 31st Dec 2024.
+The result of combining all the U-shaped densities can be seen in @fig-weighting.
 
@@
 
-In few days (at the time of writing of this paper) at 23:06 May 10, 2025
-($\pm 1.8$ days), the KOSMOS 482 Descent Craft should re-enter the
+In few days (at the time of writing of this paper) at 10:05 May 10, 2025
+($\pm 1.5$ days), the KOSMOS 482 Descent Craft should re-enter the
 Earth's atmosphere [@kosmos482reentry].
@@
 
-::: {#fig-kosmos-cell}
-```{r kosmos-cell}
-#| out-width: "45%"
-#| echo: false
-knitr::include_graphics(here::here("media", "figures", "cell_london.png"))
-```
-
-Cell with highest collision expectation due to KOSMO debris re-entry based on 2024-07-05 traffic.
-:::
-
 Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
 Using the approach in this study and the weightings for this specific debris,
-the cell with the highest impact risk is cell, `83194afffffffff` (located South
-of London, cf. @fig-kosmos-cell), with a value of $1.89 \cdot  10^{-10}$.
+the cell with the highest impact risk is cell, `83194afffffffff` (located South of London), with a value of $1.89 \cdot  10^{-10}$.
 
@@
 as logically depicted in @fig-data-densities.
+The processing of data for 1 day takes less than 30 min from
+data extraction to impact probability calculation on a
+high-end laptop (MacBook Pro M1 Max).
 
@@
 
-Code and sample data is available in the companion GitHub repository of
+Code and sample data are available in the companion GitHub repository of
 the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
 We have also developed an R package `aviodebris` to heavy-lift the
-various data processing steps to produce the results in the paper;
-it is available at <https://github.com/euctrl-pru/aviodebris>.
+various data processing steps which  is available at
+<https://github.com/euctrl-pru/aviodebris>.
 
-::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="htb"}
+::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
 ```{r ellipse-parameters}
@@
 The findings are applied to a use-case: the imminent re-entry of the
-lander module of an earlier Venus mission. The maximum collision
+lander module of an earlier failed Venus mission. The maximum collision
 expectation for the busiest day in 2024 is determined.
@@
 @fig-weighting-inclination-i:
-for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10001$,
+for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10000$,
 positions and calculate the corresponding Cartesian position and the

### 2025-05-10 paper.qmd
final version after extension

@@
 
-We follow the same mathematical approach of [@patera2008;
-@wrightAirspaceClosuresDue2025]. However, this paper augments the
-approximation of risk exposure. Instead of approximating flight density
-with population density [@patera2008] or aircraft transponder data
-[@wrightAirspaceClosuresDue2025], we use actual flown flight
-trajectories. This provides a more realistic figure of the number of affected air traffic.
-Thus, flight density is directly related to the number of flight operations per volume of airspace.
-
-This work focusses on the European region. The hourly flight density
-estimation is based on on flown \acr{IFR} trajectories as recorded by
-EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
-binned using Uber's H3 geospatial indexing system [@uber-h3]. The
-reduction of the spatial complexity enables fast processing. For this
-work, we employ a H3-cell-resolution of 3. This corresponds to hexagons
-of an average area $\qty{12393}{km^2}$ and an average edge length of
-$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
-
-<!-- CAN WE MOTIVATE WHY resolution 3??? -- e.g. is this an area used in other works, or does it do something? -->
-
 ::: {#fig-cells}
@@
 
-To build an upper threshold for the flight exposure, this study uses the
-busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
-This ensures a reasonable value, as peak airspace utilisation provides a handle on the air traffic operations and interactions of the different aircraft as part of wider *'traffic flows'*.
-<!-- does ISO do here anything? --> The flight densities are
-characterised on an hourly basis for all days in these weeks per H3-cell.
+We follow the same mathematical approach of [@patera2008;
+@wrightAirspaceClosuresDue2025]. However, this paper augments the
+approximation of risk exposure. Instead of approximating flight density
+with population density [@patera2008] or aircraft transponder data
+[@wrightAirspaceClosuresDue2025], we use actual flown flight
+trajectories. This provides a more realistic figure of the number of affected air traffic.
+Thus, flight density is directly related to the number of flight operations per volume of airspace.
 
-<!-- next sentence feels like an orphan ... do we need this here? ------>
-This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.
+This work focusses on the European region. The hourly flight density
+estimation is based on flown \acr{IFR} trajectories as recorded by
+EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
+binned using Uber's H3 geospatial indexing system [@uber-h3]. The
+reduction of the spatial complexity enables fast processing. For this
+work, we employ a H3-cell-resolution of 3. This corresponds to hexagons
+of an average area $\qty{12393}{km^2}$ and an average edge length of
+$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.
+
+The choice of hex resolution is a compromise between having to use
+highly detailed trajectories (more 4D positions and more processing and
+storage) and capturing long enough portions of flight inside the cell.
+
+To build an upper threshold for the flight impact exposure, this study uses the
+busiest week during summer 2023 and 2024.
+This ensures a reasonable value, as peak airspace utilisation provides
+a handle on the air traffic operations and interactions of the different
+aircraft as part of wider *'traffic flows'*.
+The flight densities are characterised on an hourly basis for all days
+in these weeks per H3-cell.
 
 [^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States, see blue polygon in @fig-cells.
-
-<!-- next sentence feels like an orphan ... do we need this here? ------>
-The next sections will develop the mathematical background for the
-evaluation of the collision risk for which air traffic density is
-a key component.
-
-@fig-density-summer2024-res-3 shows the hourly flight density for the 5th of
-July 2024. The chart depicts the expected outcomes in terms of major
-traffic flows, and areas of high or low traffic. For example, cells with
-a density of zero are clearly visible for areas of war or no-fly zones
-(Ukraine, Libya or Syria). At the same time, the major traffic flows (North Atlantic,
-Canaries and East-bound over Türkiye) are clearly highlighted showing
-densities of more than 10.
 
@@
 
-Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
+Flight density in the EUROCONTROL Area on the 5th of July 2024 (the busiest day of the busiest summer week in 2024) at H3 cells resolution $3$.
 The black hexagon in the South of UK marks the densest cell at 24.56 flights/hour, just around the airport of London Heathrow.
 :::
+
+@fig-density-summer2024-res-3 shows an example of binning the hourly flight
+density on the 5th of July 2024 using H3 hexagons.
+The chart depicts the expected outcomes in terms of major
+traffic flows, and areas of high or low traffic. For example, cells with
+a density of zero are clearly visible for areas of war or no-fly zones
+(Ukraine, Libya or Syria). At the same time, the major traffic flows (North Atlantic,
+Canaries and East-bound over Türkiye) are clearly highlighted showing
+densities of more than 10.
+
+The next sections will develop the mathematical background for the
+evaluation of the collision risk for which air traffic density is
+a key component.
 
@@
 Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
-
 1. if the aircraft were stationary, the area exposed to the debris impact would be the top face of the rectangular solid, and
@@
 
-where
-
-* $s_{a}$ is the average aircraft speed, 
-* $s_{d_k}$ is the average fall speed of the $k$-th debris object, 
-* $a_T$ area of the aircraft as viewed from the top, and 
-* $a_F$ area of the aircraft as viewed from the from the front of the aircraft.
+where $s_{a}$ is the average aircraft speed, 
+$s_{d_k}$ is the average fall speed of the $k$-th debris object, 
+$a_T$ area of the aircraft as viewed from the top, and 
+$a_F$ area of the aircraft as viewed from the from the front of the aircraft.
 
@@
 
+::: {#fig-weighting-inclination-i}
+```{r weighting-inclination-i}
+#| out-width: "90%"
+#| echo: false
+knitr::include_graphics("media/figures/weighting_inclination_i.png")
+```
+
+Impact probability density $\sigma(\varphi, i = 51.95)$ for debris of
+inclination $i$ and altitude $\qty{550}{km}$.
+:::
+
 ## Impact probability density for debris of inclination $i$
@@
 
+::: {#fig-weighting}
+```{r weighting}
+#| out-width: "100%"
+#| echo: false
+#| fig-pos: H
+knitr::include_graphics("media/figures/weighting.png")
+```
+
+Weighting function derived from 25 years of re-entries as an empirical
+approximation of $\sigma(\varphi)$.
+:::
+
 So for example if a building is located at latitude \ang{30} and has an area 
@@
 
-::: {#fig-weighting-inclination-i}
-```{r weighting-inclination-i}
-#| out-width: "90%"
-#| echo: false
-knitr::include_graphics("media/figures/weighting_inclination_i.png")
-```
-
-Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of
-inclination $i$ and altitude $\qty{550}{km}$.
-:::
-
 The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
-$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
+$\ang{51.95}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
 This density has been derived by simulating a full circular orbit as detailed
-in @#sec-orbitals.
+in @sec-orbitals.
 
@@
 
-::: {#fig-weighting}
-```{r weighting}
-#| out-width: "100%"
-#| echo: false
-knitr::include_graphics("media/figures/weighting.png")
-```
-
-Weighting function derived from 25 years of re-entries as an empirical
-approximation of $\sigma(\varphi)$.
-:::
-
 From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
@@
 
-::: {#fig-weighting-nm-area}
-```{r weighting-nm-area}
-#| out-width: "100%"
-#| echo: false
-knitr::include_graphics("media/figures/weighting_on_Earth.png")
-```
-
-Hourly weightings for H3 cells, $W_H(h)$, in \acr{NM} area. 
-:::
+![Hourly weightings for H3 cells, $W_H(h)$, in \acr{NM} area.](media/figures/weighting_on_Earth.png){#fig-weighting-nm-area out-width=100%}
 
@@
 
+## Traffic density per cell
+
+In order to derive $\rho_t(h)$, we have taken the trajectories of flown flights
+in the \acr{NM} Area, resampled at $\qty{30}{s}$ so that the calculation of the
+occupancy, i.e. the time spent in the hex, per aircraft type is just a matter
+of counting the number of points per cell.
+This binning operation is extremely fast to perform in Uber's H3 gridding system
+and a real advantage in processing the huge amount of data point representing
+the flight trajectories.
+
+For July, $5^{th}$ 2024 alone, we extracted $37721$ flights with
+$33,482,686$ 4D positions (at $\qty{30}{s}$ sampling), where 19 aircraft types
+cover each at least 1% of the total and the remaining 266 (!) types account
+for the residual 20.8%, see @fig-aircraft-types.
+
 ## Hourly impact probability per H3 cell
 
-When combining the hourly ($H$) aircraft density (per aircraft type) with the
-relevant weightings per $h$ cell, $W_H(h)$, we can numerically compute the
+When combining the hourly ($H$) aircraft occupancy (per aircraft type) with the
+relevant weightings per $h$ cell, $W_H(h)$, and the effective exposed
+area, $\varepsilon_t$, we can numerically compute the
 hourly impact probability for each hexagonal cell as:
@@
 
-where
+where $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$.
 
-* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
-* $W(h)$ is the mean value of the weighting function over the area of hexagon
-$h$,
-* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
-cell $h$ at hour $H$, and
-* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.
+::: {#fig-aircraft-types}
+```{r aircraft-types-july-2024}
+#| out-width: "100%"
+#| echo: false
+knitr::include_graphics("media/figures/aircraft_types.png")
+```
+
+Aircraft types on 2024-07-05 in \acr{NM} Area. 
+:::
 
@@
 # Use Case: KOSMOS 482 Descent Craft re-entry
-<!-- I would make this a separate section in the paper -->
 
-In few days (at the time of writing of this paper) at 10:05 May 10, 2025
-($\pm 1.5$ days), the KOSMOS 482 Descent Craft should re-enter the
-Earth's atmosphere [@kosmos482reentry].
-This is the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
-it is reported to weigh about $\qty{480}{kg}$ and to have a diameter of approximately $\qty{1}{m}$.
-Being built to survive Venus's atmosphere which is much denser and hotter than Earth's one (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
-it could re-enter Earth intact and not break into smaller fragments some of which could burn up.
+While this paper was being written, the KOSMOS 482 Descent Craft was being
+monitored for an uncontrolled re-entry, forecast at 06:39 May 10, 2025
+($\pm 1.5$ hours) [@kosmos482reentry].
 
-Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
+![Venera-4 lander mock-up at the Memorial Museum of Cosmonautics. The Kosmos 482 Descent Craft is probably similar.](media/venera-capsule-in-museum.jpg){#fig-kosmos width=80%}
+
+This is the lander module from a 1972 failed USSR Venera mission
+[@SpaceReviewKosmos], which is reported to weigh about $\qty{480}{kg}$ and
+to have a diameter of approximately $\qty{1}{m}$.
+Being built to survive Venus's atmosphere which is much denser and hotter than
+Earth's one (surface pressure 93 times that of Earth and temperatures reaching
+\ang{467}C (\ang{872}F)), it could re-enter Earth intact and not break into
+smaller fragments some of which could burn up.
+
+Knowing that the debris inclination is $\ang{51.95}$, 
+see @fig-weighting-inclination-i, we can derive the maximum
+collision expectation using data for the busiest day in the busiest 2024 week
+as a proxy for the traffic density on the day of re-entry.
 Using the approach in this study and the weightings for this specific debris,
-the cell with the highest impact risk is cell, `83194afffffffff` (located South of London), with a value of $1.89 \cdot  10^{-10}$.
+the cell with the highest impact risk is cell, `83194afffffffff` 
+(located South of London), with a value of $1.89 \cdot  10^{-10}$.
 
@@
 
-Eventually once the re-entry track is known, the H3 cells around \qty{35}{km} on
-each side  can be identified to retrieve the values of collision expectancy.
-It can then be decided whether it is advisable to close the airspace and re-route
-traffic around the area affected.
+Radar systems over Germany saw the descent craft at approximately 04:30 UTC and
+06:04 UTC on May $10^{th}$. Given that the descent craft was not spotted at
+the next expected pass at 07:32 UTC, it is likely that the re-entry already occurred before.
 
-# Data and Code for Reproducibility
-
-We have calculated collision risk probabilities for the two
-busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
-as logically depicted in @fig-data-densities.
-The processing of data for 1 day takes less than 30 min from
-data extraction to impact probability calculation on a
-high-end laptop (MacBook Pro M1 Max).
+In general case when the re-entry track for a debris would be known
+with some lead time, the H3 cells around, say,  \qty{35}{km} on each side
+could be identified to retrieve the values of collision expectancy.
+It would then be possible to decide whether it would advisable to close
+the airspace and re-route traffic around the affected area.
 
@@
 ```{r densities}
-#| out-width: "80%"
+#| out-width: "70%"
 #| echo: false
+#| fig-pos: H
 knitr::include_graphics("media/figures/cubes.png")
@@
 
+# Data and Code for Reproducibility
+
+We have calculated collision risk probabilities for the two
+busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
+as logically depicted in @fig-data-densities.
+
+The processing of data for 1 day takes less than 30 min from
+data extraction to impact probability calculation on a
+high-end laptop (MacBook Pro M1 Max).
+
 Code and sample data are available in the companion GitHub repository of
@@
 
-::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
+# Conclusion and Future work
+
+<!-- opening statement linking back to abstract/intro and motivation for this work --------->
+Space objects and debris re-entry is an increasing risk to aviation and the
+general public.
+The higher pace of commercial space operations will see an increase of launches 
+and subsequent re-entries.
+Dependent on the mission, objects re-entering the atmosphere can take on
+different forms, shapes, and sizes.
+On top, a higher number of these entries may be uncontrolled.
+With the continual growth of air traffic, there is a need to enhance aviation
+safety risk awareness and respond in a more tailored fashion to accommodate
+the associated uncertainties, reduce the impact on aviation, and limit the risk
+to the travelling public.
+
+<!-- what was shown in the paper ------------------------------------------------------------>
+
+This paper expands on previous work. With a focus on Europe, we establish an
+upper bound for the traffic density based on actual flights.
+In particular we derive impact risk probability for the European airspace
+using flown trajectories.
+
+The findings are applied to a use-case: the imminent re-entry of the
+lander module of an earlier failed Venus mission. The maximum collision
+expectation is determined using the busiest day in 2024 as a representative
+proxy of the traffic on the day of re-entry.
+
+<!-- benefits / strengths / weaknesses of the approach --------------------------------------->
+
+The approach presented in this paper builds on previous research.
+Our approach utilises H3-tessellation to reduce the spatial complexity.
+This cell indexing approach reduces the computation burden significantly.
+
+<!-- future work ---------------------------------------------------------------------------->
+
+Future work will address the derivation of casuality expectation using
+seats information for aircraft types in the traffic.
+<!-- or can we say something about the next concrete steps to remedy some weaknesses -->
+
+<!-- application of the results -------------------------------------------------------------->
+
+As part of this research and to inform the decision-makers it is planned
+to explore the impact of megaconstellations and flight growth forecast and
+what would happen in the mid- to long-term in various scenarios of
+uncontrolled re-entries and the risk for European airspace.
+
+::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h" height=80%}
 ```{r ellipse-parameters}
@@
 
-Definition of Keplerian elements. The orbital elements describing the
-trajectory of orbiting object are illustrated in association with an
-inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$. The $I$-axis,
-$\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$,
-is perpendicular to the $I$-axis in the equatorial plane n the sense of
-Earth's rotation with the $K$-axis pointing up. The semi-major axis,
-$a$, the eccentricity, $e$, the inclination, $i$, the right ascension of
-the ascending node, $\Omega$, and the argument of perigee, $\omega$,
-constitute, together with the mean anomaly at the initial position,
+Orbital elements, Figures 5-2 and 5-9 from [@sellers2003understanding].
+:::
+
+# Appendix: Orbital Elements {#sec-orbitals .appendix}
+
+We refer to @fig-orbital-elements to describe the orbital parameters of
+interest for the current work.
+
+The orbital elements describing the trajectory of orbiting object are
+illustrated in association with an inertial reference frame
+$\hat{I}$-$\hat{J}$-$\hat{K}$. 
+The $I$-axis, $\hat{I}$, points toward the vernal equinox and 
+the $J$-axis, $\hat{J}$, is perpendicular to the $I$-axis in the equatorial plane
+in the sense of Earth's rotation, with the $K$-axis pointing up.
+The semi-major axis, $a$, the eccentricity, $e$, the inclination, $i$,
+the right ascension of the ascending node, $\Omega$, and the argument of perigee, $\omega$, constitute, together with the mean anomaly at the initial position,
 $M_0$, a set of six constants of motion that are preserved in the
@@
 the orbiting object's position vector measured in the direction of the
-orbiting object's motion. Figures from [@sellers2003understanding],
-respectively Figure 5-2 and Figure 5-9.
-:::
+orbiting object's motion.
 
-# Conclusion and Future work
-
-<!-- opening statement linking back to abstract/intro and motivation for this work --------->
-Space objects and debris re-entry is an increasing risk to aviation and the general public.
-The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
-Dependent on the mission, objects re-entering the atmosphere can take on different forms, shapes, and sizes.
-On top, a higher number of these entries may be uncontrolled.
-With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accommodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.
-
-<!-- what was shown in the paper ------------------------------------------------------------>
-
-This paper expands on previous work. With a focus on Europe, we
-establish an upper bound for the traffic density based on actual
-flights.
-In particular we derive impact risk probability for the European airspace
-using flown trajectories.
-
-The findings are applied to a use-case: the imminent re-entry of the
-lander module of an earlier failed Venus mission. The maximum collision
-expectation for the busiest day in 2024 is determined.
-
-<!-- benefits / strengths / weaknesses of the approach --------------------------------------->
-
-The approach presented in this paper builds on previous research.
-Our approach utilises H3-tessellation to reduce the spatial complexity.
-This cell indexing approach reduces the computation significantly.
-
-<!-- future work ---------------------------------------------------------------------------->
-
-Future work will address the derivation of casuality expectation using
-seats information for aircraft types in the traffic.
-<!-- or can we say something about the next concrete steps to remedy some weaknesses -->
-
-<!-- application of the results -------------------------------------------------------------->
-
-As part of this research and to inform the decision-makers it is planned to explore the impact of megaconstellations and flight growth forecast and what would happen in the mid- to long-term in various scenarios of uncontrolled re-entries and the risk for European airspace.
-
-# Appendix: Orbital Elements {#sec-orbitals}
-
-We refer to @fig-orbital-elements to describe the orbital parameters of
-interest for the current work.
-
-The angular position, $\theta$, of the re-entry point from the ascending
-node[^ascending-node] in the orbit plane is $\theta = \omega + \nu$. It is related to the
-latitude, $\varphi$, via the Cartesian coordinate $k$,
+The angular position, $\theta$, of the re-entry debris taken from the ascending
+node[^ascending-node] in the orbit plane is $\theta = \omega + \nu$.
+It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
 $k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or
@@
 
-Mass of Earth, $M_{\oplus} = \qty{5.9722e24}{kg}$.
+In the derivation of debris orbital positions, we use the value of the mass of Earth, $M_{\oplus} = \qty{5.9722e24}{kg}$ but we neglect that of the debris
+because many orders of magnitudes smaller (and not always known).
 
-These are used to derive numerically compute the weighting function,
-@fig-weighting-inclination-i:
-for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10000$,
-positions and calculate the corresponding Cartesian position and the
-equivalent longitude and latitude.
+Of all the orbital elements, for the needs of this paper to numerically compute
+the weighting function, , @fig-weighting-inclination-i, we only need
+the inclination $i$. 
+For debris with inclination $i$ we simulate an orbit of $N$, i.e. $10000$,
+positions ($\nu$) and calculate the corresponding Cartesian positions and
+the equivalent longitudes and latitudes.
+The latitudes are then binned on half degree intervals to compute
+$\sigma(\varphi, i)$, as per @eq-impact-probability-density.
 

### 2025-05-10 R/plot-collision-risk-kosmos.R

@@ -27,7 +27,7 @@ alt <- 550e3
 
 # inclination of Kosmos 482 Descent Craft (1972-023E, cat. nr. 6073)
 # see https://sattrackcam.blogspot.com/2025/04/kosmos-842-descent-craft-reentry.html
-inclination_deg = 51.7
+inclination_deg = 51.95
 
 weightings_h3_resolution_3_hourly_kosmos <- latitude_weights(
   delta_lat = 0.5,

### 2025-05-10 R/plot_weighting_function_inclination_i.R

@@ -10,7 +10,7 @@ alt <- 550e3
 
 # inclination of Kosmos 482 Descent Craft (1972-023E, cat. nr. 6073)
 # see https://sattrackcam.blogspot.com/2025/04/kosmos-842-descent-craft-reentry.html
-inclination_deg = 51.7
+inclination_deg = 51.95
 
 www <- latitude_weights(
   delta_lat = 0.5,

### 2025-06-07 paper.qmd
address reviewer 4 remark on mission failures and possible future work.
address reviewer 4 comment on `pax`

@@
 The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{500}{m^2}$
-and $pax = 7$ for missing entries which usually are due to smaller aircraft not listed in the above sources.
+(and $pax = 7$)[^pax] for missing entries which usually are due to smaller aircraft not listed in the above sources.
+
+[^pax]: `pax` values, i.e. number of passengers, will be used for future work on
+        casuality risk calculation.
 
@@
 
+Of practical interest for the application of this work and future extension
+is the scenario of mission failure at launch can typically lead to vast airspace
+closure and traffic re-routing: The techniques developed would allow to
+quantify risk and impact of the planned measures.
+
 ::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h" height=80%}

### 2025-06-07 R/plot-weighting-latitude.R
address observation from reviewer 4 about adding y-axis label for Fig 4 and 5

@@ -6,9 +6,9 @@ weightings_half_degree_latitude |>
   ggplot() +
   geom_line(aes(x = lat, y = val), linewidth = 1.3) +
   scale_x_continuous("latitude", breaks = c(-90, -60, -30, 0, 30, 60, 90)) +
+  labs(y = "density") +
   theme_minimal(base_size = 24) +
   theme(
-    axis.title.y = element_blank(),
     panel.grid.minor.x = element_blank(),
   )
 

### 2025-06-07 R/plot_weighting_function_inclination_i.R

@@ -21,9 +21,9 @@ www |>
   ggplot() +
   geom_line(aes(x = lat, y = val), linewidth = 1.3) +
   scale_x_continuous("latitude", breaks = c(-90, -60, -30, 0, 30, 60, 90)) +
+  labs(y = "density") +
   theme_minimal(base_size = 24) +
   theme(
-    axis.title.y = element_blank(),
     panel.grid.minor.x = element_blank(),
   )
 

### 2025-06-07 paper.qmd
address Reviewer 4 observation about paper template (Figure title and reference label); remove links

@@
 title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"
+
 filters:
@@
   - abstract-section
+
 crossref:
   eq-title: Equation     # (default is "")
+  fig-prefix: Fig.   # (default is "Figure")
+  # tbl-prefix: table    # (default is "Table")
+  ref-hyperlink: false # (default is true)
+  fig-title: Fig.
 
@@
     keep-tex: true
+    colorlinks: false
     # fig-pos: 'H'
@@
 #| echo: false
-#| fig-pos: H
+#| fig-pos: t
 knitr::include_graphics("media/figures/cubes.png")

### 2025-06-07 paper.qmd
address observation from reviewer 5 regarding Figure lables/titles and numbering of equations. (Luckily) the position of the last image is now not in the middle of the References.

@@
 crossref:
-  eq-title: Equation     # (default is "")
+  # eq-title: Equation     # (default is "")
   fig-prefix: Fig.   # (default is "Figure")
+  fig-title: Fig.
   # tbl-prefix: table    # (default is "Table")
   ref-hyperlink: false # (default is true)
-  fig-title: Fig.
 
@@
 p_k = \varepsilon_k \int_a \rho(\vec{x}) \sigma_k(\vec{x}) da
-$$
+$$ {#eq-impact-probability-general}
 
@@
 p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h).
-$$
+$$ {#eq-impact-probability-h3}
 
@@
 Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
+
 1. if the aircraft were stationary, the area exposed to the debris impact would be the top face of the rectangular solid, and
@@
 \end{aligned}
-$$
+$$ {#eq-impact-probability-example}
 
@@
 
-$$ \sigma(\varphi) = \frac{1}{N} \sum_{k = 1}^N \sigma(\varphi, i_k)$$
+$$
+\sigma(\varphi) = \frac{1}{N} \sum_{k = 1}^N \sigma(\varphi, i_k)
+$$ {#eq-debris-impact-probability-historical}
 
@@
  E_{h, H} =  \sum_t W_H(h) \cdot \rho_{t,H}(h) \cdot \varepsilon_t
-$$
+$$ {#eq-impact-probability-hourly}
 

### 2025-06-07 paper.qmd
typo

@@
 [^pax]: `pax` values, i.e. number of passengers, will be used for future work on
-        casuality risk calculation.
+        casualty risk calculation.
 
@@
 
-Future work will address the derivation of casuality expectation using
+Future work will address the derivation of casualty expectation using
 seats information for aircraft types in the traffic.

### 2025-06-07 paper.qmd
re-instate true author list

@@
   - id: es
-    name: Author 1
+    # name: Author 1
+    # affiliations:
+    #   - ref: fake
+    name: Enrico Spinielli
     affiliations:
-      - ref: fake
-    # name: Enrico Spinielli
+      - ref: aiu
+    orcid: 0000-0001-8584-9131
+    email: enrico.spinielli@eurocontrol.int
+    attributes:
+      corresponding: true
+  # - name: Author 2
   #   affiliations:
-    #   - ref: aiu
-    # orcid: 0000-0001-8584-9131
-    # email: enrico.spinielli@eurocontrol.int
-    # attributes:
-    #   corresponding: true
-  - name: Author 2
+  #     - ref: fake
+  - name: Rainer Koelle
     affiliations:
-      - ref: fake
-  # - name: Rainer Koelle
+      - ref: aiu
+    orcid: 0000-0003-2810-1404
+    email: rainer.koelle@eurocontrol.int
+  # - name: Author 3
   #   affiliations:
-  #     - ref: aiu
-  #   orcid: 0000-0003-2810-1404
-  #   email: rainer.koelle@eurocontrol.int
-  - name: Author 3
+  #     - ref: fake
+  - name: Quinten Goens
     affiliations:
-      - ref: fake
-  # - name: Quinten Goens
-  #   affiliations: 
-  #     - ref: aiu
-  #   orcid: 0000-0003-2187-1924
-  #   email: quinten.goens@eurocontrol.int
+      - ref: aiu
+    orcid: 0000-0003-2187-1924
+    email: quinten.goens@eurocontrol.int
 affiliations:

### 2025-06-07 paper.qmd
hacked quarto extension for IEEE in order to display authors/affiliations correctly

@@
   - id: es
-    # name: Author 1
-    # affiliations:
-    #   - ref: fake
     name: Enrico Spinielli
@@
       corresponding: true
-  # - name: Author 2
-  #   affiliations:
-  #     - ref: fake
+    role: 
+      - conceptualization
+      - software
   - name: Rainer Koelle
@@
     email: rainer.koelle@eurocontrol.int
-  # - name: Author 3
-  #   affiliations:
-  #     - ref: fake
+    role: 
+      - editing
   - name: Quinten Goens
@@
     email: quinten.goens@eurocontrol.int
+    role: 
+      - editing
 affiliations:

### 2025-06-07 paper.qmd
fixed equation prefix and citation

@@
 crossref:
-  # eq-title: Equation     # (default is "")
+  eq-prefix: ""
   fig-prefix: Fig.   # (default is "Figure")
@@
 
-Then @eq-impact-k becomes
+Then (@eq-impact-k) becomes
 
@@
 The latitudes are then binned on half degree intervals to compute
-$\sigma(\varphi, i)$, as per @eq-impact-probability-density.
+$\sigma(\varphi, i)$, as per (@eq-impact-probability-density).
 

### 2025-06-07 paper.qmd
address observation from Reviewer 5 about missing keywords

@@
 title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"
+
+keywords: 
+  - space debris
+  - airspace closure
+  - aviation safety
 
@@
           \let\qty\SI
-          % NOTE: make siunitx's \qty work without clashing with the one in physics
+          % NOTE: make siunitx's \qty work without clashing with 
+          %      the one in physics
   ieee-html: 
@@
       - conceptualization
+      - writing
       - software

### 2025-06-08 paper.qmd
address reviewer 4 observation on methodology: explained the limitations of ADS-B data usage due to lack of coverage in certain areas and the need to filter out stationary aircraft at airports

@@
 work. 
-<!-- results / conclusions --> This work enhances the previou model by using
+<!-- results / conclusions --> This work enhances the previous model by using
 actual flown flight trajectories as a basis for time-varying flight densities
@@
 approximation of risk exposure. Instead of approximating flight density
-with population density [@patera2008] or aircraft transponder data
+with population density [@patera2008] or aircraft transponder data[^adsb]
 [@wrightAirspaceClosuresDue2025], we use actual flown flight
-trajectories. This provides a more realistic figure of the number of affected air traffic.
-Thus, flight density is directly related to the number of flight operations per volume of airspace.
+trajectories.
+This provides a more realistic figure of the number of affected air traffic.
+Thus, flight density is directly related to the number of flight operations per 
+volume of airspace.
+
+[^adsb]: Crowdsourced ADS-B/Mode S data suffer from lack of coverage away from landmasses and from the need to filter away position reports for stationary aircraft at airports which unnecessarily inflate the density of flying traffic.
 

### 2025-06-08 paper.qmd
address reviewer 4 comment on possibility to use transponder data by adding details about NM trajectories and shortcomings of ADS-B data, namely coverage and static aircraft at airports (inflating density).

@@
     country: USA
+  - id: empty
+
 bibliography: media/references.bib
@@
 
-[^adsb]: Crowdsourced ADS-B/Mode S data suffer from lack of coverage away from landmasses and from the need to filter away position reports for stationary aircraft at airports which unnecessarily inflate the density of flying traffic.
+[^adsb]: Crowdsourced ADS-B/Mode S data suffer from lack of coverage away from landmasses and from the need to filter away position reports for stationary aircraft at airports which unnecessarily inflate the density at the location.
 
 This work focusses on the European region. The hourly flight density
-estimation is based on flown \acr{IFR} trajectories as recorded by
-EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
+estimation is based on _flown_ \acr{IFR} trajectories as recorded
+by EUROCONTROL's \acr{NM}[^1]. These trajectories are reconstructed from
+the filed \acr{FP} and recalculated when big deviations are detected from
+position reports received from \acr{ATC}.
+
+To decrease spatial complexity, the data is
 binned using Uber's H3 geospatial indexing system [@uber-h3]. The

### 2025-06-09 R/plot_aircraft_types_bar_chart.R
save plot aircraft types bar char

@@ -1,0 +1,85 @@
+library(aviodebris)
+library(tidyverse)
+library(arrow)
+library(here)
+
+day <- "2024-07-05" |> as_date()
+
+flts <- here("data", "trajectories_2024-07-05_resampled_30s.parquet") |>
+  read_parquet()
+
+flts |>
+  filter(day <= timestamp, timestamp < (day + 1)) |>
+  distinct(flight_id) |>
+  count()
+
+flts |>
+  filter(day <= timestamp, timestamp < (day + 1)) |>
+  ## turn into lumped factors with capitalized names
+  dplyr::mutate(
+    aircraft_type = forcats::fct_lump(aircraft_type, n = 30)
+  ) |>
+  count(aircraft_type, sort = TRUE)
+
+
+flts |>
+  filter(day <= timestamp, timestamp < (day + 1)) |>
+  count(aircraft_type, sort = TRUE) |>
+  ## add percentage label with `scales::percent()`
+  dplyr::mutate(
+    p = n / sum(n),
+    perc = scales::percent(p, accuracy = .1, trim = FALSE),
+    perc = if_else(row_number() == 1, paste(perc, "of all types"), perc)
+  ) |>
+  filter(p >= 0.01) |>
+  summarise(tr = sum(p))
+
+ccc <- flts |>
+  filter(day <= timestamp, timestamp < (day + 1)) |>
+  ## turn into lumped factors with capitalized names
+  dplyr::mutate(
+    aircraft_type = forcats::fct_lump(aircraft_type, n = 19)
+  ) |>
+  count(aircraft_type, sort = TRUE) |>
+  ## order factor levels by number, put "Other" to end
+  dplyr::mutate(
+    aircraft_type = forcats::fct_rev(forcats::fct_inorder(aircraft_type)),
+    aircraft_type = forcats::fct_relevel(aircraft_type, "Other", after = 0)
+  ) |>
+  ## add percentage label with `scales::percent()`
+  dplyr::mutate(
+    p = n / sum(n),
+    perc = scales::percent(p, accuracy = .1, trim = FALSE),
+    perc = if_else(row_number() == 1, paste(perc, "of all types"), perc)
+  ) |>
+  mutate(
+    ## set justification based on data
+    ## so that only the first label is placed inside
+    place = if_else(row_number() <= 10, 1, 0),
+    ## add some spacing to labels since we cant use nudge_x anymore
[…]

### 2025-06-09 paper.qmd
use em dash
address reviewer 4 comment on safety level of use case: added comparison with lightening strike.

@@
 
-Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes (i.e., flight levels).
+Our focus is on commercial aircraft in flight, and --- implicitly --- operating at higher altitudes (i.e., flight levels).
 The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
@@
 Using the approach in this study and the weightings for this specific debris,
-the cell with the highest impact risk is cell, `83194afffffffff` 
-(located South of London), with a value of $1.89 \cdot  10^{-10}$.
-
-The risk is very low.
+the cell with the highest impact risk is cell, `83195dfffffffff` 
+(located North-West of London and overlapping with the airport of London Heathrow), with a daily value of probability collision risk of $2.63 \cdot  10^{-10}$.
+This corresponds to an annual probability of a collision between this rocket
+body and an aircraft of $9.6 \cdot  10^{-8}$ (1 in 10420000).
+The risk is very low, for a comparison the odds of a person being struck by
+lightning in a given year are less than one in a million
+[@cdcLightningStrikeVictim2024].
 

### 2025-06-09 paper.qmd
spell-checking

@@
 For this purpose we use a hexagon-based tessellation pattern.
-Uber's H3 method ensures non-overlapping hexagons (and 12 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
+Uber's H3 method ensures non-overlapping hexagons (and 12 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the cells. 
 While this paper focusses on the determination of the risk exposure per cell, the approach would allow to identify adjacent cells to account for the inherent uncertainty and risk appetite of decision-makers.
@@
 * constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
-  ([we adopt this value from pag 7-8 in @patera2008])
+  ([we adopt this value from pages 7-8 in @patera2008])
 * constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, under the
@@
 A characteristic point for the determination of the re-entry position is the *'ascending node'*.
-This denotes the intersection point of the object crossing the plane of Earth's equator if moving from the southern hemnisphere to the northern one.
+This denotes the intersection point of the object crossing the plane of Earth's equator if moving from the southern hemisphere to the northern one.
 The respective angular position, $\theta = \omega + \nu$, of the re-entry point from the
@@
 of counting the number of points per cell.
-This binning operation is extremely fast to perform in Uber's H3 gridding system
-and a real advantage in processing the huge amount of data point representing
-the flight trajectories.
+This binning operation is extremely fast to perform in Uber's H3 hierarchical
+geospatial indexing system and a real advantage in processing the huge amount of
+data point representing the flight trajectories.
 
@@
 
-The processing of data for 1 day takes less than 30 min from
+The processing of data for 1 day takes less than 30 minutes from
 data extraction to impact probability calculation on a

### 2025-06-09 paper.qmd
missing "which"

@@
 Of practical interest for the application of this work and future extension
-is the scenario of mission failure at launch can typically lead to vast airspace
-closure and traffic re-routing: The techniques developed would allow to
-quantify risk and impact of the planned measures.
+on casualty risk is the scenario of mission failure at launch
+which can typically lead to vast airspace closure and traffic re-routing:
+the techniques developed here  would help to quantify risk and impact of the
+planned measures.
 

### 2025-06-09 rebuttal.qmd
rebuttal info

---
title: "Rebuttal info"
date: 2025-06-09

format:
  pdf: 
    papersize: a4
---

# Introduction

Thank you very much for your comments and observations for the revision of
the paper 'Preparing for Potential Closure of European Airspaces due to
Re-entering Space Objects'.

In general we think to have addressed all comments from the two reviewers
to the best of our understanding.

We provide details below.

# Feedback

## Failed launches

**Reviewer 4**: In the introduction you stated that the launches are usually well controlled. I would mention the two recent failed launches that led to enormous airspace closures in the Caribbean after launch failures. I think that's certainly a point to also consider (probably beyond the scope of this paper but at least discuss it as future work).

**Authors**: In "Conclusions and Future Work" section we have included the
following paragraph:

> Of practical interest for the application of this work and future extension
> on casualty risk is the scenario of mission failure at launch
> which can typically lead to vast airspace closure and traffic re-routing:
> the techniques developed here  would help to quantify risk and impact of the
> planned measures.

## On transponder data vs real trajectories

**Reviewer 4**: I am not sure why you keep stating that you use real trajectories and don't rely on transponder data. I would assume that these number should match very closely?

**Authors**: we added the following footnote linked to the use of transponder
data of previous work:

> Crowdsourced ADS-B/Mode S data suffer from lack of coverage away from
> landmasses and from the need to filter away position reports for stationary 
> aircraft at airports which unnecessarily inflate the density at the location.

We also further detailed the nature of the trajectory used:

> These trajectories are reconstructed from the filed Flight Plan (FP) and 
> recalculated when big deviations are detected from position reports received
> from Air Traffic Control (ATC).

## Explain `pax`

**Reviewer 4**: You never explain what pax=7 means

**Authors**: we added a footnote to explain why we mentioned pax:

> `pax` values, i.e. number of passengers, will be used for future work on 
> casualty risk calculation.

## Missing axis labels

**Reviewer 4**: Fig 4/5 miss y axis label

**Authors**: we added the axis labels

::: {#fig-orbital-elements layout="[49,-2,49]" fig-env="figure*" fig-pos="h" height=100%}

```{r weighting-inclination-i}
#| out-width: "100%"
#| echo: false
#| fig-cap: Fig. 4
knitr::include_graphics("media/figures/weighting_inclination_i.png")
```

```{r weighting}
#| out-width: "100%"
#| echo: false
#| fig-cap: Fig. 5
knitr::include_graphics("media/figures/weighting.png")
```

:::

## Perspective about collision risk

**Reviewer 4**: You could give a reader some indications what accepted safety levels are (e.g. similar to mid-air collision risk between two aircraft) to put your numbers a bit into perspective.

**Authors**: we compared the collision risk for the use case with the odds of being struck by lightning in order to bring some more tangible example to the
reader.

## Use template

**Reviewer 5**: Use of the template: keywords missing, incorrect formatting of figures and template headers

**Authors**: we added keywords, formatted figures (and references) according to IEEE guidelines, i.e. Fig. N.  
Regarding the headers we checked but they seems composed according to IEEE class (we use a LaTeX template). We need further details to understand what does
not comply...

## Links in PDF

**Reviewer 5**: Why are there blue links in the PDF file?

**Authors**: we removed the links in the PDF.

## Formulas

**Reviewer 5**: Only a few formulas are numbered, why?

**Authors**: we numbered all equations. We also followed IEEE guidelines and changed from "...see Equation 4..." to the proper style "...(4)..."

## References' format

**Reviewer 5**: The references are not in the correct format.

**Authors**: We checked the way references are laid out (this is done from bib file from LaTeX processing) and to us they look ok. We would need more details to understand what is not correct.

## Figure 11 positioning

**Reviewer 5**: Figure 11 should not be placed in between the list of references.

**Authors**: Figure 11 is no more in the middle of the References.
